# De-identification and Data Ethics {#sec-deident}

::: {.bookepigraph}
```{=latex}
\begin{flushright}\small
```
*The right to be let alone is the most comprehensive of
rights, and the right most valued by civilized men.*

[Louis D. Brandeis]{.smallcaps}, *Olmstead v. United
States* (1928)
```{=latex}
\end{flushright}
```
:::

::: {.callout-note}
## Why this chapter exists

A biostatistician working with clinical data handles
protected health information on the first day, and the
obligations that attach to it are not optional. The survey
of peer curricula found explicit treatment of
de-identification and human-subjects protection at Karl
Broman's *Tools for Reproducible Research* and in the
Johns Hopkins statistical-computing sequence, and the
topic is conspicuous by its absence from most others. For
a book aimed at students who will spend their careers with
clinical data, the omission would be indefensible, and
this chapter fills it.
:::

## Prerequisites

Answer the following questions to see if you can bypass
this chapter. You can find the answers at the end of the
chapter in @sec-quiz-answers-02b.

1. Under the HIPAA Privacy Rule, what are the two accepted
   methods for de-identifying protected health
   information, and how do they differ?
2. Why is removing names and medical record numbers
   insufficient to de-identify a dataset, and what is a
   'quasi-identifier'?
3. What is the difference between de-identification and
   anonymisation, and why does the distinction matter for
   what you may promise a participant?

## Learning objectives

By the end of this chapter you should be able to:

- Enumerate the eighteen identifiers of the HIPAA Safe
  Harbor method and apply them to a dataset.
- Distinguish Safe Harbor from Expert Determination and
  choose appropriately.
- Recognise quasi-identifiers and the re-identification
  risk of their combination.
- Apply basic disclosure-control transformations in R:
  suppression, generalisation, date shifting, and hashing.
- State the ethical obligations that survive
  de-identification, including the terms of a data-use
  agreement.

## Orientation

De-identification is the process of transforming data so
that the individuals it describes can no longer reasonably
be identified. Under the United States HIPAA Privacy Rule,
de-identified data fall outside the Rule's restrictions,
which is why de-identification is the gateway to most
secondary analysis and data sharing. But the word carries
less certainty than it appears to: de-identified is not
the same as anonymous, and a dataset stripped of names can
still be re-identified by the combination of the ordinary
variables that remain.

This chapter covers the two regulatory methods, the
practical transformations that implement Safe Harbor, and
the residual risk that no transformation removes. It is
written for the United States regulatory context; the
principles transfer to the European GDPR and to other
regimes, though the specific rules differ.

## The statistician's contribution {#sec-deident-human}

De-identification is not a checklist to be run and
forgotten. The judgements:

**Re-identification risk lives in combinations.** No single
quasi-identifier identifies anyone, but their conjunction
can. The often-cited estimate is that the combination of
five-digit ZIP code, date of birth, and sex uniquely
identifies a large share of the United States population.
The statistician's job is to reason about the joint
distribution, not to tick off fields one at a time.

**The safest data are the data you never collect.** If the
analysis does not need date of birth, collect age; if it
does not need full ZIP, collect the three-digit prefix.
Minimising identifiers at collection is cheaper and safer
than de-identifying them afterwards, and it is the
disclosure-control analogue of the raw-data discipline in
chapter 13.

**De-identification is irreversible by design, and that is
a feature.** A per-patient hash of the medical record
number lets you link visits within your dataset without
storing the number itself. If you keep a crosswalk back to
the identifiers, you have not de-identified the data; you
have created a coded dataset, which is a different and more
regulated thing.

**Promises must match reality.** Do not tell a participant
their data are anonymous if you retain the means to
re-identify them. The consent form and the data-use
agreement are commitments; honour them in the data you
build.

These judgements are what distinguish genuine disclosure
control from a false sense of safety.

## The two HIPAA methods

The Privacy Rule recognises exactly two ways to
de-identify protected health information.

**Safe Harbor** removes eighteen specified categories of
identifier. It is mechanical and checkable, which is its
appeal, but it is blunt: it may remove information the
analysis needs, and it does not certify that
re-identification is impossible, only that the eighteen
categories are gone.

**Expert Determination** engages a qualified statistician
to certify, using accepted methods, that the risk of
re-identification is very small. It is more flexible, and
can retain more analytic detail, but it requires
documented statistical judgement and is correspondingly
more involved. This is a role a biostatistician may
themselves be asked to fill.

## The eighteen Safe Harbor identifiers

Safe Harbor requires removal of the following, for the
individual and for their relatives, employers, and
household members:

1. Names.
2. Geographic subdivisions smaller than a state, including
   street address, city, county, and ZIP code (with a
   narrow exception for the first three ZIP digits when the
   population of that region exceeds 20,000).
3. All date elements finer than a year that relate to an
   individual (birth, admission, discharge, death), and all
   ages over 89.
4. Telephone numbers.
5. Fax numbers.
6. Email addresses.
7. Social Security numbers.
8. Medical record numbers.
9. Health-plan beneficiary numbers.
10. Account numbers.
11. Certificate or licence numbers.
12. Vehicle identifiers and serial numbers, including
    licence plates.
13. Device identifiers and serial numbers.
14. Web URLs.
15. IP addresses.
16. Biometric identifiers, including fingerprints and
    voiceprints.
17. Full-face photographs and comparable images.
18. Any other unique identifying number, characteristic, or
    code, except a re-identification code assigned under the
    Rule's conditions.

The third item is the one that most often surprises
analysts: dates are identifiers. A date of admission,
combined with a hospital and a diagnosis, can single out a
patient. Safe Harbor permits the year but not the day.

## Applying Safe Harbor in R

To make the transformations concrete, we build a small
synthetic dataset carrying several identifiers and apply
Safe Harbor to it. The data are fabricated; no real
protected health information appears here.

```{r}
#| label: deident-data
#| message: false
library(dplyr)

set.seed(47)
phi <- tibble(
  name       = c("Ana Ruiz", "Ben Cole", "Cara Dey", "Dan Fox"),
  mrn        = c("MRN-40021", "MRN-40022", "MRN-40023", "MRN-40024"),
  dob        = as.Date(c("1948-03-11", "1971-09-02",
                         "1933-12-19", "1960-06-30")),
  admit_date = as.Date(c("2026-01-14", "2026-02-03",
                         "2026-01-27", "2026-03-08")),
  zip        = c("92037", "92093", "92037", "59001"),
  sbp        = c(148, 152, 139, 145)
)
phi
```

The transformations: remove the direct identifiers, replace
the medical record number with a salted hash so that
within-dataset linkage survives without the number itself,
shift each patient's dates by a fixed per-patient random
offset (preserving intervals while destroying the actual
calendar dates), generalise ages above 89, and truncate the
ZIP to its first three digits, suppressing prefixes from
sparsely populated regions.

```{r}
#| label: deident-transform
salt <- "study-specific-secret-not-in-the-repo"

# per-patient date offset: preserves intervals, hides calendar dates
set.seed(1)
offsets <- sample(-90:90, nrow(phi), replace = TRUE)

# ZIP prefixes with population under 20,000 are suppressed by
# Safe Harbor; here we treat 590 as such an example
sparse_zip3 <- c("590", "036", "692", "878")

deid <- phi |>
  mutate(
    subject_id = vapply(paste0(salt, mrn),
                        function(x) substr(digest::digest(x, "sha256"),
                                           1, 12),
                        character(1)),
    age        = as.integer(floor(
                   as.numeric(admit_date - dob) / 365.25)),
    age        = if_else(age > 89L, NA_integer_, age),  # 90+ suppressed
    admit_day  = admit_date + offsets,                  # shifted
    zip3       = substr(zip, 1, 3),
    zip3       = if_else(zip3 %in% sparse_zip3, NA_character_, zip3)
  ) |>
  select(subject_id, age, admit_day, zip3, sbp)

deid
```

Two features are worth noting. The `subject_id` is a
one-way hash: the same medical record number always
produces the same code, so a patient's visits still link,
but the code cannot be reversed to recover the number,
provided the salt is kept out of the repository. The date
shift preserves the *interval* between a patient's
admission and any other of their dates, which is what a
longitudinal analysis needs, while destroying the actual
admission day. What Safe Harbor does not do is guarantee
that the surviving combination of age, region, and clinical
values is unique to no one, which is the subject of the
next section.

::: {.callout-note icon=false collapse=true}
## Check your understanding: is this anonymous?

**Question.** After the transformation above, a colleague
proposes releasing the de-identified table publicly,
describing it as 'anonymous'. The dataset has four rows.
What is your concern?

**Answer.**

On four rows, the combination of age, three-digit ZIP, and
systolic blood pressure may still single out an individual
to anyone who knows a patient was in the study, because
each row is nearly unique on those quasi-identifiers. Safe
Harbor removes the eighteen categories but does not certify
that the remainder is non-unique; on small datasets the
residual re-identification risk can be high. 'Anonymous'
overstates what has been achieved. The honest description
is 'de-identified under Safe Harbor', and a public release
of a small dataset should be evaluated for
re-identification risk (Expert Determination) rather than
assumed safe.

:::

## Quasi-identifiers and residual risk

A quasi-identifier is a variable that does not identify
anyone alone but contributes to identification in
combination: age, sex, ZIP, admission date, occupation,
rare diagnosis. The formal tools for reasoning about their
joint risk come from statistical disclosure control. The
simplest is *k-anonymity*: a dataset is k-anonymous on a
set of quasi-identifiers if every combination of their
values that appears is shared by at least $k$ records, so
that no individual is distinguishable from at least $k-1$
others. Achieving a target $k$ is a matter of generalising
(coarsening age into bands, ZIP into prefixes) and
suppressing (removing the rare rows) until the condition
holds. Refinements such as l-diversity and t-closeness
address weaknesses that k-anonymity alone does not, and the
`sdcMicro` package implements these methods for R.

The practical lesson is that de-identification is a
risk-reduction exercise, not a binary state. The
statistician's contribution is to quantify the residual
risk and to match the transformation to the sensitivity of
the data and the openness of the release.

## Obligations that survive de-identification

De-identifying the data does not discharge every
obligation. If the data arrived under a data-use agreement,
its terms bind you regardless of de-identification: a DUA
commonly forbids any attempt at re-identification,
restricts redistribution, and requires secure storage and
eventual destruction. The ADNI case study in @sec-adni
turns on exactly such an agreement. Human-subjects
protections, institutional review board conditions, and
the promises made in the consent form persist as well.
De-identification changes what the HIPAA Privacy Rule
requires; it does not override a contract or a promise.

## Collaborating with an LLM on de-identification

LLMs know the eighteen categories and can draft
transformation code; the danger is treating their output
as a compliance guarantee.

**Prompt 1: auditing a schema.** Paste a list of column
names and ask which correspond to Safe Harbor identifiers.

*What to watch for.* The model will catch names and MRNs
readily; it is less reliable on the subtle cases, dates,
the ZIP exception, and free-text fields that may embed
identifiers. Treat the output as a first pass, not a
certification.

*Verification.* Confirm against the official eighteen
categories, and inspect any free-text field by hand, since
a notes column can contain names the schema does not
reveal.

**Prompt 2: writing the transformation.** Describe the
identifiers present and ask for R code to suppress,
generalise, hash, and date-shift them.

*What to watch for.* Whether the hash is salted and whether
the salt is kept out of the repository; whether date
shifting preserves within-patient intervals; whether ages
over 89 are handled.

*Verification.* Run the code on synthetic data; confirm no
direct identifier survives and that within-patient linkage
still works through the hash.

## Principle in use

Three habits define defensible de-identification:

1. **Minimise at collection.** Collect age, not date of
   birth; the three-digit ZIP, not the full one. The safest
   identifier is the one you never held.
2. **Reason about combinations.** Re-identification risk is
   a property of the joint distribution of quasi-identifiers,
   not of any single field.
3. **Match the claim to the method.** 'De-identified under
   Safe Harbor' is a defensible claim; 'anonymous' usually
   is not. Honour the data-use agreement regardless.

## Exercises

1. Enumerate the eighteen Safe Harbor identifiers from
   memory, then check your list against the official one.
   Which did you miss?
2. Take the synthetic `phi` dataset from this chapter and
   add a free-text `notes` column containing a sentence
   that mentions a patient's name. Write code that detects
   and flags the identifier in the free text; explain why
   this is harder than handling a structured column.
3. Construct a synthetic dataset of 200 patients with age,
   sex, and three-digit ZIP. Compute the largest $k$ for
   which the dataset is k-anonymous on those three
   quasi-identifiers. Generalise age into decade bands and
   recompute; by how much does $k$ improve?
4. Implement date shifting that preserves within-patient
   intervals but uses a different random offset per patient.
   Verify on synthetic data that the interval between two
   of a patient's dates is unchanged while the calendar
   dates differ.
5. Explain, in a paragraph suitable for a data-management
   plan, why a salted hash of the medical record number is
   preferable to either keeping the number or dropping the
   ability to link visits.

## Further reading

- The HHS guidance on de-identification methods at
  `hhs.gov/hipaa`, the authoritative source for Safe Harbor
  and Expert Determination.
- The `sdcMicro` package and its vignettes, for k-anonymity
  and statistical disclosure control in R.
- El Emam and Arbuckle (2013), *Anonymizing Health Data*,
  a practical book-length treatment.
- @sec-federal on the federal data-sharing requirements
  that de-identification enables.

## Prerequisites answers {#sec-quiz-answers-02b}

1. Safe Harbor removes eighteen specified categories of
   identifier; it is mechanical and checkable but blunt.
   Expert Determination has a qualified statistician certify,
   by accepted methods, that the re-identification risk is
   very small; it is more flexible and can retain more
   analytic detail but requires documented statistical
   judgement.
2. Names and medical record numbers are direct identifiers,
   but the ordinary variables that remain, age, ZIP,
   admission date, sex, so-called quasi-identifiers, can
   identify an individual in combination even though none
   does alone. The classic result is that ZIP, date of
   birth, and sex together are unique for a large share of
   the population.
3. De-identification reduces re-identification risk under a
   defined method (such as Safe Harbor) but does not
   guarantee anonymity; the residual risk may be
   non-trivial, especially on small datasets. Anonymisation
   implies re-identification is effectively impossible. The
   distinction matters because you may only promise a
   participant what the data actually deliver, and because a
   retained crosswalk makes data 'coded', not
   de-identified.
