# Ethics Beyond Compliance {#sec-ethics}

::: {.bookepigraph}
```{=latex}
\begin{flushright}\small
```
*The first principle is that you must not fool yourself, and
you are the easiest person to fool.*

[Richard P. Feynman]{.smallcaps}, *Cargo Cult Science* (1974)
```{=latex}
\end{flushright}
```
:::

::: {.callout-note}
## Why this chapter exists

The book's treatment of ethics has so far been entirely
regulatory. @sec-federal covers what the funders require,
@sec-deident covers what the HIPAA Privacy Rule requires, and
@sec-cdisc covers what the FDA requires. All three are
necessary and none is sufficient, because an analysis can
satisfy every one of them and still mislead every reader it
reaches. This chapter is about the failures that compliance
does not prevent, and about which the statistician, not the
institutional review board, is the last line of defense.
:::

## Prerequisites

Answer the following questions to see if you can bypass this
chapter. You can find the answers at the end of the chapter in
@sec-quiz-answers-02c.

1. A dataset is assembled entirely from information that
   individuals published about themselves in public. Explain how
   releasing it can nonetheless harm them.
2. Two group means have 95% confidence intervals that overlap.
   May you conclude that the groups do not differ significantly?
3. Name three ways a figure can mislead a reader without
   containing a single false number.

## Learning objectives

By the end of this chapter you should be able to:

- Distinguish the harm of disclosure from the harm of
  aggregation, and recognize that public data can carry the
  second without the first.
- Identify causal language attached to associational evidence,
  in your own writing and in the manuscripts you review.
- Recognize the standard visual distortions: the truncated axis,
  the area-for-quantity encoding, and the dual axis.
- Demonstrate why overlapping confidence intervals do not imply
  a non-significant difference, and stop making the inference.
- Distinguish a sample from the target population, and state the
  groups to which an estimate does and does not transport.
- Locate where in the workflow each of these failures is
  introduced, and where it can be caught.

## Orientation

The regulatory chapters of this book describe a floor. They
tell you what you must not do: release identifiers, skip the
data-management plan, submit a dataset that fails conformance.
Clearing that floor is a necessary condition for defensible
practice and it is nowhere near a sufficient one, because the
floor is written in terms of *procedures* and the failures that
matter most are failures of *inference*.

Five such failures recur in biomedical work, and they have a
common structure: at no point does anyone state a falsehood.
The data are real, the computation is correct, the p-value is
what the software returned, the figure plots the numbers it
claims to plot. The misleading happens in the gap between what
was computed and what the reader concludes, and it is precisely
in that gap that the statistician's professional obligation
lives, because the statistician is the only person on the team
who can see the gap at all.

## The statistician's contribution {#sec-ethics-human}

This chapter is judgement from beginning to end.

**You are the last person who can catch this.** The clinical
collaborator cannot evaluate whether the confidence interval
supports the sentence in the abstract; that is why they have
you. The reviewer may or may not look. The reader certainly
will not. When the press release says 'exercise prevents
cancer' and your analysis found an association in an
observational cohort of volunteers, you are not a bystander to
that sentence. You are its last opportunity.

**Public is not the same as consented, and available is not the
same as ethical.** The question 'may I' has a legal answer and
an ethical one, and they are different questions. A dataset
that violates no rule can still hurt the people in it. Ask what
happens to a person in your data if the analysis is right, and
then ask what happens if it is wrong.

**Distortion is usually accidental, which does not make it
harmless.** Almost nobody sets out to deceive with a truncated
axis. They set out to make the effect visible, and a truncated
axis makes it visible, and the reader who glances at the figure
carries away an impression the data do not support. Intent is
not the standard. The standard is what a competent reader would
conclude.

**Refusing is part of the job.** There is a point at which the
correct professional act is to decline: to refuse to sign the
manuscript, to refuse to produce the figure as specified, to
say that the analysis as requested is not one you will perform.
@sec-team-science distinguishes preference from principle
precisely so that you spend your credibility on the cases where
this matters. This chapter is the catalog of those cases.

## Aggregation harm

The first failure concerns data that were never private.

In 2016 a group of researchers released a dataset of some
70,000 profiles scraped from a dating website, including
usernames alongside answers to intimate questions about
sexuality, drug use, and political views. The defense offered
was that the data 'are already publicly available', which was
true and beside the point. Each user had disclosed a fact to an
audience of their choosing, in a context they understood. The
release did something none of them had chosen: it collected
those disclosures into a single, searchable, permanent,
downloadable artifact, and attached them to identifiers.

This is **aggregation harm**, and it is the same phenomenon
that @sec-deident treats formally under the heading of
quasi-identifiers, arriving from the other direction. There,
the point was that no single ordinary variable identifies
anyone but their conjunction can. Here, the point is that no
single disclosure harms anyone but their conjunction can. The
mathematics is the mathematics of joint distributions; the
ethics is the ethics of context. Information carries with it an
implicit expectation about where it will travel, and moving it
somewhere else is an act with consequences even when every
individual datum was freely given.

The practical tests, before you assemble a dataset from public
sources:

- **Would the people in it be surprised?** Not 'would they
  object', which invites you to answer for them, but 'would
  they be surprised to find themselves in it'. Surprise is a
  reliable signal that a contextual expectation is being
  violated.
- **Does the aggregate reveal what the parts did not?** Someone
  who posted in a support forum has disclosed a health
  condition to that forum. A dataset linking forum posts to
  employer names has disclosed it to their employer.
- **Can it be de-identified without destroying the analysis?**
  If yes, do that (@sec-deident). If no, that is informative:
  it means the identifiers are load-bearing, and you should be
  able to say why.

## Causal overclaiming

The second failure is the most common one in biomedical
publishing, and it usually happens in the last paragraph.

An observational cohort of self-selected volunteers reports
that those who exercised more had lower cancer incidence,
adjusting for age, sex, smoking, and body-mass index. The
finding is an association. The discussion section says the
finding 'suggests that exercise reduces cancer risk'. The press
office says exercise prevents cancer. The newspaper says it
cures it. At no step did anyone fabricate a number.

The statistician's job is to insist that the verb match the
design. The three registers, from @sec-team-science, are worth
restating as a hierarchy of claims:

| Design | What you may say |
|---|---|
| Randomized, adequately powered, pre-specified | 'X reduced Y by $\delta$' |
| Observational, well adjusted, no unmeasured confounding assumed | 'X was associated with lower Y; the association is consistent with a causal effect but the design cannot establish one' |
| Observational, self-selected sample, self-reported exposure | 'Among volunteers who reported X, Y was lower. Volunteers who exercise differ from those who do not in ways we did not measure.' |

The third row is the one to keep in view. The problem with a
volunteer cohort reporting its own exercise is not that the
confounder adjustment was inadequate. It is that people who
exercise and report it differ systematically from people who do
not, along dimensions (conscientiousness, health-seeking
behavior, unmeasured comorbidity, the very early symptoms of
undiagnosed disease) that no covariate list captures. The
adjustment cannot fix this, and stating that you adjusted for
age and smoking implies to a reader that it did.

The specific place this fails is the transition from Results to
Discussion, where the verb changes from 'was associated with'
to 'reduces' without anyone noticing. Read your own discussion
section as a hostile reviewer, and grep for causal verbs:
reduces, prevents, protects, causes, leads to, improves. Each
one is a claim your design must license.

## Visual misrepresentation

The third failure needs no numbers at all, only their encoding.
The figure below plots the same two group means twice.

```{r}
#| label: fig-truncated
#| fig-cap: "The same two means, encoded twice. Panel A begins the y-axis
#|   at zero and the difference is what it is: small. Panel B begins the
#|   axis at 71 and the same difference fills the panel. Neither panel
#|   contains a false number, and a reader who glances at B carries away
#|   a conclusion the data do not support."
#| fig-width: 7
#| fig-height: 3.5
#| message: false
library(tidyverse)
library(patchwork)

d <- tibble(
  arm  = c("Control", "Treatment"),
  mean = c(72.4, 74.1)
)

base <- ggplot(d, aes(arm, mean, fill = arm)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  scale_fill_manual(values = c(Control = "#9d2235",
                               Treatment = "#1f4e79")) +
  labs(x = NULL, y = "Score") +
  theme_minimal()

p_honest    <- base +
  coord_cartesian(ylim = c(0, 100)) +
  labs(title = "A. Axis from zero")

p_truncated <- base +
  coord_cartesian(ylim = c(71, 75)) +
  labs(title = "B. Axis from 71")

p_honest + p_truncated
```

The difference is 1.7 points on a 100-point scale. Panel A
tells you that. Panel B tells you that the treatment arm is
roughly three times the control arm, which is false, and it
tells you so without printing a single incorrect value.

The three standard distortions:

**The truncated axis.** As above. The rule is not 'always start
at zero', because for many quantities zero is meaningless (body
temperature, calendar year, pH) and forcing it wastes the panel.
The rule is that *when the visual encoding is length or area*,
as in a bar chart, the baseline must be zero, because the
reader is comparing lengths and a truncated bar's length is not
proportional to its value. For a point-and-interval plot, where
the encoding is position rather than length, a truncated axis is
legitimate and often necessary. The question is always: what is
the reader's eye measuring?

**Area for quantity.** A map colored by which arm won each
county encodes *land area*, and land does not vote. A bubble
whose *radius* is proportional to a count exaggerates it, because
the eye reads area, which goes as the square of the radius. In
biomedicine the recurring case is a map of disease counts that
is really a map of where people live, and the fix is a rate, a
denominator, and a cartogram or a population-weighted encoding.

**The dual axis.** Two series on two differently scaled y-axes
can be made to cross, converge, or diverge at the analyst's
choice, because the relative scaling is arbitrary. Any apparent
relationship between them is an artifact of a decision you made.
Avoid dual axes entirely; if two series must be compared, index
them to a common baseline or use two panels.

## The overlapping-interval fallacy

The fourth failure is the one a biostatistician is most likely
to commit personally, and it is a genuine technical error rather
than a rhetorical one.

The reasoning goes: the 95% confidence intervals for the two
group means overlap, therefore the difference is not significant
at the 5% level. This is wrong, and it is wrong in the direction
of missing real effects. The interval for a *difference* depends
on the standard error of the difference, which is smaller than
the sum of the two individual standard errors, and intervals for
the two means can overlap substantially while the interval for
their difference excludes zero.

A demonstration, on synthetic data constructed to make the point
sharply:

```{r}
#| label: overlap-demo
set.seed(18)

a <- rnorm(60, mean = 100, sd = 12)
b <- rnorm(60, mean = 105, sd = 12)

ci <- function(x) {
  m  <- mean(x)
  se <- sd(x) / sqrt(length(x))
  c(lower = m - 1.96 * se, mean = m, upper = m + 1.96 * se)
}

rbind(group_a = ci(a), group_b = ci(b)) |> round(1)

t.test(b, a)
```

The two intervals overlap: group A reaches 102.7 and group B
begins at 101.7. The two-sample test nonetheless rejects at the
5% level ($p = 0.02$). Anyone reading the overlap off a figure
and concluding 'no significant difference' has reached the
opposite conclusion from the one the data support.

The overlap here is about one point, which is deliberate: the
fallacy does its damage in exactly this regime, where the
intervals graze each other and the difference is real. As the
difference grows the intervals separate and the two readings
agree; as it shrinks the test stops rejecting and they agree
again. It is the middle, where an analyst is most likely to be
looking for guidance from the figure, that the overlap heuristic
gets wrong.

@fig-overlap shows why. The correct object to look at is not
the pair of intervals but the interval for their difference,
which is a different quantity and answers the question actually
being asked.

```{r}
#| label: fig-overlap
#| fig-cap: "Left: the two group means with their 95% intervals, which
#|   overlap. Right: the 95% interval for the difference, which excludes
#|   zero. These are not in conflict; they are answers to different
#|   questions. Reading significance off the overlap on the left is the
#|   error."
#| fig-width: 7
#| fig-height: 3.2
means <- bind_rows(
  as_tibble_row(ci(a)) |> mutate(group = "A"),
  as_tibble_row(ci(b)) |> mutate(group = "B")
)

tt <- t.test(b, a)
diff <- tibble(
  group = "B - A",
  mean  = unname(diff(rev(tt$estimate))),
  lower = tt$conf.int[1],
  upper = tt$conf.int[2]
)

p_means <- ggplot(means, aes(mean, group)) +
  geom_pointrange(aes(xmin = lower, xmax = upper)) +
  labs(title = "A. Group means", x = "Score", y = NULL) +
  theme_minimal()

p_diff <- ggplot(diff, aes(mean, group)) +
  geom_pointrange(aes(xmin = lower, xmax = upper)) +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "#9d2235") +
  labs(title = "B. Difference", x = "Difference in score",
       y = NULL) +
  theme_minimal()

p_means + p_diff
```

The practical consequences are two. When you *report*, give the
interval for the contrast, not merely intervals for the arms,
because the contrast is the estimand. When you *review*, treat
any sentence of the form 'the intervals overlapped, so the
groups did not differ' as an error, and say so.

## Unrepresentative samples

The fifth failure is the one whose costs fall on people who
are not in the room, and often not in the data. It is the
failure of **fair representation**, and it is the reason a
competent statistician asks who a sample contains before,
not after, generalizing from it.

The structure is by now familiar. A cohort is assembled, an
estimate is computed, and the estimate is valid for the
population that was actually sampled. The falsehood, once
again, is in no number; it is in the silent step from 'this
holds in the sample' to 'this holds', taken when the sample
and the target are not the same population and no one has
said so. This book's own source material shows the
mechanism: the survey behind this chapter's team-science
reading drew respondents who were 82.8% White, and its
authors state plainly that their conclusions may not
transport to a more diverse workforce [@slade2023team]. The
honesty is the model; the skew is the hazard.

**A sample is evidence about the population it was drawn
from, and no other.** A drug trial that enrolls
overwhelmingly at academic medical centers, among patients
who can attend fortnightly visits, skews toward the younger,
the insured, the urban, and the un-comorbid. The estimate it
produces is an estimate for those patients. Applied to the
elderly, rural, multiply-morbid patient to whom the drug
will actually be prescribed, it is an extrapolation, and an
extrapolation is a claim the data do not contain. Adjusting
for age does not repair this, for the same reason it did not
repair the volunteer cohort earlier in this chapter: the
groups differ along dimensions no covariate list enumerates.

**Under-representation is not a diversity checkbox; it is a
bias-and-variance problem with a face.** When a subgroup is
a small fraction of the sample, two things hold at once. The
estimate for that subgroup is imprecise, because its $n$ is
small, so a real effect specific to them is the effect you
are least able to detect. And if the effect genuinely
differs there, the pooled estimate, dominated by the
majority, is biased for them in particular. A result
reported as though it applied to everyone therefore fails
exactly the people the sample underweighted, and fails them
silently, because the interval on the pooled estimate looks
reassuringly tight.

**A model learns the sample's inequities and returns them as
findings.** Where an analysis feeds a decision rule, a risk
score, an eligibility threshold, a resource-allocation
model, an unrepresentative or biased training sample is no
longer merely imprecise; it is a mechanism for propagating
the bias at scale. The canonical case is a health-system
algorithm that used prior health-care spending as a proxy
for health need and, because less had historically been
spent on Black patients at equal levels of illness,
systematically understated their need
[@obermeyer2019dissecting]. Every number was correctly
computed. The proxy was the error, and only someone asking
what the variable *stood for* could have caught it.

The practical discipline is threefold and cheap:

- **Report who is in the sample beside who the result is
  for.** A Table 1 giving the demographic composition is the
  minimum, and the Discussion should name the populations to
  which the estimate does and does not transport. 'Our cohort
  was 91% White and privately insured; generalization to
  other populations is unestablished' is one sentence, and it
  is the honest one.
- **Pre-specify the subgroups you are powered to examine, and
  label the rest exploratory.** This is the discipline of
  @sec-analysis-plan applied to representation: a subgroup you
  were not powered for yields an imprecise estimate, not an
  absence of effect, and the report must say which it is.
- **When a variable feeds a decision, ask what it is a proxy
  for.** Before a predictor enters a model that will act on
  people, ask whether it encodes a protected characteristic
  and whether the model's errors are borne equally. This is
  the fairness audit, and it is cheap relative to the harm it
  averts.

Fair representation has a second face, downstream of the
analysis: whether every reader can in fact read the result.
A figure that carries its only distinction in red versus
green is illegible to the roughly eight per cent of men with
a red-green color-vision deficiency, and the fix, a
color-blind-safe palette verified before publication, is
treated concretely in @sec-graphics. Accessibility of the
artifact is the same principle as representativeness of the
sample, arriving at the far end of the pipeline: a result
that reaches only some of its readers, or speaks for only
some of its subjects, is not the finished product it appears
to be.

::: {.callout-note icon=false collapse=true}
## Check your understanding: the compliant misleading paper

**Question.** A paper reports an observational study. The IRB
approved it, the data are properly de-identified under Safe
Harbor, the DMSP is in place, the code is on Zenodo with a DOI,
and every number in the paper is correctly computed from the
data. The abstract concludes that the intervention 'reduces
30-day readmission'. Has anything gone wrong?

**Answer.**

Yes, and nothing in the compliance apparatus was built to catch
it. Every requirement in @sec-federal and @sec-deident is
satisfied, the analysis is reproducible in the sense of
@sec-why, and the reported numbers are right. What is wrong is a
verb: an observational design licenses 'was associated with
lower readmission', not 'reduces'. The distinction is not
pedantry. A reader, a guideline committee, or a health system
deciding where to spend money will act on 'reduces', and the
evidence does not support the action.

Note what this implies about the rest of the book.
Reproducibility, compliance, and correctness of computation are
each necessary and they are jointly insufficient: a perfectly
reproducible, fully compliant, numerically correct analysis can
still mislead every reader it reaches. The tools cannot close
that gap. Only the analyst can.

:::

## Where each failure is caught

The five failures enter the workflow at different points, which
means they are caught at different points.

| Failure | Introduced at | Caught by |
|---|---|---|
| Aggregation harm | Data acquisition (@sec-acquisition) | Asking, before you assemble, whether the people in the data would be surprised |
| Causal overclaiming | The Results-to-Discussion transition | Reading your own discussion for causal verbs; peer review (@sec-communicating) |
| Visual misrepresentation | Figure construction (@sec-graphics) | Asking what the reader's eye is measuring |
| Overlapping-interval fallacy | Interpretation, and often the figure | Reporting the contrast, not the arms |
| Unrepresentative sample | Study design and recruitment; sometimes acquisition | Reporting who was sampled beside who the result is for; a proxy audit before a model acts |

None of them is caught by a test suite, a lockfile, or a
container. That is the point of the chapter, and it is worth
stating plainly: the entire infrastructural apparatus of this
book is orthogonal to whether the analysis misleads.
Reproducibility guarantees that a wrong conclusion can be
regenerated exactly.

## Collaborating with an LLM on ethical review

Models are useful here and dangerous in a specific way: they are
agreeable, and every failure in this chapter is one an agreeable
collaborator would let pass.

**Prompt 1: auditing your own discussion section.** Paste it and
ask: 'list every sentence whose verb implies causation, and for
each, state what design would license it.'

*What to watch for.* This works well, because it is a
text-classification task rather than a judgement. Expect a few
false positives. The sentences it flags are the sentences a
skeptical reviewer will flag.

*Verification.* For each flagged sentence, check the design
against the table in this chapter. If the design is
observational and the verb is causal, rewrite it.

**Prompt 2: critiquing a figure.** Describe the encoding (bar
chart, y-axis from 71 to 75, two groups) and ask what a reader
would wrongly conclude.

*What to watch for.* The model will identify the truncated
baseline reliably. It is weaker on subtler distortions, such as
an area encoding or an inappropriate denominator, because those
require knowing what the quantity means.

*Verification.* Render the figure both ways, as this chapter
does, and look at them side by side. The comparison is more
persuasive than any critique.

**Prompt 3: the one it will get wrong.** Ask it whether you may
release a dataset assembled from public profiles.

*What to watch for.* The model is likely to reason from
legality, because legality is what is written down, and to
conclude that public data may be republished. Push it: ask what
harm a person in the dataset could suffer, and ask whether they
would be surprised. The considerations exist in its training
data but it will not volunteer them, and this is exactly the
territory where an agreeable collaborator is worse than none.

*Verification.* Take the question to your IRB, and to a
colleague who will tell you no.

## Principle in use

In closing, three habits define what we would regard as defensible practice beyond compliance:

1. **Match the verb to the design.** Observational studies
   license association. If the discussion section says
   'reduces', the design had better be randomized.
2. **Ask what the reader's eye is measuring.** A bar encodes
   length, so its baseline is zero. A map encodes area, and land
   does not have a disease. A dual axis encodes an arbitrary
   choice you made.
3. **Compliance is a floor, not a ceiling.** Every regulatory
   requirement in this book can be satisfied by an analysis that
   misleads every reader it reaches. You are the only safeguard
   against that, and it is the part of the job no tool performs.

## Exercises

1. Take the discussion section of a paper you have co-authored.
   List every causal verb. For each, state the design that would
   license it and the design you actually had. Rewrite any that
   fail.
2. Reproduce the truncated-axis figure in this chapter with data
   of your own. Show it to a colleague without comment and ask
   them to describe the effect size in words. Then show them the
   zero-baseline version and ask again.
3. Construct a dataset in which two group means have overlapping
   95% intervals and a two-sample t-test rejects at the 5%
   level. State the largest overlap you can achieve while
   retaining significance, and explain the relationship between
   the standard error of a mean and the standard error of a
   difference.
4. Find a published figure in your field that encodes a quantity
   as area or uses a dual axis. Redraw it honestly. Write a
   paragraph on what changes.
5. Write a one-page policy, suitable for a research group, on
   when data assembled from public sources may be released.
   Include the questions you would require an analyst to answer
   before assembling such a dataset.
6. Take a published clinical study in your field and find its
   Table 1. Identify the demographic groups under-represented
   relative to the population the intervention is meant to
   serve. Write the single sentence the Discussion should
   contain about generalization, and check whether it does.

## Further reading

- [@altman1995absence], on the specific inferential error that
  the overlapping-interval fallacy is a cousin of.
- [@wilke2019fundamentals], whose chapters on misleading axes
  and on visualizing uncertainty are the extended treatment of
  this chapter's third and fourth sections.
- [@elemam2013anonymizing], for the disclosure-control side of
  the aggregation problem.
- [@goodman2016what], for the reminder that reproducibility and
  correctness are independent qualities.
- [@obermeyer2019dissecting], the health-system algorithm whose
  proxy variable encoded and propagated racial bias at scale.
- [@rothwell2005external], on external validity and the question
  a trial silently answers: to whom do its results apply?
- @sec-deident, @sec-federal, and @sec-graphics, the chapters
  this one sits between.

## Prerequisites answers {#sec-quiz-answers-02c}

1. Because the harm of aggregation differs from the harm of
   disclosure. Each individual disclosed a fact to an audience of
   their choosing, in a context they understood; assembling those
   disclosures into a single searchable, permanent, downloadable
   artifact and attaching identifiers to it does something none
   of them chose. Information carries an implicit expectation
   about where it will travel, and the aggregate can reveal what
   no individual datum did, for example by linking a health
   disclosure in one forum to an employer named in another. This
   is the same joint-distribution phenomenon that makes
   quasi-identifiers dangerous in @sec-deident, arriving from the
   other direction. The test is not 'was each item public' but
   'would the people in the dataset be surprised to find
   themselves in it'.
2. No. This is a common and consequential error. The confidence
   interval for the *difference* depends on the standard error of
   the difference, which is smaller than the sum of the two
   individual standard errors, so two means can have substantially
   overlapping intervals while the interval for their difference
   excludes zero and the two-sample test rejects. The chapter
   demonstrates exactly this case. The error runs in the direction
   of missing real effects. Report and inspect the interval for
   the contrast, which is the estimand, rather than reading
   significance off the overlap of the arms.
3. (Any three of:) a truncated axis on a length-encoded chart, so
   that a 1.7-point difference on a 100-point scale fills the
   panel; encoding a quantity as area, including a map colored by
   land rather than by population, or a bubble whose radius rather
   than area is proportional to the count; a dual axis, whose
   relative scaling is an arbitrary analyst choice that can be
   tuned to make two series appear to converge or diverge;
   omitting the uncertainty entirely, so that a point estimate
   invites more confidence than it has earned; and choosing a
   denominator that flatters the conclusion. None of these
   requires printing a false number.
