# Getting Data In: Files, APIs, and Scraping {#sec-acquisition}

::: {.bookepigraph}
```{=latex}
\begin{flushright}\small
```
*It is a capital mistake to theorize before one has data.*

[Arthur Conan Doyle]{.smallcaps}, *A Scandal in Bohemia*
(1891)
```{=latex}
\end{flushright}
```
:::

::: {.callout-note}
## Why this chapter exists

Every data chapter that follows this one begins with a data
frame already in memory, as though it had arrived there by
itself. It did not. Someone read a file whose encoding was
wrong, or queried an API whose token had expired, or was
handed a spreadsheet in which the dates had turned into
five-digit numbers. That work is where a substantial share of
an analyst's time goes, and it is where a substantial share of
the errors that survive into publication are introduced,
because a value that is silently misread is a value that no
downstream test will catch. This chapter covers the three ways
data actually arrive: as a file, through an interface, and, as
a last resort, off a web page.
:::

## Prerequisites

Answer the following questions to see if you can bypass this
chapter. You can find the answers at the end of the chapter in
@sec-quiz-answers-12b.

1. `read_csv()` guesses the type of every column from the first
   thousand rows. Name two ways this guess can be wrong in a
   manner that no error message will reveal, and say what you
   should do instead.
2. You obtain your analytic dataset by querying an institutional
   API on the day you begin the analysis. What must you commit
   to the repository so that the analysis is reproducible, and
   what must you *not* commit?
3. A public web page contains the table you need. What questions
   should you answer before writing the scraper, and what is the
   first thing you should do with the data once you have it?

## Learning objectives

By the end of this chapter you should be able to:

- Read delimited files with an explicit column specification,
  and treat that specification as a data-integrity assertion
  rather than a convenience.
- Diagnose the common file-level pathologies: wrong delimiter,
  wrong encoding, byte-order marks, sentinel missing values, and
  Excel serial dates.
- Query a token-authenticated REST API from R, keeping the token
  out of the repository, and handle pagination and rate limits.
- Parse a nested JSON response into a tidy data frame.
- Scrape an HTML table when no interface exists, and decide
  whether you should.
- Cache every acquisition to a dated, immutable raw file, so that
  the analysis does not depend on the network or on the state of
  a remote system at render time.

## Orientation

Data arrive in three ways, and they differ in how much control
you have over what arrives.

A **file** is a snapshot someone made for you. It is inert, it
does not change under you, and its pathologies are all present
at the moment you receive it: the encoding, the delimiter, the
sentinel values that some upstream system used to mean 'missing'.
Reading a file correctly is a matter of asserting what you
expect and checking that you got it.

An **interface**, by which I mean a REST API such as REDCap's or
an institutional data warehouse's, is a query against a living
system. What it returns today is not what it returned last month,
because rows have been added, corrected, and occasionally
deleted. The reproducibility problem is therefore different in
kind: the code is deterministic but the data source is not, and
the only defense is to freeze what you received.

A **scrape** is what you resort to when the data exist but no
interface does. It is the most fragile of the three, because the
page's structure is not a contract and can change without notice,
and it is the only one of the three that raises a question about
whether you are entitled to the data at all.

The three converge on a single discipline, which is the
raw-data rule of @sec-wrangling stated one step earlier in the
pipeline: whatever you acquire, write it down, date it, and never
acquire it again in the same breath as analyzing it.

## The statistician's contribution {#sec-acquisition-human}

Reading a file is mechanical. The judgements:

**The column specification is an assertion, not a convenience.**
Every reader guesses your column types, and the guess is right
until it is not: a sentinel -99 becomes a blood pressure, an
ambiguous date is read under the wrong convention, an identifier
of leading zeros becomes an integer, so that the join in
@sec-joins drops a patient and reports no error. Worse, the
guesses differ between readers, so a file that imports correctly
for you imports differently for a collaborator who reached for a
different function. Writing the `col_types` out in full is fifteen
seconds of typing that converts a silent corruption into a loud
parse failure. Do it on every read of any file you did not create.

**Acquisition is a step, not a preamble.** The instinct is to
treat 'getting the data' as something that happens before the
analysis starts, and therefore something that need not be
scripted to the same standard. This is exactly backwards. The
acquisition determines which rows exist, and no amount of care
downstream can recover a row that was silently dropped at
import. The script that acquires is the first script in the
pipeline and is held to the same standard as the last.

**Freeze what you receive.** A warehouse query run in March and
re-run in June returns different rows, because the warehouse is
alive: patients are added, records are corrected, a duplicate is
merged. If you re-query at render time, your figures change
under you and you cannot tell whether a shifted estimate is a
finding or a data refresh. Write the response to a dated raw
file, commit it if you may, and read *that* in the analysis.
The query is the recipe; the dated file is the ingredient.

**Prefer the interface to the scrape, and the file to neither.**
Scraping is the technique of last resort. It is brittle, it is
often against the terms under which the data are published, and
its output is a guess about someone else's HTML. Before writing
a scraper, ask whether an API exists, whether the data are
published as a file somewhere, and whether you could simply ask
the person who has them. The answer is surprisingly often yes.

These judgements determine whether the data your analysis rests
on are the data you think they are.

## Reading files, deliberately

The whole tidyverse import story is one function per format:
`readr::read_csv()` and its relatives for delimited text,
`readxl::read_excel()` for spreadsheets, `haven::read_sas()` and
`haven::read_xpt()` for the SAS world of @sec-sas. They share an
interface and, more importantly, a philosophy: they parse, they
report what they did, and they never fail silently if you ask
them not to.

Consider a small CRF extract with three of the pathologies that
recur in clinical data: an identifier with leading zeros, a
sentinel value that means 'missing', and a date that a site typed
in the American convention.

```{r}
#| label: acq-csv
#| message: false
library(tidyverse)

extract <- "patient_id,visit_date,sbp,site
007,03/14/2026,148,Durham
012,03/15/2026,-99,Durham
019,03/17/2026,152,Chapel Hill
"

naive <- read_csv(extract, show_col_types = FALSE)
naive
```

Every column parsed without complaint, and the result is wrong in
two ways. The `sbp` of -99 is a sentinel that some upstream system
used to mean 'not recorded', and it is now a blood pressure of
negative ninety-nine, which will pass through `mean()` without a
murmur and drag the estimate down. The `visit_date` is not a date
at all but a string, so it will sort lexicographically, which puts
January of next year before March of this one.

The `patient_id`, on the other hand, survived: `readr` guessed
character and kept the leading zeros. That is a property of
`readr`, not of R, and relying on it is a mistake. The same file
read with base R's reader loses them:

```{r}
#| label: acq-base
f <- tempfile(fileext = ".csv")
writeLines(extract, f)

base_read <- utils::read.csv(f)
str(base_read$patient_id)
```

The identifier is now the integer 7. It will never again match the
string `'007'` in the demographics file, the join of @sec-joins
will silently drop that patient, and nothing will have complained.
Which of your columns survive an unspecified read is therefore a
fact about which function you happened to call, and it will change
under you the day a collaborator uses `read.csv()`, or the file
arrives through `readxl`, or someone upgrades a package. This is
the whole argument for not leaving it to the guess.

The remedy is to say what you expect:

```{r}
#| label: acq-csv-spec
clean <- read_csv(
  extract,
  col_types = cols(
    patient_id = col_character(),          # leading zeros matter
    visit_date = col_date(format = "%m/%d/%Y"),
    sbp        = col_double(),
    site       = col_factor(levels = c("Durham", "Chapel Hill"))
  ),
  na = c("", "NA", "-99")                  # the sentinel, declared
)
clean
```

The identifier survives as a string, the date is a date, the
sentinel is a missing value rather than a number, and the site is
a factor whose levels you have declared, which means that a
fourth site appearing in next month's extract will arrive as `NA`
and be visible rather than silently swelling the level set.

That last property is the one to internalize. A full `col_types`
specification is a contract with the file. When the file changes,
the contract is violated, and `readr` tells you so through the
problems it reports rather than by handing you a data frame that
is quietly different from last month's.

```{r}
#| label: acq-problems
problems(clean)
```

An empty problems table is a passed assertion. On a real extract,
inspect it every time; a non-empty one names the row, the column,
and what it expected.

### The pathologies that do not announce themselves

**The wrong delimiter.** A file named `.csv` that is
tab-separated will read as a single column with commas nowhere in
it. The shell triage of @sec-shell, `head` and `file`, catches
this in three seconds, before R is involved.

**Encodings and the byte-order mark.** A file exported from a
Windows tool may be Latin-1 rather than UTF-8, so that a patient
from Zürich arrives as `Z\xfcrich`, or may carry a byte-order
mark that turns the first column's name into `ï»¿patient_id`, so
that `select(patient_id)` fails on a column you can see with your
own eyes. Declare the encoding (`locale = locale(encoding =
"latin1")`) rather than repairing the strings afterward.

**Excel serial dates.** A date column arriving as five-digit
numbers is Excel's internal representation, and parsing it as a
date yields nonsense. @sec-types has the fix
(`janitor::excel_numeric_to_date()`); the rule here is the
detection: any column whose *name* says date and whose *values*
are numbers is a bug until proven otherwise.

**Trailing whitespace and case.** `'Durham'` and `'Durham '` and
`'durham'` are three sites as far as a join is concerned. The
`janitor::clean_names()` of @sec-wrangling handles the column
names; the values need `str_squish()` and a deliberate decision
about case.

::: {.callout-note icon=false collapse=true}
## Check your understanding: the silent identifier

**Question.** Your cohort has 1,000 patients. You read the
demographics file without a column specification, join it to the
visits file on `patient_id`, and the result has 1,000 rows, which
is what you expected. Everything looks correct. What might still
be wrong?

**Answer.**

The join may have matched on a coerced type. Suppose the visits
file came to you as a CSV read with `readr`, which keeps
`'007'` as a string, while the demographics came from a
collaborator's script that used `read.csv()`, or from an Excel
export, either of which turns it into the integer 7. The join then
fails outright with a type error, which is the *good* case,
because you find out.

The bad case is subtler: if both files were read by a tool that
coerces, both identifiers are now integers, the join succeeds, the
row count is right, and every identifier that originally carried a
leading zero is a different value from the one in the source
system. The analysis is internally consistent and externally
wrong: you can no longer link a row back to the patient it came
from, and the day someone asks you to re-identify a case for a
data query, you cannot.

The row count told you nothing, because it was never sensitive to
the failure. This is the pattern of @sec-testing: a structural
check that passes for the wrong reason. Read identifiers as
character, always, and assert it.

:::

## Querying an interface

Most clinical data now reach the statistician through an
interface rather than a file. REDCap exposes one; so do
institutional warehouses, ClinicalTrials.gov, the FDA's openFDA
endpoints, and PubMed. The pattern is the same in every case: you
send an authenticated HTTP request describing what you want, and
you receive JSON.

The `httr2` package expresses this as a pipeline that mirrors the
structure of the request itself.

```r
library(httr2)

resp <- request("https://redcap.example.edu/api/") |>
  req_body_form(
    token   = Sys.getenv("REDCAP_TOKEN"),   # <1>
    content = "record",
    format  = "json",
    fields  = "patient_id,visit_date,sbp",
    type    = "flat"
  ) |>
  req_retry(max_tries = 3) |>               # <2>
  req_throttle(rate = 30 / 60) |>           # <3>
  req_perform()

raw_json <- resp_body_string(resp)          # <4>
```
1. The token comes from the environment, never from the source.
2. Transient network failures are retried rather than crashing
   the pipeline.
3. At most thirty requests per minute, so that you do not
   inconvenience a shared institutional server.
4. Keep the response as a string; it is about to become a file.

The single most important line is the first. **A credential in a
repository is a credential that has been disclosed**, and a
REDCap token is not a password to an inconvenience, it is
authenticated access to protected health information. It belongs
in `.Renviron`, which is listed in `.gitignore`:

```
# .Renviron  (never committed)
REDCAP_TOKEN=A1B2C3D4E5F6...
```

and the repository ships a `.Renviron.example` naming the
variable without its value, so that a collaborator knows what
they must obtain. The de-identification discipline of
@sec-deident applies to the pipeline as much as to the data, and
a leaked token is the fastest route from a defensible study to an
institutional incident report. If you do commit one, rotating the
token is not optional and deleting the commit is not sufficient.

### From JSON to a data frame

The response is nested, because JSON is. Flattening it is the
step where rows quietly vanish, so it deserves the same care as a
join.

Here is a REDCap-shaped payload, fabricated so that this chapter
renders without a network call. It has the two features that
matter: the values are all strings, because REDCap sends them
that way, and one record is missing a field entirely rather than
sending it as null.

```{r}
#| label: acq-json
#| message: false
library(jsonlite)

payload <- '[
  {"patient_id":"007","visit_date":"2026-03-14","sbp":"148"},
  {"patient_id":"012","visit_date":"2026-03-15","sbp":""},
  {"patient_id":"019","visit_date":"2026-03-17"}
]'

records <- fromJSON(payload, simplifyDataFrame = TRUE) |>
  as_tibble()
records
```

Note what arrived. Every column is character, including the blood
pressure, because the interface does not know or care about your
types. The third record has no `sbp` field at all, and `fromJSON`
has filled it with `NA`, which is the behavior you want but
should verify rather than assume. The empty string in the second
record is *not* `NA` yet; it is a zero-length string that will
become `0` if you coerce it carelessly.

The coercion, therefore, is explicit and is where the column
specification of the previous section reappears in a different
costume:

```{r}
#| label: acq-json-types
visits <- records |>
  mutate(
    patient_id = as.character(patient_id),
    visit_date = as.Date(visit_date),
    sbp        = na_if(sbp, "") |> as.numeric()
  )

visits

stopifnot(
  nrow(visits) == 3,
  is.character(visits$patient_id),
  !anyNA(visits$patient_id)
)
```

The `stopifnot()` is the point. An interface can change its
response shape without telling you, and the assertion is what
turns a silent shape change into a failed pipeline. Assert the
row count you expect, assert the types, and assert that the
identifier is present on every record.

### Pagination

An interface will not hand you a million rows at once. It will
hand you the first thousand and a token for the next page, and a
loop that forgets to follow the token will produce an analysis of
the first thousand patients that reports itself as an analysis of
all of them. This is the acquisition-layer version of the silent
row loss that @sec-joins warns about, and it is more dangerous,
because there is no second table to compare against.

```r
fetch_all <- function(base_query) {
  out  <- list()
  page <- 1
  repeat {
    resp <- request(base_query) |>
      req_url_query(page = page, page_size = 1000) |>
      req_perform() |>
      resp_body_json()

    if (length(resp$records) == 0) break     # <1>
    out[[page]] <- resp$records
    if (is.null(resp$next_page)) break       # <2>
    page <- page + 1
  }
  bind_rows(out)                             # <3>
}
```
1. Stop on an empty page.
2. Stop when the interface says there is no next page. Trusting
   only the empty-page test will loop forever against an API that
   repeats its last page.
3. One data frame, whose row count you now assert against the
   total the interface reported.

Whatever the pagination scheme, verify the total. Most interfaces
report a record count in the response or a header; compare it to
`nrow()` and fail loudly if they differ.

## Scraping, and whether to

Sometimes the data are on a web page and nowhere else: a
regulator's table of approvals, a registry's list of trials, an
institution's directory. `rvest` will get them, and the mechanics
are straightforward. The prior question is not.

**Can you?** is a technical question and the answer is usually
yes. **Should you?** is the question that matters, and it has
three parts. Does the site's `robots.txt` permit it? Do the terms
of service? And would the volume and rate of your requests impose
a cost on someone else's server that they did not agree to bear?
A scraper that hammers a small registry's server at fifty
requests a second is a denial-of-service attack with good
intentions. The `polite` package encodes the etiquette (check
`robots.txt`, identify yourself with a user agent that names you,
rate-limit, cache), and it is the right default.

Ask also whether the data are personal. A page listing the
attendees of a support group is public in the sense that anyone
can read it, and aggregating it into a dataset is a different act
with different consequences. That the data are technically
accessible does not make their aggregation ethical, and the fact
that a website has published something is not a license for you
to redistribute it.

When the answer to all of this is yes, the mechanics are these.
The example below parses a literal HTML fragment rather than
fetching a page, so that the chapter renders offline and so that
no server is touched by a build of this book.

```{r}
#| label: acq-scrape
#| message: false
library(rvest)

page <- minimal_html('
  <html><body>
    <h1>Site enrollment</h1>
    <table class="enrollment">
      <tr><th>Site</th><th>Enrolled</th><th>Target</th></tr>
      <tr><td>Durham</td><td>142</td><td>150</td></tr>
      <tr><td>Chapel Hill</td><td>118</td><td>150</td></tr>
      <tr><td>Raleigh</td><td>97</td><td>150</td></tr>
    </table>
  </body></html>
')

enrollment <- page |>
  html_element("table.enrollment") |>
  html_table() |>
  janitor::clean_names()

enrollment
```

`html_element()` takes the first match for a CSS selector and
`html_elements()` takes all of them; `html_table()` turns an HTML
table into a data frame. Selecting by a class (`table.enrollment`)
rather than by position (`table` and hope) is what makes the
scraper survive the addition of a second table to the page. It
will not survive a redesign, and nothing will, which is why the
next section matters more than this one.

## The discipline that makes any of this reproducible

An acquisition that runs at render time is a reproducibility bug
wearing a convenience costume. If your Quarto document queries
the warehouse when it renders, then the document produces
different numbers on different days, your collaborator cannot
render it at all without a token, and your continuous-integration
run of @sec-ci fails or, worse, succeeds against different data.

The pattern is two scripts and one dated file.

```r
# scripts/01-acquire.R  (run deliberately, by a human, rarely)
source("R/redcap.R")

raw <- fetch_redcap(Sys.getenv("REDCAP_TOKEN"))

stopifnot(nrow(raw) > 0)

stamp <- format(Sys.Date(), "%Y%m%d")
write_rds(raw, glue::glue("data/raw/redcap_{stamp}.rds"))   # <1>
write_lines(                                                # <2>
  glue::glue("Extracted {nrow(raw)} records on {Sys.Date()} \\
              from {Sys.getenv('REDCAP_URL')}"),
  glue::glue("data/raw/redcap_{stamp}.README")
)
```
1. Dated, so that a later extraction does not overwrite the one
   your published numbers rest on.
2. A provenance note beside it: how many rows, from where, when.
   Six months from now this file is the only thing that can tell
   you which extraction Table 2 came from.

```r
# scripts/02-clean.R  (runs on every render; touches no network)
raw <- read_rds("data/raw/redcap_20260314.rds")   # <1>
```
1. The date is written into the analysis, deliberately. When you
   refresh the extract, you change this line in a commit that
   says so, and the change in the numbers is attributable to the
   refresh rather than mysterious.

Whether the dated raw file can be committed depends on what is in
it. Protected health information cannot go in a repository
(@sec-deident, @sec-federal), and in that case the compendium
holds the acquisition script, the provenance note, and a
synthetic file of the same shape so that a reviewer without
access can still run the pipeline. This is exactly the
arrangement @sec-adni describes for ADNI, and the reason it
generalizes is that the constraint is not about ADNI: it is about
any dataset you are not free to redistribute.

## Collaborating with an LLM on acquisition

Models write `readr`, `httr2`, and `rvest` fluently. The failure
mode is uniform across all three: they produce code that runs,
against data they have not seen.

**Prompt 1: writing the column specification.** Paste the first
twenty lines of the file and ask for a full `col_types`
specification with the sentinel values declared.

*What to watch for.* The model infers types from the sample you
pasted, exactly as `read_csv()` would, so it inherits the same
blindness: it will make an identifier an integer if the first
twenty rows contain no leading zeros. Tell it which columns are
identifiers. Ask it to justify each type.

*Verification.* Read the file with the specification, then check
`problems()` and the distinct values of every factor column
against what you expected.

**Prompt 2: parsing a JSON response.** Paste one record of the
response and ask for the code that flattens the whole thing to a
tibble.

*What to watch for.* Whether it handles the record that is
missing a field, and whether it distinguishes an empty string
from a missing value. A model given one well-formed record will
write code that assumes every record is well-formed, and the one
that is not is precisely the one you needed to know about.

*Verification.* Assert the row count against the total the
interface reports, and assert that no identifier is `NA`.

**Prompt 3: 'write me a scraper for this page'.** The model will
comply, immediately and without asking the question you should
have asked.

*What to watch for.* It will not raise `robots.txt`, the terms of
service, rate limiting, or whether the data are personal, because
you did not ask and it is agreeable. The judgement is yours and
the model will not prompt you for it. Ask it explicitly to argue
the case *against* scraping the page, which is a prompt it
handles well and which surfaces the considerations you should
have weighed.

*Verification.* Run the scraper once, on one page, and inspect
what came back before you point it at anything larger.

## Principle in use

In closing, three habits define what we would regard as defensible acquisition:

1. **Specify every column.** The type specification is an
   assertion about the file, and the file will eventually violate
   it. Better a loud parse failure than a silent coercion.
2. **Freeze what you receive.** Acquire deliberately, to a dated
   raw file with a provenance note beside it. Never acquire and
   analyze in the same breath, and never over the network at
   render time.
3. **Keep credentials out of the repository, and ask whether you
   should scrape before asking how.** The token belongs in
   `.Renviron`; the ethics belong to you and not to the tool.

## Exercises

1. Take a delimited file you did not create. Read it once with no
   arguments and once with a full `col_types` specification.
   Diff the two results. Document every column on which they
   disagree and say which is correct.
2. Construct a small CSV containing each of the pathologies in
   this chapter (leading-zero identifier, sentinel missing value,
   American-format date, trailing whitespace in a factor level, a
   byte-order mark). Write a read that handles all five and an
   assertion suite that would fail if any of them regressed.
3. Query any public API with `httr2` (ClinicalTrials.gov and
   openFDA both permit anonymous access). Write the response to a
   dated raw file with a provenance note. Then write a second
   script that reads the dated file and touches no network, and
   confirm the second runs with the network disconnected.
4. Take the paginated fetch in this chapter and break it
   deliberately, by removing the `next_page` check. Point it at an
   interface with more than one page of results and show that the
   result is silently incomplete. Write the assertion that would
   have caught it.
5. Find a web page with a table you would like as data. Before
   writing any code, write a paragraph answering: does
   `robots.txt` permit this, do the terms of service, is an API
   available instead, are the data personal, and would your
   request rate cost the host anything. Only then write the
   scraper, or, having answered honestly, do not.

## Further reading

- [@wickham2023r4ds], the data-import and web-scraping chapters,
  the canonical applied treatment of `readr` and `rvest`.
- The `httr2` documentation at `httr2.r-lib.org`, in particular
  its articles on authentication and on wrapping an API.
- The `polite` package, which encodes the scraping etiquette this
  chapter argues for as a set of defaults.
- @sec-databases, for the case where the data live in a warehouse
  and the query, not the file, is the provenance.
- @sec-deident and @sec-federal, for what you may keep, commit,
  and redistribute once you have it.

## Prerequisites answers {#sec-quiz-answers-12b}

1. The guess is made from a sample (the first thousand rows by
   default), so any pathology that does not appear in the sample
   is invisible to it. Two consequences that produce no error.
   First, a sentinel value used upstream to mean 'missing', such
   as -99 or 9999, is guessed as a number and enters the analysis
   as a real measurement, dragging every mean it touches. Second,
   a column of dates in an ambiguous format is guessed as
   character and then sorts lexicographically, or, worse, is
   guessed as a date under the wrong convention, so that
   03/04/2026 is March in one file and April in another.

   A third hazard is worth stating precisely, because it is
   commonly misdescribed: whether an identifier with leading
   zeros survives depends on *which reader you call*. Modern
   `readr` guesses character and keeps `'007'`; base R's
   `read.csv()` and most spreadsheet exports coerce it to the
   integer 7. Relying on your reader's good behavior is
   therefore relying on a fact about your tooling rather than
   about your data, and it will break the day a collaborator uses
   a different function.

   The remedy in every case is the same: pass a full `col_types`
   specification and an explicit `na` argument, treat both as
   assertions about the file, and check `problems()` after every
   read.
2. Commit the acquisition script, the query it sends, and a
   provenance note recording how many records came back, from
   which endpoint, on which date; and commit the dated raw file
   itself if, and only if, you are permitted to redistribute its
   contents. Do not commit the credential: the API token belongs
   in `.Renviron`, which is gitignored, with a
   `.Renviron.example` naming the variable so a collaborator
   knows what to obtain. Do not commit protected health
   information; ship a synthetic file of the same shape instead,
   so that the pipeline can be run by a reviewer who lacks
   access. And do not have the analysis query the interface at
   render time, because a living system returns different rows on
   different days and your figures would change under you.
3. Before writing the scraper: does an API or a published file
   exist that would make the scrape unnecessary; does the site's
   `robots.txt` permit it; do the terms of service; are the data
   personal, such that aggregating them causes a harm that
   publishing them individually did not; and would your request
   rate impose a cost on the host. Once you have the data, the
   first thing to do is write them to a dated raw file with a
   provenance note, so that the analysis never scrapes again. The
   page will change, and when it does you want the failure to be
   an explicit re-acquisition rather than a figure that silently
   changed.
