# Databases and SQL for Biostatisticians {#sec-databases}

::: {.bookepigraph}
```{=latex}
\begin{flushright}\small
```
*The most valuable commodity I know of is information.*

[Gordon Gekko]{.smallcaps}, *Wall Street* (1987)
```{=latex}
\end{flushright}
```
:::

::: {.callout-note}
## Why this chapter exists

Clinical data rarely arrive as a tidy CSV. They live in a
REDCap project, an institutional data warehouse, or an
electronic health record extract, and the biostatistician
reaches them through a query rather than a file. The
survey of peer programmes found relational databases and
SQL taught as a core skill in at least seven of the
comparison courses (Berkeley 243, UNC 735, Michigan 625,
Minnesota 8054, the Johns Hopkins data-science sequence,
and both Software and Data Carpentry). This chapter brings
the Practicum to parity.
:::

## Prerequisites

Answer the following questions to see if you can bypass
this chapter. You can find the answers at the end of the
chapter in @sec-quiz-answers-15b.

1. What is the difference between querying a table with
   `DBI::dbGetQuery()` and reading a whole table into
   memory with `dbReadTable()`, and why does the
   distinction matter for a ten-million-row table?
2. In the `dbplyr` model of database access, what does a
   pipeline of `dplyr` verbs actually do before you call
   `collect()`, and where does the computation happen?
3. Why is it preferable, for reproducibility, to record
   the SQL query that produced an analytic dataset rather
   than to save the extracted CSV alone?

## Learning objectives

By the end of this chapter you should be able to:

- Connect to a database from R with `DBI` and an
  appropriate backend driver (`RSQLite`, `duckdb`,
  `RPostgres`).
- Write basic `SELECT` queries with filtering, joining,
  grouping, and aggregation.
- Express the same query with `dbplyr` using familiar
  `dplyr` verbs, and inspect the SQL it generates.
- Distinguish lazy from eager evaluation and use
  `collect()` deliberately.
- Reason about when to push work to the database and when
  to pull data into R.

## Orientation

A database is a store that lets many users query
structured data efficiently and safely. For the
biostatistician, the practical point is that the data are
usually too large, too sensitive, or too frequently
updated to hand around as flat files. Querying at the
source has three advantages: you pull only the rows and
columns you need, you leave the authoritative copy
untouched, and you record, in the query itself, exactly
how your analytic dataset was derived.

This chapter is not a database course. It teaches the
subset of SQL and the `DBI`/`dbplyr` tooling that covers
the recurring needs of applied biostatistics: connect,
filter, join, aggregate, and extract. The deeper matters
of schema design, indexing, and transaction management
belong to a database text and to your data-engineering
colleagues.

## The statistician's contribution {#sec-databases-human}

The query mechanics are mechanical. The judgements:

**Query at the source, extract deliberately.** The
temptation is to pull the whole table into R and work
there, because R is where you are comfortable. On a small
table this is harmless. On a large one it is slow,
memory-hungry, and it obscures the derivation. Push the
filtering and aggregation to the database; pull only the
analytic dataset.

**The query is the provenance.** A CSV named
`analytic_data_final_v3.csv` records nothing about how it
was made. The SQL that produced it records everything.
Keep the query in the compendium, under version control,
and treat the extracted file as a build product in the
manner of chapter 13's raw-data rule.

**Beware the silent row multiplication.** A join in SQL
inflates rows exactly as a join in `dplyr` does, and the
database will not warn you. The verification habit from
chapter 15, count the rows and check against
expectation, applies with equal force here.

**Respect the boundary between analyst and warehouse.**
You usually have read-only access to an institutional
warehouse, and that is correct. You are a consumer of the
data, not its custodian. Do not write back; extract, and
do your derivations in your own space.

These judgements are what make database-backed analysis
reproducible and safe.

## Connecting with DBI

The `DBI` package provides a backend-independent interface;
a driver package supplies the connection to a specific
database engine. For teaching we use `RSQLite`, which
gives a complete SQL database in a single file or in
memory, with no server to configure. The same code works
against PostgreSQL (`RPostgres`), DuckDB (`duckdb`), or
SQL Server (`odbc`) by changing only the `dbConnect` line.

To make the example self-contained, we create an in-memory
database and populate it with a tiny clinical schema: a
one-row-per-patient demographics table and a
one-row-per-visit table of blood-pressure measurements.

```{r}
#| label: db-setup
#| message: false
library(DBI)
library(RSQLite)
library(dplyr)
library(dbplyr)

con <- dbConnect(RSQLite::SQLite(), ":memory:")

demographics <- data.frame(
  patient_id = 1:5,
  age        = c(64, 71, 58, 66, 73),
  arm        = c("Active", "Placebo", "Active", "Placebo", "Active")
)

visits <- data.frame(
  patient_id = c(1, 1, 2, 3, 3, 3, 4, 5),
  visit      = c(1, 2, 1, 1, 2, 3, 1, 1),
  sbp        = c(140, 138, 150, 145, 143, 141, 139, 152)
)

dbWriteTable(con, "demographics", demographics)
dbWriteTable(con, "visits", visits)

dbListTables(con)
```

In a real setting the `dbConnect` call would name a
server, a database, and credentials, ideally drawn from an
environment variable or a `config` file rather than
written into the script:

```r
con <- dbConnect(
  RPostgres::Postgres(),
  host     = "warehouse.example.edu",
  dbname   = "clinical",
  user     = Sys.getenv("DB_USER"),
  password = Sys.getenv("DB_PASS")
)
```

## Querying with SQL

The `SELECT` statement is the workhorse. It names the
columns, the source table, and optional filtering,
joining, grouping, and ordering. Here we ask for the mean
systolic blood pressure and the number of visits per
treatment arm, joining the two tables on `patient_id`.

```{r}
#| label: db-sql
dbGetQuery(con, "
  SELECT d.arm,
         COUNT(*)     AS n_visits,
         AVG(v.sbp)   AS mean_sbp
  FROM visits AS v
  JOIN demographics AS d
    ON v.patient_id = d.patient_id
  GROUP BY d.arm
")
```

The clauses read in the order the database applies them:
`FROM` and `JOIN` assemble the rows, `WHERE` (absent here)
filters them, `GROUP BY` partitions them, the aggregate
functions summarise each partition, and `SELECT` names
what comes back. `dbGetQuery()` returns the result as an
ordinary data frame, which is what you want for a small
summary. For a large intermediate result you would not use
it, for the reason discussed next.

## Querying with dbplyr

Writing SQL by hand is a skill worth having, but for most
extractions you can stay in `dplyr`. The `dbplyr` package
translates a `dplyr` pipeline into SQL, sends it to the
database, and leaves the result there until you ask for
it. A reference to a database table behaves almost exactly
like a tibble.

```{r}
#| label: db-dbplyr
visits_tbl <- tbl(con, "visits")
demo_tbl   <- tbl(con, "demographics")

summary_q <- visits_tbl |>
  inner_join(demo_tbl, by = "patient_id") |>
  group_by(arm) |>
  summarise(n_visits = n(),
            mean_sbp = mean(sbp, na.rm = TRUE))

summary_q
```

The object `summary_q` is *lazy*: no computation has yet
crossed into R. The pipeline has been recorded, and when
printed it shows only a preview fetched from the database.
To see the SQL that `dbplyr` generated, ask for it:

```{r}
#| label: db-showquery
summary_q |> show_query()
```

The generated SQL is the same query we wrote by hand,
which is the point: `dbplyr` lets you think in `dplyr` and
still push the work to the database. When you are ready to
bring the result into R, for plotting, modelling, or
anything the database cannot do, call `collect()`.

```{r}
#| label: db-collect
analytic <- summary_q |> collect()
analytic
```

Everything up to `collect()` ran in the database; only the
small summary crossed into memory. This is the pattern to
internalise: filter and aggregate lazily against the
table, then `collect()` the modest result. The
verification habit still applies, so we check the row
count is what a per-arm summary should give.

```{r}
#| label: db-verify
stopifnot(nrow(analytic) == n_distinct(demographics$arm))
```

```{r}
#| label: db-disconnect
#| echo: false
dbDisconnect(con)
```

::: {.callout-note icon=false collapse=true}
## Check your understanding: where does the work happen?

**Question.** You write
`tbl(con, "labs") |> filter(test == "HbA1c") |> summarise(m = mean(value))`
against a table of fifty million laboratory results, then
call `collect()`. How much data crosses from the database
into R, and what would have happened had you called
`collect()` before the `filter`?

**Answer.**

With `collect()` at the end, only the single summarised
value crosses into R: the database does the filtering and
the averaging, and returns one number. Had you called
`collect()` first, all fifty million rows would have been
pulled into memory before R did any work, which on most
laptops would exhaust RAM and fail. The rule is to
`collect()` as late as possible, after the database has
reduced the data to the size you actually need.

:::

## When to push, when to pull

The database is good at set operations on large data:
filtering, joining, grouping, aggregating. R is good at
everything else: modelling, plotting, iteration, anything
statistical. The dividing line is `collect()`. Push the
data-reduction work to the database; pull the reduced
result into R for the analysis. A useful test: if a step
makes the data smaller (a filter, an aggregate), it
usually belongs before `collect()`; if a step needs the
full statistical machinery of R (fitting a mixed model,
drawing a figure), it belongs after.

Two cautions. First, not every R function has a SQL
translation; `dbplyr` will tell you when it cannot
translate a verb, and then you must either rewrite it or
`collect()` earlier. Second, a query that is cheap to
write can be expensive to run; on a shared warehouse, a
careless cross-join can inconvenience every other user, so
test on a `LIMIT`-ed subset before running the full
extraction.

## Reproducibility of the extraction

The analytic dataset is a build product, and the query is
its recipe. Keep the query in the compendium as a `.sql`
file or an R script, so that the derivation is
version-controlled and re-runnable. Record the date of
extraction and, where the warehouse supports it, the data
snapshot or freeze identifier, because a warehouse is a
living system and today's query may return different rows
next month. The pattern mirrors the raw-data discipline of
chapter 13: the query and its output date are the audit
trail that lets a reviewer reconstruct exactly which rows
your analysis rested on.

## Collaborating with an LLM on databases

LLMs write competent SQL and `dbplyr`; the traps are
correctness on joins and the silent scale of a query.

**Prompt 1: drafting a query.** Describe the tables, their
keys, and the summary you want, and ask for both the SQL
and the `dbplyr` equivalent.

*What to watch for.* The join key and the join type. An
LLM will happily write an inner join where you needed a
left join, dropping patients with no visits. State the
cardinality you expect.

*Verification.* Run against a `LIMIT`-ed subset; count the
rows; confirm the count matches the expected number of
groups or patients.

**Prompt 2: translating hand-written SQL to `dbplyr`.**
Paste a `SELECT` statement and ask for the `dplyr`
pipeline that generates it.

*What to watch for.* Aggregations and `GROUP BY`
translate cleanly; window functions and correlated
subqueries less so. Compare `show_query()` on the result
against the original.

*Verification.* Confirm the two produce identical results
on a small table before trusting the translation on the
full one.

## Principle in use

Three habits define defensible database work:

1. **Query at the source; `collect()` late.** Push
   filtering and aggregation to the database; pull only the
   analytic result into R.
2. **The query is the provenance.** Version-control the
   SQL that produced the analytic dataset; treat the
   extract as a build product with a recorded date.
3. **Verify joins in SQL as in R.** Count the rows; a
   silent many-to-many inflates the data in a database
   exactly as it does in `dplyr`.

## Exercises

1. Create an in-memory SQLite database with the
   `demographics` and `visits` tables from this chapter.
   Write a SQL query that returns, per patient, the number
   of visits and the change in systolic blood pressure
   from the first to the last visit.
2. Express the same query with `dbplyr`. Use
   `show_query()` to inspect the generated SQL and compare
   it to your hand-written version.
3. Add a third table, `labs`, with columns `patient_id`,
   `test`, and `value`. Write a `dbplyr` pipeline that
   returns the mean HbA1c per treatment arm, joining all
   three tables. Verify the row count.
4. Deliberately write a query that produces a
   many-to-many join between `visits` and a table with
   duplicate `patient_id` values. Observe the inflated row
   count and explain it.
5. Using the `duckdb` backend instead of `RSQLite`, read a
   large CSV file directly with `duckdb`'s
   `read_csv_auto` and run an aggregation without ever
   loading the file into R. Compare the memory use to
   `readr::read_csv()` followed by `dplyr`.

## Further reading

- The `DBI` documentation at `dbi.r-dbi.org`, the
  interface reference.
- The `dbplyr` documentation at `dbplyr.tidyverse.org`,
  including the list of translatable verbs.
- [@wickham2023r4ds], the databases chapter, applied
  introduction to `DBI` and `dbplyr`.
- The DuckDB documentation at `duckdb.org`, for
  file-backed analytical queries without a server.

## Prerequisites answers {#sec-quiz-answers-15b}

1. `dbGetQuery()` sends a query and returns only its
   result; a `SELECT` with filtering and aggregation
   returns a small summary regardless of table size.
   `dbReadTable()` reads an entire table into memory, so
   on a ten-million-row table it pulls all ten million
   rows across, which is slow and may exhaust memory. Use
   `dbGetQuery()` (or lazy `dbplyr`) to let the database do
   the reduction.
2. A `dbplyr` pipeline builds up a query lazily: each verb
   adds to a SQL statement that has not yet run. No
   computation crosses into R until you call `collect()`
   (or print, which fetches a preview). The computation
   happens in the database; R receives only the collected
   result.
3. The extracted CSV records the rows but not their
   derivation. The SQL query records exactly which tables,
   filters, joins, and aggregations produced the analytic
   dataset, so a reviewer can reconstruct or re-run it.
   A warehouse is also a living system, so pairing the
   query with an extraction date documents the snapshot the
   analysis rested on.
