# Plotting with ggplot2 and purrr {#sec-graphics}

::: {.bookepigraph}
```{=latex}
\begin{flushright}\small
```
*The greatest value of a picture is when it forces us to
notice what we never expected to see.*

[John W. Tukey]{.smallcaps}, *Exploratory Data Analysis*
(1977)
```{=latex}
\end{flushright}
```
:::

::: {.callout-note}
## Sources

Stat 545 Part VII; blog posts `16-plotsfrompurrr`,
`38-tableplacementrmarkdown`; `zzlongplot` for
longitudinal examples (@sec-adni).
:::

## Prerequisites

Answer the following questions to see if you can bypass
this chapter. You can find the answers at the end of the
chapter in @sec-quiz-answers-16.

1. How do you produce one plot per level of a
   categorical variable, returning the results as a
   named list of `ggplot` objects?
2. What does `patchwork::wrap_plots()` provide that
   `facet_wrap()` does not?
3. Why might you save a figure as PDF rather than PNG
   for a journal submission, and when would you choose
   PNG instead?

## Learning objectives

By the end of this chapter you should be able to:

- Build publication-quality `ggplot2` figures for
  continuous, categorical, and time-varying outcomes.
- Generate many plots programmatically with
  `purrr::map()` over a list of groups.
- Compose multi-panel figures with `patchwork` and
  arrange shared legends cleanly.
- Produce PDF and PNG outputs at appropriate DPIs for
  journal and Word submissions.
- Build a house theme via a `theme_*()` function and
  reuse it across all figures in a project.
- Recognize when to reach for `zzlongplot` and
  `zztable1` rather than rolling your own.

## Orientation

A figure that survives peer review is a figure whose
data, encoding, and aesthetic choices all serve the
single claim the figure exists to make. This chapter
covers the mechanical skills; the principles of
effective visualization are covered in the companion
textbook (chapters 15--16 of *Statistical Computing in
the Age of AI*).

The combination `ggplot2` + `purrr` + `patchwork` is
the modern stack for going from a single exploratory
plot to a polished multi-panel figure: ggplot2 for
the plot, purrr for iteration over groups, patchwork
for composition.

```{r}
#| label: setup-graphics
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(patchwork)
library(palmerpenguins)
library(broom)
penguins_clean <- na.omit(penguins)
```

## The statistician's contribution {#sec-graphics-human}

Software handles the mechanics. The judgements:

**One plot, one message.** A figure trying to convey
three relationships succeeds at none. Three figures
each conveying one are usually clearer. Resist the
'pack everything into one figure to save space'
impulse.

**Encode the claim, not the data.** A scatter plot
with 5,000 points obscures any pattern. The same
relationship as a hex-binned plot or a 2D density
contour shows the structure without the visual
noise. Choose encoding for legibility, not
faithfulness to the raw data.

**Consistency across the project.** All figures in a
paper should share fonts, color palette, and theme
defaults. A house theme function applied via
`theme_set()` makes consistency the default. Without
it, ggplot2 produces nine variants of gray.

**Display uncertainty.** A point estimate without a
CI invites over-trust. A regression line without a
confidence band the same. The tools are simple
(`geom_errorbar`, `se = TRUE` on `geom_smooth`); the
discipline is the analyst's.

**Ask what the reader's eye is measuring.** A bar
encodes length, so a truncated baseline makes a small
difference fill the panel; a map encodes area, and land
does not have a disease; a bubble's radius is read as
area, so a radius proportional to a count exaggerates
it. None of these prints a false number, and all of
them mislead. @sec-ethics treats the failure modes and
their remedies at length; the habit to carry into every
figure is to name the visual channel the reader is
decoding before choosing it.

These judgements are what make figures publishable.

## One plot per group with purrr

The pattern: split data by a grouping variable, apply
a plotting function to each group, collect into a
named list.

```{r}
#| label: fig-purrr-adelie
#| fig-cap: "One plot per species, produced by mapping a plotting
#|   function over a split data frame. Only the Adelie panel is shown;
#|   `plots` holds all three."
#| fig-width: 6
#| fig-height: 4
library(tidyverse)
library(palmerpenguins)

penguins_clean <- na.omit(penguins)

# split by species
plots <- penguins_clean |>
  split(penguins_clean$species) |>
  imap(\(d, name) {
    ggplot(d, aes(flipper_length_mm, body_mass_g)) +
      geom_point() +
      geom_smooth(method = "lm") +
      labs(title = name,
           x = "Flipper length (mm)", y = "Body mass (g)") +
      theme_minimal()
  })

# inspect one
plots$Adelie
```

The same iteration writes every panel to disk. This
is a side effect, so we reach for `walk2` rather than
`map2`:

```r
# save all three panels as publication-quality PDFs
walk2(plots,
      paste0("figures/penguin-", names(plots), ".pdf"),
      \(p, path) ggsave(path, p, width = 6, height = 4))
```

Three things to internalize:

1. **`split()` returns a named list.** The names
   propagate through `imap` and `walk2`.
2. **`imap` and `walk2` use both the value and the
   name.** Useful for titles, file paths, and
   identifying which plot is which.
3. **`walk2` for side effects.** Returns invisibly;
   `map2` would build a list of `ggsave`'s NULL
   returns.

For a `dplyr`-native version:

```r
penguins_clean |>
  group_by(species) |>
  group_map(\(d, k) {
    ggplot(d, aes(flipper_length_mm, body_mass_g)) +
      geom_point() +
      labs(title = k$species)
  })
# returns a list of plots, with the group keys in `k`
```

## Multi-panel figures with patchwork

`patchwork` composes distinct plots into one figure.
Where `facet_wrap` produces small multiples of the
same plot, `wrap_plots` (or the `+` and `/`
operators) combines independent plots that may
differ in everything.

```{r}
#| label: fig-patchwork
#| fig-cap: "Three distinct views of the penguin data composed into one
#|   figure with patchwork. The species legend is collected once and
#|   placed at the bottom."
#| fig-width: 7.5
#| fig-height: 6
library(patchwork)

p1 <- ggplot(penguins_clean,
             aes(flipper_length_mm, body_mass_g, color = species)) +
        geom_point() +
        labs(title = "A. Body mass vs flipper length")

p2 <- ggplot(penguins_clean,
             aes(species, body_mass_g, fill = species)) +
        geom_boxplot() +
        labs(title = "B. Body mass by species") +
        guides(fill = "none")     # hide redundant fill legend

p3 <- ggplot(penguins_clean,
             aes(bill_length_mm, bill_depth_mm, color = species)) +
        geom_point() +
        labs(title = "C. Bill morphology")

# 2x2 grid with shared legend at bottom
(p1 + p2) / (p3 + plot_spacer()) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

Operators:

- `+` juxtaposes side by side.
- `/` stacks vertically.
- `|` is equivalent to `+` (horizontal).
- `&` applies a theme to every plot in the
  composition.
- `plot_layout(guides = "collect")` collects
  duplicate legends into one.

For more control, `plot_layout` takes `widths`,
`heights`, `nrow`, `ncol`, `byrow`, and other
arguments.

`facet_wrap` for small multiples of the same plot;
`patchwork` for distinct plots in one figure.

::: {.callout-note icon=false collapse=true}
## Check your understanding: facet vs. patchwork

**Question.** You have one dataset and want to show
'distribution of body mass' as a histogram and
'distribution of body mass by species' as boxplots,
side by side. Should you use `facet_wrap`, `patchwork`,
or both?

**Answer.**

`patchwork`. The two panels show *different views* of
the same data: a histogram (one geom) and boxplots
(a different geom). `facet_wrap` shows the same plot
across panels conditioned on a variable; both panels
must use the same geom and aesthetic mapping. Build
each plot separately and combine with `p1 + p2`.

```r
p_hist <- ggplot(d, aes(body_mass_g)) + geom_histogram()
p_box  <- ggplot(d, aes(species, body_mass_g)) + geom_boxplot()
p_hist + p_box
```

If the second plot was 'histogram of body mass for
each species' (same geom, conditioned on species),
`facet_wrap(~species)` would be the right tool.

:::

## House themes

A function that returns a `theme()` object, applied
once per session:

```{r}
#| label: theme-practicum
theme_practicum <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title         = element_text(face = "bold"),
      plot.subtitle      = element_text(color = "gray40"),
      axis.title         = element_text(face = "bold"),
      axis.text          = element_text(color = "gray20"),
      panel.grid.minor   = element_blank(),
      legend.position    = "bottom",
      strip.background   = element_rect(fill = "gray95", color = NA),
      strip.text         = element_text(face = "bold")
    )
}

# apply globally
theme_set(theme_practicum())

# update geom defaults to match
update_geom_defaults("point", list(size = 1.5, alpha = 0.7))
update_geom_defaults("line",  list(linewidth = 0.6))
```

For project consistency, define `theme_practicum()`
in a setup chunk at the top of every analysis file,
or in a project-level helper that gets sourced. Every
plot then matches without per-plot theme calls.

For palettes, define and reuse:

```r
practicum_palette <- c(
  "Adelie"    = "#1f4e79",
  "Chinstrap" = "#9d2235",
  "Gentoo"    = "#2e8b57"
)

scale_color_practicum <- function(...)
  scale_color_manual(values = practicum_palette, ...)

# in plots
ggplot(...) + ... + scale_color_practicum()
```

For color-blind safety, verify the palette with the
Coblis simulator
(`colorbrewer2.org/learnmore/colorblind-simulator.html`)
or `colorBlindness::cvdPlot()`.

## Export formats

`ggsave()` writes to many formats, picked by
extension:

```r
# vector for journal submission
ggsave("figure1.pdf", plot = p, width = 6, height = 4,
       device = cairo_pdf)

# raster for Word, slides, web
ggsave("figure1.png", plot = p, width = 6, height = 4,
       dpi = 300)

# vector for the web (preserves quality on zoom)
ggsave("figure1.svg", plot = p, width = 6, height = 4)
```

DPI considerations:

- 72 DPI: web display only.
- 96 DPI: monitor default.
- 300 DPI: print quality (the standard for journal
  submission as PNG).
- 600 DPI: high-quality scientific figures.

For LaTeX submission, prefer PDF (vector). For Word,
prefer PNG at 300 DPI. For web, SVG (vector).

`device = cairo_pdf` ensures non-default fonts are
embedded so the PDF renders correctly on a system
that lacks the font.

For journals, check size requirements (typically
column widths of 85--90 mm or 170--180 mm) and produce
figures at exactly the target size, not larger.

## Integrating with `zzlongplot` and `zztable1`

The in-house `zzlongplot` package provides
opinionated longitudinal-study plots: spaghetti plots
with confidence ribbons, mean trajectories with
group-specific overlays, ICC-aware error bars. The
defaults match this practicum's conventions.

```r
library(zzlongplot)
spaghetti(adni_long, time = "VISIT", value = "ADAS",
          subject = "RID", group = "DX")
```

When to reach for `zzlongplot`:

- Longitudinal data with a standard time variable.
- You want consistent styling with other figures in
  this practicum.
- You do not want to write fifty lines of ggplot2
  every time.

When to write your own ggplot2:

- Custom encoding the package does not support.
- Cross-sectional or non-longitudinal data.
- You need a non-standard layout that fights the
  package's defaults.

The `zztable1` package similarly handles Table 1
generation; covered in the analysis-plan chapter
(@sec-analysis-plan).

## Worked example: regression diagnostic figure

```{r}
#| label: fig-diagnostics
#| fig-cap: "A composed regression-diagnostic figure: residuals versus
#|   fitted, a normal Q-Q plot, and Cook's distance, all sharing the
#|   house theme."
#| fig-width: 7
#| fig-height: 6
theme_set(theme_practicum())

fit <- lm(body_mass_g ~ flipper_length_mm + species,
          data = na.omit(penguins))
diag <- augment(fit)

p_resid <- ggplot(diag, aes(.fitted, .resid)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE,
              color = "red", linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "A. Residuals vs. fitted",
       x = "Fitted (g)", y = "Residual")

p_qq <- ggplot(diag, aes(sample = .std.resid)) +
  geom_qq(alpha = 0.5) +
  geom_qq_line(color = "red") +
  labs(title = "B. Normal Q-Q",
       x = "Theoretical", y = "Standardized residual")

p_cook <- ggplot(diag, aes(seq_len(nrow(diag)), .cooksd)) +
  geom_col(width = 0.6) +
  geom_hline(yintercept = 4 / nrow(diag),
             linetype = "dashed", color = "red") +
  labs(title = "C. Cook's distance",
       x = "Observation", y = "Cook's distance")

(p_resid + p_qq) / p_cook +
  plot_layout(heights = c(1, 0.7))
```

To write the composed figure to disk for a manuscript,
add a `ggsave` call after building it:

```r
ggsave("figures/diagnostics.pdf", width = 7, height = 6,
       device = cairo_pdf)
```

The composition reads naturally: top row two related
diagnostics, bottom row a single observation-level
diagnostic. House theme applied via
`theme_set(theme_practicum())`. Saved as a
publication-quality PDF.

## Worked example: a longitudinal trajectory figure

The penguin data are cross-sectional, and cross-sectional
scatter plots are the easy case. The figure a
biostatistician is asked for most often is harder: the
trajectory of a repeated-measures outcome over time, by
treatment arm, with the individual patients visible
behind the group means. To illustrate, we simulate a
small two-arm trial in which a cognitive score (higher
is worse, in the manner of ADAS-Cog) declines more
slowly on the active arm than on placebo.

```{r}
#| label: fig-trajectory
#| fig-cap: "A longitudinal trajectory figure for a simulated two-arm
#|   trial. Thin lines are individual patients; heavy lines are arm
#|   means with a 95% confidence ribbon. This is the figure the
#|   in-house `zzlongplot` package produces by default."
#| fig-width: 7
#| fig-height: 4.5
set.seed(1)
n_id   <- 60
months <- c(0, 6, 12, 18, 24)

sim <- expand_grid(id = seq_len(n_id), month = months) |>
  mutate(arm = if_else(id <= n_id / 2, "Placebo", "Active")) |>
  group_by(id) |>
  mutate(
    slope = if_else(first(arm) == "Placebo", 0.45, 0.20) +
            rnorm(1, 0, 0.08),
    adas  = 20 + rnorm(1, 0, 4) + slope * month + rnorm(n(), 0, 1.5)
  ) |>
  ungroup()

arm_means <- sim |>
  group_by(arm, month) |>
  summarize(mean = mean(adas), se = sd(adas) / sqrt(n()),
            .groups = "drop")

pal <- c(Placebo = "#9d2235", Active = "#1f4e79")

ggplot(sim, aes(month, adas, color = arm)) +
  geom_line(aes(group = id), alpha = 0.15) +
  geom_ribbon(data = arm_means,
              aes(month, ymin = mean - 1.96 * se,
                  ymax = mean + 1.96 * se, fill = arm),
              alpha = 0.2, color = NA, inherit.aes = FALSE) +
  geom_line(data = arm_means, aes(month, mean, color = arm),
            linewidth = 1.1, inherit.aes = FALSE) +
  scale_color_manual(values = pal) +
  scale_fill_manual(values = pal) +
  labs(x = "Month", y = "ADAS-Cog (higher is worse)",
       color = "Arm", fill = "Arm",
       title = "Simulated cognitive decline by treatment arm")
```

The individual trajectories, drawn at low opacity,
convey the within-arm heterogeneity that a plot of
group means alone would hide. The confidence ribbon on
the arm means keeps the reader from over-reading a
separation that is within the noise. The real ADNI
version of this figure appears in @sec-adni.

## The same figure in Python

Many students arrive with more Python than R, and a
good deal of collaborative biostatistics now crosses
the two languages. Where a comparison is instructive we
show the Python equivalent alongside the R. Quarto
executes Python chunks through `reticulate`, and the
`r` object makes an R data frame available to Python
without writing an intermediate file. Here is the
body-mass-versus-flipper-length scatter in `pandas` and
`matplotlib`.

```{python}
#| label: fig-py-scatter
#| fig-cap: "The penguin scatter reproduced in Python. The cleaned R
#|   data frame crosses into Python through reticulate's `r` object,
#|   so no intermediate file is written."
#| fig-width: 6
#| fig-height: 4
import matplotlib
matplotlib.use("Agg")
import matplotlib.pyplot as plt

df = r.penguins_clean
colors = {"Adelie": "#1f4e79", "Chinstrap": "#9d2235",
           "Gentoo": "#2e8b57"}

fig, ax = plt.subplots(figsize=(6, 4))
for species, sub in df.groupby("species", observed=True):
    ax.scatter(sub["flipper_length_mm"], sub["body_mass_g"],
               label=species, color=colors[species], alpha=0.7, s=18)
ax.set_xlabel("Flipper length (mm)")
ax.set_ylabel("Body mass (g)")
ax.legend(title="Species")
fig.tight_layout()
plt.show()
```

For a reader who wants the grammar of graphics in
Python rather than the imperative `matplotlib` style,
the `plotnine` package reproduces the `ggplot2` API
almost verbatim (`from plotnine import ggplot, aes,
geom_point`). The choice between them is a matter of
taste and of what the rest of the analysis is written
in. The judgement the figure requires, one message per
plot, uncertainty shown, color chosen for legibility,
does not change with the language.

## Collaborating with an LLM on graphics

LLMs handle ggplot well; the trap is busy plots that
encode too much.

**Prompt 1: drafting a plot.** Describe the data and
the question, ask: 'write a ggplot that addresses
the question. Use a color-blind-safe palette and
clear axis labels with units.'

*What to watch for.* The default LLM plot tends to be
busy: too many aesthetics encoded, default ggplot
theme. Push for clarity. Multiple iterations of
'simpler' tend to improve.

*Verification.* Render the plot. Ask whether a reader
who has never seen the data could state the message
in one sentence.

**Prompt 2: combining plots.** Describe four plots,
ask: 'combine these into a 2x2 grid with shared
legend, panel labels A through D.'

*What to watch for.* `patchwork::plot_layout(guides = "collect")`
for the legend. Panel labels usually go in `labs(title = "A. ...")`
rather than `tag_levels`. The LLM may use either; both
work.

*Verification.* Render the combined plot. Are
legends shared? Are panels labeled in the right
order?

**Prompt 3: theme function.** Ask: 'write a
`theme_practicum()` function with serif body text,
sans-serif axis labels, and a color-blind-safe
default palette.'

*What to watch for.* The output is a starting point.
Test with a few plots; iterate.

*Verification.* Apply to several plots; ensure
consistency. Verify palette via Coblis or
`colorBlindness::cvdPlot()`.

## Principle in use

Three habits define defensible plotting:

1. **One plot, one message.** Resist combining
   relationships into one figure.
2. **Set the theme once.** `theme_set(theme_practicum())`
   at the top of every analysis script.
3. **Export for the destination.** PDF for LaTeX,
   PNG at 300 DPI for Word, SVG for web. Embed fonts
   in PDFs.

## Exercises

1. Using the `palmerpenguins` data, build a
   three-panel patchwork: (a) scatter of body mass
   vs flipper length colored by species; (b)
   residuals from a linear fit of (a); (c) QQ plot
   of residuals. Share the species legend across
   all three.
2. Write a function `plot_per_site(data, site_col, outcome)`
   that returns a named list of ggplot objects
   (one per site level) and a helper that saves
   each to `figures/<site>.pdf`.
3. Define `theme_practicum()` and apply it to three
   plots from any prior exercise. Verify the plots
   render consistently in HTML, PDF, and Word
   output.
4. Replicate one published figure from a recent
   biomedical paper using ggplot2. Compare your
   replication to the original; identify what is
   the same and what is different.
5. Verify your color palette with the Coblis
   simulator. Adjust if any pair of categories
   become indistinguishable under deuteranopia.

## Further reading

- [@wickham2016ggplot2], *ggplot2: Elegant Graphics
  for Data Analysis*, the canonical reference.
- [@wilke2019fundamentals], *Fundamentals of Data Visualization*
  at `clauswilke.com/dataviz`, effective-
  visualization principles.
- The `patchwork`, `cowplot`, and `gganimate`
  package vignettes.

## Prerequisites answers {#sec-quiz-answers-16}

1. Use `split(data, data$group) |> map(\(d) ggplot(d, aes(x, y)) + geom_point())`
   to produce a named list of plots, one per group.
   Alternatively, in a tidyverse pipeline:
   `data |> group_by(group) |> group_map(\(d, k) ggplot(d, aes(x, y)) + geom_point())`
   (note: `group_map` returns a list, with names
   taken from the group keys if you pass
   `.keep = TRUE` and post-process).
2. `facet_wrap()` splits a single plot into panels
   by a factor. Every panel shares the same `aes`
   mapping, same geom, same scales.
   `patchwork::wrap_plots()` composes *distinct*
   plots that can differ in every respect (different
   geoms, different data, different scales). Use
   `facet_wrap()` for small-multiples of the same
   plot; `wrap_plots()` for multi-panel figures
   showing different views.
3. PDF is a vector format: it scales cleanly, looks
   crisp at any zoom level, embeds fonts, and is
   what journal production systems prefer for
   typeset PDFs. PNG is raster: appropriate for
   embedding in Word documents, PowerPoint slides,
   or web pages where vector support is limited.
   Save PDF for LaTeX submissions, PNG at 300 DPI
   for Word and web.
