# Clinical Data Standards: CDISC, SDTM, and ADaM {#sec-cdisc}

::: {.callout-note}
## Sources

Authored directly for this book at
`~/Dropbox/prj/tch/01-phb228-stat-computing/phb228-2026/textbook/19-cdisc.qmd`.
Target audience: students heading to pharma or CRO statistical
programming roles who have no prior exposure to CDISC. Written
in response to reports from alumni that a gap in this area was
felt on the first day of their industry positions.
:::

## Prerequisites

Answer the following questions to see if you can bypass this
chapter. You can find the answers at the end of the chapter in
@sec-quiz-answers-19.

1. What is the difference between SDTM and ADaM, and which one
   does a biostatistician typically build?
2. In the CDISC convention, does `CNSR = 1` indicate a censored
   observation or an event?
3. What is the role of `ADSL` relative to other ADaM datasets
   such as `ADLB` or `ADTTE`?

## Learning objectives

By the end of this chapter you should be able to:

- Name the four main CDISC standards (SDTM, ADaM, CDASH, SEND)
  and identify which are required for FDA submissions.
- Describe the data flow from CRF through SDTM to ADaM to
  tables, listings, and figures (TLFs).
- Read an SDTM domain and an ADaM dataset and explain how rows
  and variables relate.
- Derive a minimal `ADSL` and `ADTTE` from synthetic SDTM
  domains using R.
- Recognise the common ADaM variable names (`USUBJID`,
  `PARAMCD`, `AVAL`, `CNSR`, and the population flags).
- Explain the purpose of Define-XML, the annotated CRF (aCRF),
  and the reviewer guides (SDRG, ADRG).

## Orientation

Alumni of this programme working in pharmaceutical companies
and contract research organisations (CROs) have reported that
their first weeks on the job involved encountering data
structures, variable names, and regulatory conventions that the
statistics curriculum had not prepared them for. This chapter
addresses that gap.

The Clinical Data Interchange Standards Consortium (CDISC) is
a non-profit that publishes data standards for clinical trials.
Since December 2016, the United States Food and Drug
Administration (FDA) has required that sponsors submit clinical
trial data in CDISC formats to support new drug applications
(NDAs) and biologics license applications (BLAs). The two
formats a statistician will touch are SDTM (Study Data
Tabulation Model) and ADaM (Analysis Data Model).

The organising insight for the chapter is this: SDTM is how
the data arrives at the statistician's desk; ADaM is what the
statistician builds, documents, and analyses. Your deliverables
as an industry biostatistician will, with high probability, be
either ADaM datasets and their specifications, or tables
derived from them.

## The regulatory pipeline

The flow of clinical trial data from collection to submission
follows a standardised pipeline:

1. **Protocol** defines the trial design, endpoints, and
   statistical analysis plan (SAP).
2. **CRF** (Case Report Form) is the instrument sites use to
   record observations on each participant.
3. **SDTM** domains are built by data management from the
   collected CRF data and submitted to the FDA as the raw
   tabulations layer.
4. **ADaM** datasets are derived from SDTM by biostatistics
   programmers. They are analysis-ready and also submitted to
   the FDA.
5. **TLFs** (Tables, Listings, and Figures) are produced from
   ADaM for the Clinical Study Report (CSR) that accompanies
   the submission.

Two rules organise this pipeline:

- **Immutability of source.** Once SDTM is locked, it is not
  edited to accommodate new analyses. A new analytic need
  produces a new ADaM variable, not a change to SDTM.
- **Traceability.** Every ADaM variable must point back to its
  SDTM source and its derivation logic, documented in the
  ADaM specification and the Define-XML.

## SDTM: the FDA's view of the raw data

SDTM organises a trial's observations into **domains**, each a
dataset covering one topic area. Core domains you will see
include:

- `DM`: demographics (one row per subject).
- `EX`: exposure or dosing (one row per dose).
- `AE`: adverse events (one row per event).
- `LB`: laboratory results (one row per test result).
- `VS`: vital signs.
- `DS`: disposition and end-of-study status.

Two features of SDTM are worth noting. First, its structure is
**vertical**: within a domain, each row is a single
observation, and the same variable names (`USUBJID`, `VISIT`,
`--TESTCD`, `--ORRES`, `--STRESN`) recur across domains.
Second, dates are stored as ISO 8601 character strings such
as `'2024-03-15'` or `'2024-03-15T08:30'`, not as R `Date`
objects; this is an FDA submission requirement.

You do not usually build SDTM as a biostatistician. You read
it.

## ADaM: the analysis-ready layer

ADaM datasets take SDTM as input and produce rows and columns
shaped for direct use by statistical procedures. There are
three principal structures:

- **ADSL** (Subject-Level Analysis Dataset): exactly one row
  per subject. Contains treatment assignment, demographics,
  baseline covariates, key dates, and population flags. Every
  other ADaM dataset joins back to `ADSL` by `USUBJID`.
- **BDS** (Basic Data Structure): one row per subject per
  parameter per analysis visit. Used for longitudinal,
  laboratory, vital-signs, and time-to-event analyses. Common
  examples include `ADLB`, `ADVS`, and `ADTTE`.
- **OCCDS** (Occurrence Data Structure): one row per
  occurrence. Used for adverse events (`ADAE`) and
  concomitant medications (`ADCM`).

A few variable-name conventions recur across every ADaM
dataset you will encounter:

| Variable | Meaning |
|---|---|
| `USUBJID` | Unique subject identifier, sponsor-wide. |
| `TRT01P`, `TRT01A` | Planned and actual treatment, period 1. |
| `PARAMCD`, `PARAM` | Parameter code and decoded name. |
| `AVAL`, `AVALC` | Analysis value (numeric, character). |
| `AVISIT`, `AVISITN` | Analysis visit (character, numeric). |
| `CNSR` | Censoring indicator (1 = censored). |
| `ITTFL`, `SAFFL`, `EFFFL` | Population flags. |
| `ANL01FL` | Analysis record flag. |
| `DTYPE` | Derivation type (e.g. `'LOCF'`). |

The `CNSR` convention is the single most common source of
day-one errors for new statisticians. CDISC uses `CNSR = 1`
for censored and `CNSR = 0` for event. The `survival::Surv()`
function in R uses the opposite convention: `event = 1`. When
moving between `ADTTE` and `Surv()`, invert.

**Controlled terminology.** CDISC publishes codelists that
enumerate the valid values of categorical variables. `SEX`,
for instance, takes values in `{'M', 'F', 'U',
'UNDIFFERENTIATED'}`, not `'Male'`, `'male'`, or `'1'`. Your
ADaM specification will cite the relevant CDISC codelist
version for each controlled variable.

## Worked example: SDTM to ADaM to survival in R

The example below fabricates minimal SDTM domains, derives
`ADSL` and `ADTTE` for overall survival, and fits a
Kaplan-Meier estimator and a Cox model. The row granularity
changes at each step; tracking it is the concept.

```{r}
#| label: cdisc-setup
#| message: false

library(dplyr)
library(survival)

set.seed(47)
n <- 100
```

### Fabricate SDTM domains

Real SDTM has many more required variables per domain. We keep
only what ADaM needs so that the pipeline is visible.

```{r}
#| label: cdisc-sdtm

dm <- tibble(
  USUBJID = sprintf('STUDY01-%03d', 1:n),
  AGE     = round(rnorm(n, 62, 10)),
  SEX     = sample(c('M', 'F'), n, replace = TRUE),
  RACE    = sample(
    c('WHITE', 'BLACK OR AFRICAN AMERICAN', 'ASIAN'),
    n, replace = TRUE, prob = c(0.70, 0.20, 0.10)),
  ARM     = sample(c('Placebo', 'Drug 50mg'), n,
                   replace = TRUE),
  RFSTDTC = as.Date('2024-01-01') +
            sample(0:180, n, replace = TRUE)
)

ex <- dm |>
  mutate(EXSTDTC = RFSTDTC + sample(0:3, n(), replace = TRUE),
         EXTRT   = ARM) |>
  select(USUBJID, EXSTDTC, EXTRT)

ds <- dm |>
  mutate(
    hazard   = if_else(ARM == 'Drug 50mg', 0.0015, 0.0030),
    tt_event = rexp(n(), rate = hazard),
    t_admin  = 365 + sample(0:180, n(), replace = TRUE),
    obs_time = pmin(tt_event, t_admin),
    DSDECOD  = if_else(tt_event <= t_admin,
                       'DEATH', 'COMPLETED'),
    DSSTDTC  = RFSTDTC + round(obs_time)
  ) |>
  select(USUBJID, DSDECOD, DSSTDTC)
```

### Derive ADSL

`ADSL` has exactly one row per subject. Every population flag,
every treatment variable, and every demographic that a
downstream analysis will stratify on is placed here.

```{r}
#| label: cdisc-adsl

adsl <- dm |>
  left_join(ex |> select(USUBJID, EXSTDTC),
            by = 'USUBJID') |>
  left_join(ds, by = 'USUBJID') |>
  mutate(
    TRT01P = ARM,
    TRT01A = ARM,
    TRTSDT = EXSTDTC,
    TRTEDT = DSSTDTC,
    AGEGR1 = cut(AGE, c(-Inf, 65, Inf),
                 labels = c('<65', '>=65')),
    SAFFL  = if_else(!is.na(TRTSDT), 'Y', 'N'),
    ITTFL  = 'Y',
    EFFFL  = SAFFL
  ) |>
  select(USUBJID, AGE, AGEGR1, SEX, RACE,
         TRT01P, TRT01A, TRTSDT, TRTEDT,
         SAFFL, ITTFL, EFFFL)

stopifnot(nrow(adsl) == n_distinct(adsl$USUBJID))
```

### Derive ADTTE

`ADTTE` follows the BDS structure: one row per subject per
time-to-event parameter. Here we build a single parameter,
overall survival (OS). A real `ADTTE` might stack additional
rows for progression-free survival (PFS), disease-free survival
(DFS), or time to treatment failure (TTF).

```{r}
#| label: cdisc-adtte

adtte <- adsl |>
  left_join(ds, by = 'USUBJID') |>
  mutate(
    PARAMCD  = 'OS',
    PARAM    = 'Overall Survival (days)',
    STARTDT  = TRTSDT,
    ADT      = DSSTDTC,
    AVAL     = as.numeric(ADT - STARTDT),
    CNSR     = if_else(DSDECOD == 'DEATH', 0L, 1L),
    EVNTDESC = if_else(CNSR == 0L, 'Death',
                       'Administrative censoring'),
    ANL01FL  = 'Y'
  ) |>
  select(USUBJID, TRT01P, TRT01A, SAFFL, EFFFL,
         PARAMCD, PARAM, STARTDT, ADT, AVAL, CNSR,
         EVNTDESC, ANL01FL)

stopifnot(all(adtte$AVAL >= 0),
          all(adtte$CNSR %in% c(0L, 1L)))

head(adtte)
```

### Analyse from ADTTE

Note the `1L - CNSR` inversion from the CDISC convention to
`survival::Surv()`.

```{r}
#| label: cdisc-survival

analysis <- adtte |>
  filter(EFFFL == 'Y', PARAMCD == 'OS')

fit <- survfit(Surv(AVAL, 1L - CNSR) ~ TRT01P,
               data = analysis)
print(fit)

cox <- coxph(Surv(AVAL, 1L - CNSR) ~ TRT01P,
             data = analysis)
summary(cox)
```

The progression of row granularity across the pipeline is the
concept to retain: `dm` is one row per subject, `ex` is one row
per dose, `ds` is one row per disposition event, `adsl` returns
to one row per subject, and `adtte` is one row per subject per
parameter. Tracking this at every join is how you avoid
silently duplicating subjects; the `stopifnot()` calls above
are the doctrinal minimum.

## The submission package

A complete submission to the FDA is not only SDTM and ADaM
datasets. It includes several supporting artefacts:

- **Define-XML**: machine-readable metadata describing every
  variable in every dataset, with controlled terminology
  references, derivations, and provenance. Reviewers ingest
  it to navigate the submission.
- **Annotated CRF (aCRF)**: the blank CRF with SDTM variable
  names annotated on each field. It is the bridge from what
  was collected to what was submitted.
- **Study Data Reviewer's Guide (SDRG)** and **Analysis Data
  Reviewer's Guide (ADRG)**: narrative PDFs that explain
  conformance decisions, known issues, and analysis choices.
- **Statistical Analysis Plan (SAP)**: pre-specifies every
  analysis that appears in the CSR, including populations,
  estimands, and sensitivity analyses.

You will not usually author Define-XML by hand. Tools such as
`pinnacle21` and the open-source pharmaverse packages generate
it from ADaM specifications.

## The pharmaverse

Until the late 2010s, most ADaM work in industry was done in
SAS. Since approximately 2020, a coordinated open-source
initiative called the **pharmaverse** has produced R packages
that implement ADaM derivations with validation intended for
regulatory use. The principal packages include:

- `admiral`: modular ADaM derivations. The reference
  implementation for `ADSL`, `ADAE`, `ADLB`, `ADTTE`, and
  others.
- `metacore` and `metatools`: read and validate ADaM
  specifications.
- `xportr`: export to SAS transport (XPT) format, which the
  FDA still requires.
- `pharmaverseadam`: reference ADaM datasets for teaching and
  testing.

At a sponsor or CRO that has adopted R, expect to encounter
these packages in production pipelines. They are a more useful
thing to learn than any proprietary alternative.

## With AI assistance {.unnumbered}

::: {.callout-tip}
## Prompts to try

1. Paste an SDTM domain (even a synthetic one) and ask the LLM
   to describe its shape, its role, and its relationship to
   other domains. Check whether the model correctly identifies
   the granularity (one row per what).
2. Give the LLM an `ADSL` specification and ask it to generate
   an `admiral`-style derivation. Compare its output to the
   `admiral` reference implementation. Does it use the correct
   helper functions, or does it reinvent them?
3. Ask the LLM to write a reviewer-facing explanation of the
   `CNSR` convention. Evaluate whether the explanation would
   be accepted at face value by a new statistician, or whether
   it contains subtle errors about which value means what.
:::

## Exercises

1. Extend the `ADTTE` in the worked example to include a
   second parameter, progression-free survival (PFS), stacked
   as additional rows. Add a synthetic progression event to
   the `DS` domain to support it. Verify that
   `nrow(adtte) == 2 * n`.
2. Build a minimal `ADAE` (OCCDS structure) from a fabricated
   `AE` SDTM domain. Include `AEDECOD`, `AESER`, `AESEV`, and
   a treatment-emergent flag `TRTEMFL` derived from whether
   `AESTDTC >= TRTSDT`.
3. Install `admiral` and `pharmaverseadam` from CRAN. Load
   `pharmaverseadam::adsl` and compare its column list to the
   `ADSL` built in this chapter. Which variables are missing
   from our simplified version, and which of them would matter
   for a real oncology analysis?
4. Write an ADaM specification table (in Quarto) for the
   `ADTTE` built above: variable name, label, type, source,
   and derivation. Render it to PDF.

## Further reading

- CDISC Foundational Standards at `cdisc.org/standards/foundational` --- authoritative reference.
- `admiral` documentation at `pharmaverse.github.io/admiral` --- the CRAN implementation.
- FDA Study Data Technical Conformance Guide --- the regulatory requirements.

## Prerequisites answers {#sec-quiz-answers-19}

1. SDTM (Study Data Tabulation Model) holds the regulatory
   view of the raw, collected clinical data, organised into
   domains such as `DM`, `AE`, `LB`, and `EX`. ADaM (Analysis
   Data Model) holds analysis-ready datasets derived from
   SDTM. A biostatistician typically builds ADaM; SDTM is
   built by data management or clinical programmers.
2. `CNSR = 1` indicates a censored observation. `CNSR = 0`
   indicates the event of interest. This convention is the
   inverse of `survival::Surv()`, which uses `event = 1`.
3. `ADSL` is the subject-level anchor dataset. It contains
   exactly one row per subject, with treatment assignment,
   demographics, baseline covariates, key dates, and
   population flags. Every other ADaM dataset (`ADLB`,
   `ADTTE`, `ADAE`, and so on) joins back to `ADSL` by
   `USUBJID` and inherits treatment and population flags
   from it.
