# Case Study: ADNI MCI Prediction {#sec-adni}

::: {.callout-note}
## Sources

PHB 243B Lectures 12–17 (`~/Dropbox/prj/tch/02-phb243b-biostat/lec12`
through `lec17/`), including `analysis_plan_template.qmd`,
`grassi.Rmd`, `ADNIMERGEtadpole.csv`, and the Grassi et al.
(2019) paper. Uses the in-house packages `zzcollab` (@sec-zzcollab),
`zztable1`, and `zzlongplot`.
:::

## Prerequisites

Answer the following questions to see if you can bypass this
chapter. You can find the answers at the end of the chapter in
@sec-quiz-answers-24.

1. What is the scientific question posed by the ADNI MCI
   prediction project, and why is it clinically meaningful?
2. What features of the ADNI data make a reproducible
   compendium especially important relative to a typical
   pilot dataset?
3. Why must ADNIMERGE data be preprocessed before modelling,
   and what preprocessing step is most commonly overlooked?

## Learning objectives

By the end of this chapter you should be able to:

- Frame a concrete clinical prediction question from a
  longitudinal cohort dataset.
- Scaffold an ADNI compendium using `zzcollab`.
- Produce a Table 1 (demographics, baseline clinical
  variables) using the in-house `zztable1` package.
- Produce trajectory plots for cognitive outcomes using
  `zzlongplot` with CDISC-style formatting.
- Pre-specify an analysis plan that distinguishes primary
  from exploratory analyses.
- Deliver a reproducible report that satisfies federal
  data-sharing requirements (@sec-federal).

## Orientation

This chapter works through a single end-to-end case study:
predicting conversion from Mild Cognitive Impairment (MCI)
to Alzheimer's disease (AD) dementia within a three-year
horizon, using baseline sociodemographic, clinical, and
neuropsychological measures. The goal is not to solve the
prediction problem (that is a book-length exercise) but to
demonstrate the Practicum workflow end-to-end: scaffolding,
planning, wrangling, figuring, reporting, and depositing.

The statistical modelling details are deliberately light;
for those, consult the companion book *Statistical Computing
in the Age of AI*. Here, the focus is on everything *around*
the modelling.

## The research question

From the 243B analysis-plan template:

> Can baseline sociodemographic characteristics, clinical
> scales, and neuropsychological test scores predict
> conversion from MCI to AD dementia within 3 years?

This is a well-posed prediction question. Secondary
questions from the same template:

- Which predictor categories (sociodemographic, clinical,
  neuropsychological) contribute most to prediction
  accuracy?
- Does prediction performance differ between Early MCI
  (EMCI) and Late MCI (LMCI) subgroups?

TODO: brief paragraph on clinical context and on prior work
(Grassi et al. 2019).

## The ADNI dataset

**Source.** ADNI (Alzheimer's Disease Neuroimaging
Initiative) is a longitudinal multicentre cohort study
funded since 2004. Data are obtained under a data-use
agreement at <https://adni.loni.usc.edu/>.

**The `ADNIMERGE` package** is the community-standard R
interface, a single long-format tibble of approximately
15,000 observations on ~2000 participants, with 115
variables spanning demographics, APOE genotype, clinical
scales (CDR, FAQ), and neuropsychological tests (ADAS, RAVLT,
logical memory, MMSE).

**Key variables for this case study.**

| Variable      | Description                                   |
| ------------- | --------------------------------------------- |
| `RID`         | Participant ID                                |
| `VISCODE`     | Visit code (`bl`, `m06`, `m12`, …)            |
| `DX`          | Diagnosis at visit (CN/MCI/Dementia)          |
| `DX_bl`       | Baseline diagnosis (CN/SMC/EMCI/LMCI/AD)      |
| `AGE`, `PTGENDER`, `PTEDUCAT` | Demographics                  |
| `APOE4`       | Number of APOE4 alleles (0/1/2)               |
| `ADAS13`      | 13-item Alzheimer's Disease Assessment Scale  |
| `RAVLT.immediate` | Rey Auditory Verbal Learning Test        |
| `MMSE`        | Mini-Mental State Examination                 |
| `LDELTOTAL`   | Logical Memory II delayed recall              |

TODO: one or two paragraphs on the psychometric
instruments and their clinical interpretation.

## Why a compendium matters here

ADNI analyses have three properties that make reproducibility
especially critical:

1. **Data-use agreement.** Data cannot be shared, but the
   analysis code and the `renv.lock` can. Collaborators
   reproduce the work by independently obtaining the data
   and running the same code.
2. **Long time horizon.** Papers using ADNI are often
   revisited years later (reviewer comments, follow-up
   analyses). A Docker-pinned environment means the original
   analysis can be rerun at year 3 on the exact same R and
   package versions.
3. **Version churn in ADNIMERGE.** The dataset is updated
   periodically; new participants are added and variables
   are renamed. Analyses must pin to a specific ADNIMERGE
   version (documented in `renv.lock` if installed from
   GitHub, or captured as a processed data file in the
   compendium).

## Scaffolding the compendium with zzcollab

```bash
mkdir adni-mci
cd adni-mci
zzc modeling             # modeling profile: glmnet, survival
make r                   # enter the container
```

Inside the container, add the ADNI data access layer:

```{r eval=FALSE}
# One-off: install ADNIMERGE from Github (pinned in renv.lock)
renv::install('ADNI/ADNIMERGE')
renv::snapshot()
```

Under `analysis/`, the compendium grows:

```
analysis/
├── data/
│   ├── raw/adnimerge_20260301.rds      # frozen copy
│   └── derived/analytic_cohort.rds     # processed
├── paper/paper.qmd                     # the manuscript
└── scripts/
    ├── 01-build-cohort.R
    ├── 02-table1.R
    ├── 03-trajectories.R
    └── 04-model.R
```

## Building the analytic cohort

TODO: code walkthrough for `01-build-cohort.R`: filter
`DX_bl %in% c('EMCI', 'LMCI')`, create a three-year
follow-up window, derive the binary outcome 'converted to AD
within 3y'.

## Table 1: demographics with `zztable1`

```{r eval=FALSE}
library(zztable1)

table1(
  data     = analytic_cohort,
  vars     = c('AGE', 'PTGENDER', 'PTEDUCAT', 'APOE4',
               'ADAS13', 'MMSE'),
  strata   = 'converted_3y',
  overall  = TRUE,
  output   = 'latex'
) |>
  zztab2fig::to_pdf('analysis/paper/tables/table1.pdf')
```

TODO: discuss `zztable1`'s defaults (continuous → mean [SD],
categorical → n (%)), how to override, and how to produce
CDISC-compliant output with the `theme = 'regulatory'`
argument.

## Cognitive trajectories with `zzlongplot`

```{r eval=FALSE}
library(zzlongplot)

longplot(
  data       = adnimerge_clean,
  time       = 'years_since_baseline',
  outcome    = 'ADAS13',
  id         = 'RID',
  group      = 'DX_bl',
  facet      = 'APOE4_any',
  stat       = 'mean_ci',
  theme      = 'regulatory'
) +
  labs(
    title    = 'ADAS13 trajectories by baseline diagnosis',
    subtitle = 'Stratified by any APOE4 allele (ADNI 1/GO/2/3)'
  )
```

TODO: explain how `zzlongplot`'s `stat = 'mean_ci'` computes
group-wise means with Wald CIs at each visit, and how the
`theme = 'regulatory'` preset produces CDISC-style plots
that satisfy journal figure standards without custom
`ggplot2::theme()` code.

## Pre-specified analysis plan

Before touching the analytic cohort, the team writes an
SAP (@sec-analysis-plan). For ADNI, the 243B template
includes placeholders for:

1. Background and rationale
2. Research questions and hypotheses
3. Cohort definition (inclusion/exclusion)
4. Exposure, outcome, and covariate definitions
5. Primary analysis: pre-specified predictor set, model,
   metric
6. Secondary analyses: EMCI vs LMCI, predictor-category
   contributions
7. Missing-data handling
8. Sensitivity analyses

TODO: call out what in this list is pre-specified (locked
before data access) vs exploratory (documented but not
pre-registered).

## Modelling (summary only)

The modelling step itself belongs to the companion book;
for Practicum purposes, the key operational points are:

- Use `caret` or `tidymodels` for the cross-validated
  predictive pipeline, with the outer CV fold defined at
  the `RID` level (not the observation level).
- Report AUC with bootstrap CIs.
- Decision threshold should be reported alongside
  sensitivity, specificity, PPV, and NPV.
- Store predictions in the compendium as
  `analysis/data/derived/predictions.rds` so figures and
  tables can be regenerated without refitting.

## Reporting

The manuscript lives at `analysis/paper/paper.qmd`. Its
frontmatter imports a journal template from `rticles`
and the bibliography from `references.bib`. Rendering is
driven by `make render`, which invokes Quarto inside the
Docker container so that the environment matches the
`renv.lock`.

TODO: describe the handoff: what gets deposited with the
paper (Zenodo DOI of the compendium without data, plus
Docker image), what requires ADNI data-use agreement
approval (the data itself).

## Verifying reproducibility

Before submitting:

1. `make check-renv` to audit the lockfile.
2. `docker image prune` then `make docker-build` to verify
   the image builds from scratch.
3. `make render` to verify the paper builds from scratch.
4. On a colleague's machine: `git clone`, `make r`, `make
   render`. Expected: bit-identical output.

## With AI assistance {.unnumbered}

::: {.callout-tip}
## Prompts to try

1. Paste the 243B ADNI analysis-plan template and ask the
   LLM to draft the 'Inclusion/Exclusion' section for EMCI
   and LMCI participants with a 3-year follow-up
   requirement. Check whether it correctly handles
   participants who die or drop out before year 3.
2. Ask the LLM to critique the proposed `zztable1` call
   above. Does it catch that `APOE4` is currently numeric
   (0/1/2) and should probably be a factor for Table 1?
3. Ask the LLM to write the SAP's 'Missing Data Handling'
   section for a prediction model with 30% missingness in
   one predictor. Compare its proposal (mean imputation?
   multiple imputation? complete-case analysis with a
   sensitivity check?) to what you would recommend.
:::

## Exercises

1. Reproduce this chapter's compendium scaffold for a
   different ADNI outcome (e.g., progression from CN to
   MCI). Adapt the SAP template accordingly.
2. Add a second table to the compendium using `zztable1`:
   comparison of baseline cognitive scores between
   converters and non-converters. Produce both LaTeX and
   HTML output.
3. Use `zzlongplot` to produce a MMSE trajectory plot
   stratified by `DX_bl` with a `theme = 'regulatory'`
   override of the default colour palette. Save at 300 DPI
   for journal submission.

## Quiz answers {#sec-quiz-answers-24}

1. The question is: *can baseline sociodemographic,
   clinical, and neuropsychological measures predict
   conversion from MCI to AD within three years?* It is
   clinically meaningful because MCI is a heterogeneous
   state: only some participants progress to AD on a
   meaningful timescale, and identifying those at highest
   risk enables targeted intervention trials and earlier
   patient counselling.
2. ADNI analyses carry three distinguishing features: a
   data-use agreement that prevents data sharing (so the
   compendium must be code-only); long time horizons
   (years between analysis and reviewer response, during
   which the R ecosystem drifts); and a dataset that
   updates periodically with renamed variables. All three
   increase the value of Docker-pinned environments and
   `renv`-pinned R package versions.
3. ADNIMERGE is long-format (one row per participant per
   visit), with mixed diagnostic labels and inconsistent
   visit codes across ADNI 1, GO, 2, and 3. Analytic
   cohorts require derivation of baseline-anchored
   follow-up windows, careful handling of participants
   who drop out before the outcome window closes, and
   consistent visit alignment. The step most commonly
   overlooked is *censoring dropouts correctly*: a
   participant who leaves the study after year 2 has not
   been observed for the three-year outcome and must be
   excluded or handled with time-to-event methods, not
   treated as a non-converter.
