# Factors, Strings, and Dates {#sec-types}

::: {.bookepigraph}
```{=latex}
\begin{flushright}\small
```
*There are only two hard things in Computer Science: cache
invalidation and naming things.*

[Phil Karlton]{.smallcaps}, *attributed* (c. 1996)
```{=latex}
\end{flushright}
```
:::

::: {.callout-note}
## Sources

Stat 545 Chapters 10--13 (Jenny Bryan, UBC); the
`forcats`, `stringr`, and `lubridate` packages.
:::

## Prerequisites

Answer the following questions to see if you can bypass
this chapter. You can find the answers at the end of the
chapter in @sec-quiz-answers-14.

1. Why has `stringsAsFactors = FALSE` been the default
   since R 4.0, and what problems did the old default
   cause?
2. What does `stringr::str_detect()` return, and how
   does it differ from `stringr::str_extract()`?
3. Given `c('2026-04-23', '04/23/2026', 'April 23, 2026')`,
   write one `lubridate`/`readr` pipeline that parses
   all three into `Date` objects.

## Learning objectives

By the end of this chapter you should be able to:

- Manage factors idiomatically with `forcats`
  (reorder, lump, relevel, drop, infreq).
- Use `stringr` verbs confidently for detection,
  extraction, replacement, splitting, and
  interpolation.
- Parse dates robustly with `lubridate`, including
  mixed formats and time zones.
- Diagnose and fix the common 'character vs. factor
  vs. ordered factor' decision that causes modeling
  errors downstream.
- Recognize locale, time-zone, and Excel-date pitfalls.

## Orientation

Numeric columns misbehave loudly. A value that should be a
number and is not will stop the analysis, and the fix follows
within the minute. Text, categories, and dates are the opposite
case, and that is why they merit a chapter of their own. A date
parsed under the wrong convention is still a date. A factor
with an unintended reference level still fits. A string with
trailing whitespace still joins, to nothing. Each of these
produces a result rather than an error. Unfortunately, the
result is wrong in a way that no amount of staring at the
output will reveal, because the output looks exactly as it
should.

We take up here the three packages that make these types
tractable. Master them, and a large share of biomedical data
cleaning becomes routine. More to the point, the failures that
remain become the loud kind.

The packages are:

- **`forcats`** for factors (categorical variables with
  defined levels).
- **`stringr`** for character operations.
- **`lubridate`** for dates and times.

All three are part of the tidyverse, with consistent
naming and pipe-friendly interfaces.

## The statistician's contribution {#sec-types-human}

The three packages are well documented and pleasant to
use. The bugs they exist to prevent are not bugs of syntax.
They are bugs of timing and assumption:

**Factor or character?** A character vector is
flexible: any value, any order, any new level later. A
factor is constrained: a fixed set of levels, a defined
order, attached metadata. For analysis, factors are
correct: the model needs to know the reference level
and the level set. For data cleaning and joining,
characters are usually safer (no surprises when a
new value arrives). Convert to factor late, near the
modeling step.

**Reference levels matter.** A logistic regression with
factor `treatment` and reference 'high-dose' produces
coefficients with a different interpretation from one
with reference 'placebo'. The default (alphabetical) is
rarely what you want. Set the reference deliberately,
in writing, in the data-cleaning script.

**Time zones are tricky.** A 'date' value '2026-04-23'
without a time zone is fine for daily-resolution
analyses. A 'time' value '2026-04-23 02:00:00' is
ambiguous: which time zone? UTC? US/Pacific? The local
time zone of the machine that recorded it? Time-zone
errors propagate quietly through analyses; daylight-
saving transitions cause spurious 1-hour gaps.

**Excel dates.** When data come from Excel, dates may
arrive as 5-digit serial numbers ('45405') rather than
date strings. Parsing them as dates produces nonsense
(year 124, depending on the epoch convention). The fix
is `janitor::excel_numeric_to_date()` or careful
import. Verify any column whose name suggests dates
but whose values are numeric.

These judgments are what distinguish a working data
cleaning script from one that silently mishandles a
common type.

The factor-or-character decision is not a matter of taste.
Nor is it a decision made once. It is a decision about *when*
in the pipeline the conversion happens, and @fig-factor-timing
gives the answer this chapter argues for.

```{mermaid}
%%| echo: false
%%| label: fig-factor-timing
%%| fig-alt: "Left-to-right chain. Read, clean and recode, and join
%%|   all stay in character type. A highlighted node marks where to
%%|   convert to factor, setting the reference level deliberately.
%%|   From there the chain branches to Model and to Plot, both
%%|   working in factor type."
%%| fig-cap: "Where to convert. Character is the safe type through
%%|   reading, cleaning, and joining, because a new or misspelled
%%|   value is merely a new string rather than an invalid level.
%%|   Factor is the correct type at modeling and plotting, because
%%|   the model needs a fixed level set and a declared reference."
flowchart LR
  R["Read<br/><i>character</i>"] --> C["Clean and recode<br/><i>character</i>"]
  C --> J["Join<br/><i>character</i>"]
  J --> X["<b>Convert here</b><br/><i>factor, reference level<br/>set deliberately</i>"]
  X --> M["Model<br/><i>factor</i>"]
  X --> P["Plot<br/><i>factor, ordered<br/>by fct_reorder</i>"]
```

## Factors with `forcats`

The standard operations:

```{r}
#| label: types-forcats
#| message: false
library(forcats)
library(dplyr)
library(palmerpenguins)

f <- factor(c("low", "high", "low", "medium", "high"))

# levels in current order
levels(f)
#> [1] "high"   "low"    "medium"

# alphabetical by default; reorder by frequency
fct_infreq(f)

# reorder by another variable
penguins |>
  mutate(species = fct_reorder(species, body_mass_g, mean,
                               .na_rm = TRUE))

# specify order explicitly
fct_relevel(f, "low", "medium", "high")

# combine rare levels
fct_lump_n(f, n = 2)         # keep top 2; rest -> 'Other'
fct_lump_min(f, min = 5)     # keep levels with >= 5 obs

# drop unused levels
fct_drop(f)

# rename levels
fct_recode(f, "L" = "low", "M" = "medium", "H" = "high")
```

When to convert character to factor:

- Just before modeling: ensures the model knows the
  level set and the reference.
- Just before plotting: enables `fct_reorder` for
  ordered axis labels.

When to keep as character:

- During data cleaning: avoids 'invalid factor level'
  surprises when a new value arrives.
- For joining: factor matching can fail when levels
  differ between tables.

## Strings with `stringr`

Pattern detection, extraction, replacement, and
splitting:

```{r}
#| label: types-stringr
library(stringr)

s <- c("ABC123", "ABC", "XYZ-789", NA)

# detect: returns logical
str_detect(s, "[0-9]")
#> [1]  TRUE FALSE  TRUE  NA

# extract: returns first match per input, or NA
str_extract(s, "[0-9]+")
#> [1] "123" NA    "789" NA

# extract all matches per input (returns list)
str_extract_all(s, "[0-9]+")

# replace
str_replace(s, "[0-9]+", "***")

# split
str_split("a,b,c,d", ",", simplify = TRUE)
#>      [,1] [,2] [,3] [,4]
#> [1,] "a"  "b"  "c"  "d"

# interpolate
name <- "World"
str_glue("Hello, {name}!")
#> Hello, World!

# case
str_to_lower("ABC")           # "abc"
str_to_upper("abc")           # "ABC"
str_to_title("hello world")   # "Hello World"

# trim whitespace
str_trim("  abc  ")           # "abc"
str_squish("  a   b   c  ")   # "a b c"
```

The regex fundamentals you need:

- `[abc]`: any of a, b, c.
- `[a-z]`, `[A-Z]`, `[0-9]`: ranges.
- `\\d`, `\\w`, `\\s`: digit, word char, whitespace.
- `*`, `+`, `?`: zero-or-more, one-or-more, zero-or-one.
- `{n}`, `{n,m}`: exactly n, between n and m.
- `^`, `$`: start, end of string.
- `()`: capture group; `(?:...)` non-capturing.
- `|`: alternation.
- `\\`: escape; in R strings this is `\\\\` for a
  literal backslash.

For complex regex, build incrementally and test each
step.

::: {.callout-note icon=false collapse=true}
## Check your understanding: why character, then factor?

**Question.** This chapter argues for keeping categorical
variables as character through reading, cleaning and
joining, and converting to factor only at modeling and
plotting. What actually goes wrong if you convert at read
time instead?

**Answer.**

Two things, and the second is why the rule exists.

Recoding becomes harder. A factor's permitted values are
fixed at creation, so assigning a value that is not
already a level yields `NA` with a warning rather than the
value you assigned. Every cleaning step then has to
manage levels alongside data, and `str_replace()` on a
factor silently coerces to character anyway.

Joining becomes unsafe. Two factors built independently
from different sources almost never share a level set, and
their underlying integer codes mean different things.
Joining on them compares codes, not labels, so rows match
that should not and fail to match that should. Character
comparison is on the text and cannot go wrong this way.

The general principle: factor is a *modeling* type. It
encodes a fixed set of categories and a reference level,
which is exactly what a model needs and exactly what makes
it brittle while the data are still moving.
:::

## Dates with `lubridate`

Parsing:

```{r}
#| label: types-lubridate-parse
library(lubridate)

# fixed format
ymd("2026-04-23")
mdy("4/23/2026")
dmy("23 April 2026")

# mixed formats: try each
parse_date_time(c("2026-04-23", "04/23/2026", "April 23, 2026"),
                orders = c("ymd", "mdy", "Bdy"))
#> [1] "2026-04-23" "2026-04-23" "2026-04-23"

# date-time
ymd_hms("2026-04-23 14:30:00")
ymd_hm("2026-04-23 14:30")
```

Components:

```{r}
#| label: types-date-components
d <- ymd("2026-04-23")
year(d)         # 2026
month(d)        # 4
day(d)          # 23
wday(d)         # 5 (1 = Sunday by default)
wday(d, label = TRUE)  # Thu
```

Arithmetic:

```{r}
#| label: types-date-arithmetic
d + days(7)              # 2026-04-30
d %m+% months(3)         # 2026-07-23 (handles month length safely)

# differences
ymd("2026-12-31") - ymd("2026-01-01")  # Time difference of 364 days
as.numeric(ymd("2026-12-31") - ymd("2026-01-01"), units = "days")
```

Time zones:

```{r}
#| label: types-timezone
# parse as a specific time zone
t <- ymd_hms("2026-04-23 14:30:00", tz = "America/Los_Angeles")
t

# convert
with_tz(t, "UTC")          # same instant, different zone
force_tz(t, "UTC")         # same wall clock, different zone
```

`with_tz` changes the displayed zone without changing
the instant; `force_tz` changes the instant by
declaring the wall clock was always in the new zone.
Confusing them is a common source of 1- to 24-hour
errors.

## A runnable example: parsing a messy clinical field

The three tools of this chapter earn their place on the
free-text fields that clinical data are full of. To illustrate,
consider a single column recording a dosing instruction as the
site typed it, with the drug, the amount, and the start date
all run together and formatted inconsistently. We use a
regular expression to pull out the structured pieces, and
`lubridate` to parse the dates, which arrive in three
different formats.

```{r}
#| label: types-regex-dates
#| message: false
reticulate::py_require("pandas")
library(tidyverse)

dosing <- tibble(
  raw = c("Metformin 500mg started 2026-01-14",
          "Lisinopril 10 mg started 01/22/2026",
          "Atorvastatin 20mg started Feb 3 2026",
          "Aspirin 81 mg started 2026-02-19")
)

parsed <- dosing |>
  mutate(
    drug   = str_extract(raw, "^[A-Za-z]+"),
    dose   = as.integer(str_extract(raw, "\\d+(?=\\s?mg)")),
    date_s = str_extract(raw, "(?<=started ).*$"),
    start  = parse_date_time(date_s, orders = c("ymd", "mdy", "bdy")) |>
             as_date()
  ) |>
  select(drug, dose, start)

parsed
```

Three regular-expression constructs do the work. The anchor
`^[A-Za-z]+` takes the leading word, the drug. The lookahead
`\\d+(?=\\s?mg)` takes the digits that precede `mg` without
capturing the `mg` itself. The lookbehind `(?<=started )`
takes everything after the word `started`, which is the date
in whatever format the site used. `parse_date_time` then
tries each candidate format in turn. The result is a tidy
three-column table drawn from a single messy string.

### The same parsing in Python

`pandas` exposes the same operations through its `.str`
accessor and `to_datetime`. The regular expressions are
identical, because the regex dialect is shared. Only the
surrounding idiom differs.

```{python}
#| label: types-py
import pandas as pd

dosing = r.dosing            # the R tibble, via reticulate

out = pd.DataFrame({
    "drug": dosing["raw"].str.extract(r"^([A-Za-z]+)")[0],
    "dose": dosing["raw"].str.extract(r"(\d+)\s?mg")[0].astype(int),
    "start": pd.to_datetime(
        dosing["raw"].str.extract(r"started (.*)$")[0],
        format="mixed"),
})
out
```

The `format="mixed"` argument is `pandas`' counterpart to
supplying several `orders` to `parse_date_time`. It lets each
value be parsed by whichever format fits. The lesson is that
regex and date parsing are portable skills. The drug, dose,
and date fall out the same way in either language, and a
biostatistician who knows the constructs can work in
whichever the project uses.

::: {.callout-note icon=false collapse=true}
## Check your understanding: factor reference level

**Question.** You fit
`glm(outcome ~ treatment, family = binomial)` where
`treatment` is a factor with levels `"placebo"`,
`"low"`, `"high"`. The default reference is alphabetical
(`"high"`). The intercept estimates 18.2 and the
`treatmentplacebo` coefficient is -3.1. What is the
mean outcome under high-dose treatment?

**Answer.**

The intercept estimates the linear-predictor value at the
reference level (`"high"`). The high-dose mean on the link
scale is therefore 18.2 (or its inverse-link transform,
depending on the family). The `treatmentplacebo` coefficient
(-3.1) is the difference from high-dose, so the placebo mean
on the link scale is 18.2 - 3.1 = 15.1.

For clinical reporting, it is usually clearer to set the
reference deliberately to `"placebo"`, so the intercept gives
the placebo baseline and the treatment coefficients give the
increments from baseline:

```{r}
#| label: types-reference-level
treatment <- c("high", "placebo", "low", "high", "placebo")

treatment <- factor(treatment,
                    levels = c("placebo", "low", "high"))

levels(treatment)      # placebo is now the reference

```

The default alphabetical ordering is, unfortunately, a
frequent source of misreported effect direction.

:::

## Common pitfalls

**Excel dates as serial numbers.** A date column from
Excel may arrive as `45405` (= 2024-04-23 in Excel's
1900-epoch numbering). Parse with:

```{r}
#| label: types-excel-date
janitor::excel_numeric_to_date(45405)
#> [1] "2024-04-23"
```

Verify: any 'date' column whose values are 5-digit
numbers should ring an alarm.

**Mixed-locale month names.** `lubridate::dmy()` can
read 'avril' (French April) only if the locale
supports it. For multi-language data:

```r
lct <- Sys.getlocale("LC_TIME")
Sys.setlocale("LC_TIME", "fr_FR.UTF-8")
dmy("23 avril 2026")
Sys.setlocale("LC_TIME", lct)             # restore
```

Or use explicit format strings.

**Time-zone-naïve datetimes.** A datetime read without
a time zone defaults to UTC (lubridate) or the system
zone (base R). Be explicit; never assume.

**Daylight-saving transitions.** Adding 24 hours and
adding 1 day are not always the same:

```{r}
#| label: types-dst
ymd_hms("2026-03-09 01:00:00", tz = "US/Pacific") + days(1)
#> [1] "2026-03-10 01:00:00 PDT"
ymd_hms("2026-03-09 01:00:00", tz = "US/Pacific") + hours(24)
#> [1] "2026-03-10 02:00:00 PDT"
```

The first is a calendar day. The second is 24 wall-clock
hours, which crosses a DST boundary. Use whichever notion of
a day is appropriate for the analysis at hand.

::: {.callout-note icon=false collapse=true}
## Check your understanding: which date bug is silent?

**Question.** Three things go wrong with dates in this
chapter: a mixed-format column parsed with a single
format, an Excel serial number read as a number, and
`days(1)` used across a daylight-saving boundary. Which of
the three will your code tell you about, and which will it
not?

**Answer.**

Only the first announces itself, and even then partially.
`as.Date(x, format = "%Y-%m-%d")` on `03/14/2026` returns
`NA`, so the failure is visible if anyone counts the `NA`s
afterward. If the pipeline filters missing dates without
counting, it disappears too.

The other two are silent by construction.

An Excel serial date is a perfectly valid number. `45405`
reads as 45405, sorts, means, and plots without complaint;
nothing marks it as a date at all. It is only wrong
relative to an intention the data do not record.

The DST case is worse, because both answers are correct.
`d + days(1)` gives the same clock time the next calendar
day; `d + hours(24)` gives the instant 24 hours later.
Across a spring-forward boundary these differ by an hour,
and which you want depends on whether your interval is a
*calendar* day or a *duration*. R cannot know which you
meant, so it cannot warn you.

The lesson generalizes past dates: a bug that produces
`NA` is cheap, and a bug that produces a plausible number
is expensive. Assertions exist for the second kind.
:::

## Common antipatterns

**Hand-coded `ifelse` chains for date parsing**:

```{r}
#| label: types-antipattern-dates
x <- c("04/23/2026", "2026-04-23")

# bad: ifelse strips the Date class and returns bare numbers
ifelse(grepl("/", x),
       as.Date(x, format = "%m/%d/%Y"),
       as.Date(x, format = "%Y-%m-%d"))

# good
parse_date_time(x, orders = c("mdy", "ymd"))
```

**Letting characters become factors silently in old
code**:

```{r}
#| label: types-read-strings
#| message: false
f <- file.path(tempdir(), "file.csv")
write.csv(data.frame(site = "A", n = 3), f, row.names = FALSE)

# always specify
class(read.csv(f, stringsAsFactors = FALSE)$site)
class(readr::read_csv(f, show_col_types = FALSE)$site)  # readr default
```

**Comparing factors by value when levels differ**:

```{r}
#| label: types-factor-compare
f1 <- factor("a", levels = c("a", "b"))
f2 <- factor("a", levels = c("a", "c"))
f1 == f2                          # works, but levels differ
# better: convert to character for cross-table comparisons
as.character(f1) == as.character(f2)
```

## Worked example: cleaning a CRF export

```{r}
#| label: types-worked-cleanup
#| message: false
#| warning: false
library(tidyverse)
library(janitor)
library(lubridate)
library(forcats)

# A stand-in for the messy export: mixed date formats, one Excel
# serial date, inconsistent treatment spellings, a rare site that
# should be lumped, and blank sex entries. Twenty-four rows, so that
# fct_lump_min's threshold of 10 has something to bite on.
crf_path <- file.path(tempdir(), "crf_export.csv")
set.seed(3)
tibble(
  `Visit Date`       = rep(c("2026-01-14", "02/19/2026",
                             NA, "2026-03-02"), 6),
  `Visit Date Excel` = rep(c(NA, NA, 45405, NA), 6),
  Treatment          = rep(c("placebo", "PBO",
                             "low dose", "HIGH DOSE"), 6),
  Site               = c(rep("A", 12), rep("B", 10), "Y", "Z"),
  Sex                = rep(c("F", "M", "", "F"), 6)
) |>
  write_csv(crf_path, na = "")

# na = "NA" keeps blank Sex as an empty string rather than NA, which
# is what the fct_recode below is there to handle.
raw <- read_csv(crf_path, na = "NA", show_col_types = FALSE)

clean <- raw |>
  clean_names() |>
  mutate(
    # parse mixed-format date column
    visit_date = parse_date_time(visit_date,
                                 orders = c("ymd", "mdy")),

    # convert numeric Excel dates if any
    visit_date = if_else(is.na(visit_date) & !is.na(visit_date_excel),
                         excel_numeric_to_date(visit_date_excel),
                         visit_date),

    # standardize treatment text
    treatment = str_to_title(treatment),
    treatment = str_replace(treatment, "^Pbo$", "Placebo"),

    # convert to factor with deliberate reference
    treatment = factor(treatment,
                       levels = c("Placebo", "Low Dose", "High Dose")),

    # collapse small categories
    site = fct_lump_min(site, min = 10, other_level = "Other"),

    # recode missing-as-unknown
    sex = fct_recode(factor(sex), "Unknown" = "")
  ) |>
  filter(!is.na(visit_date), !is.na(treatment))

count(clean, treatment, site, sex)
```

The cleaned data have dates parsed from mixed formats,
treatment standardized and explicitly factored, small sites
collapsed, and missing sex coded explicitly. That is, the
script documents every transformation as code, rather than
leaving it to memory.

## Collaborating with an LLM on types

LLMs handle these packages well. The trap, as far as we have
observed, is silent type coercion.

**Prompt 1: parsing dates.** Paste a sample of date
strings (10 examples covering the formats present)
and ask: 'parse these robustly, flagging any that fail.'

*What to watch for.* `parse_date_time()` with multiple
orders is the canonical answer. If the LLM produces
nested `ifelse` or `case_when`, push back.

*Verification.* Run on the full column. Count NAs
produced; investigate any unexpected ones.

**Prompt 2: regex extraction.** Describe the pattern
you want and provide examples. Ask the LLM to write a
regex.

*What to watch for.* Test the regex on edge cases the
LLM may not have considered (empty strings, unusual
separators, mixed case). LLM regexes often fail at
the margins.

*Verification.* `str_detect` and `str_extract` on a
test set; count mismatches.

**Prompt 3: factor reference level.** Paste a
modeling formula and ask: 'is the default reference
level for `treatment` clinically sensible? If not,
how should it be set?'

*What to watch for.* The LLM should know that
alphabetical reference is rarely what you want.
Specifying via `factor(..., levels = ...)` or
`fct_relevel` is the canonical fix.

*Verification.* Re-fit the model with the explicit
reference; check coefficients match the intended
interpretation.

## Principle in use

Three habits keep text, categories, and dates from becoming
entries on the project's bug list:

1. **Convert to factor late.** Cleaning in
   character; modeling in factor. Specify reference
   levels deliberately.
2. **Parse dates with named formats.**
   `parse_date_time` with explicit orders, not
   guesses. Watch for Excel serial numbers.
3. **Be explicit about time zones.** A datetime
   without a zone is a bug waiting to happen.

## Exercises

1. Using `palmerpenguins::penguins`, reorder the
   `species` factor by mean body mass (ascending)
   using `fct_reorder()`. Produce a bar plot that
   reflects the new order.
2. A character column contains hospital names like
   `'General Hospital Main Campus'`,
   `'Gen. Hospital'`, `'GENHOSP'`. Write a
   regex-based cleaning pipeline that collapses
   these into a single canonical level.
3. A dataset has a `visit_date` column in mixed
   formats. Write a pipeline that parses everything
   with `parse_date_time()` and flags rows that fail
   to parse.
4. Write a function that takes any character column
   and returns a tibble of value, count, and
   proportion.
5. Create a dataset with a datetime in PST and convert
   it to UTC two ways: with `with_tz` and
   `force_tz`. Explain the difference.

**On your own data.** No answer key; the answer depends
on your data.

6. Use the function from exercise 4 to audit the
   categorical variables in a recent analysis of
   yours. Report any level you did not expect.

::: {.callout-tip icon=false collapse=true}
## Solutions to exercises 1 to 5

**1. Reorder a factor by a summary of another variable.**

```{r}
#| label: soln-14-1
#| fig-cap: "Mean body mass by species, with the factor reordered so
#|   the bars ascend. The ordering is a property of the data, not of
#|   the alphabet."
#| fig-alt: "Bar chart of mean body mass for three penguin species,
#|   ascending from Chinstrap through Adelie to Gentoo."
#| fig-width: 5
#| fig-height: 3
#| message: false
library(ggplot2)

penguins |>
  filter(!is.na(body_mass_g)) |>
  mutate(species = fct_reorder(species, body_mass_g, mean)) |>
  ggplot(aes(species, body_mass_g)) +
  stat_summary(fun = mean, geom = "col")
```

`fct_reorder()` changes the level order, and the plot
follows because ggplot2 draws factors in level order.
Sorting the data frame instead would not have worked.

**2. Collapsing spelling variants to a canonical level.**

```{r}
#| label: soln-14-2
hosp <- c("General Hospital Main Campus", "Gen. Hospital",
          "GENHOSP", "Mercy Regional")

canonical <- case_when(
  str_detect(str_to_lower(str_remove_all(hosp, "[^A-Za-z]")),
             "^gen") ~ "General Hospital",
  .default = hosp
)
tibble(raw = hosp, canonical)
```

Stripping punctuation and case before matching is what
lets one pattern cover all three variants. Note that this
collapses `GENHOSP` correctly and leaves `Mercy Regional`
alone, which is the behavior to check.

**3. Parsing mixed formats and flagging failures.**

```{r}
#| label: soln-14-3
#| warning: false
visit_raw <- c("2026-01-14", "02/19/2026", "Feb 3 2026",
               "not recorded")

parsed <- tibble(visit_raw) |>
  mutate(
    visit_date = parse_date_time(visit_raw,
                                 orders = c("ymd", "mdy", "Bdy")),
    failed = is.na(visit_date)
  )
parsed
```

The fourth value cannot parse, and the point of the
`failed` column is that it is now data rather than a
warning nobody reads. Filtering silently on
`!is.na(visit_date)` would lose the count.

**4. A reusable level audit.**

```{r}
#| label: soln-14-4
audit_levels <- function(df, col) {
  df |>
    count({{ col }}, sort = TRUE) |>
    mutate(proportion = n / sum(n))
}

audit_levels(penguins, island)
```

**5. `with_tz` against `force_tz`.**

```{r}
#| label: soln-14-5
t <- ymd_hms("2026-04-23 14:30:00", tz = "America/Los_Angeles")

with_tz(t, "UTC")    # same instant, clock reads differently
force_tz(t, "UTC")   # same clock reading, different instant
```

`with_tz` preserves the moment and relabels it; `force_tz`
preserves the digits and moves the moment. Use `with_tz`
when combining data recorded in different zones, and
`force_tz` only to repair a timestamp that was stamped with
the wrong zone to begin with.
:::

## Further reading

- [@wickham2023r4ds] Chapters 14--18, `forcats`,
  `stringr`, `lubridate`.
- [@bryan2019stat545] Chapters 10--13, detailed
  type-specific coverage.
- [@grolemund2011dates], the `lubridate` paper, for
  the date and time-zone handling in this chapter.
- The `regular-expressions.info` website,
  comprehensive regex reference.

## Prerequisites answers {#sec-quiz-answers-14}

1. Before R 4.0, `read.csv()` and `data.frame()` converted
   character columns to factors by default. This caused
   surprise when levels were invented, ordered, or appeared
   where a character string was expected, for example as
   merge keys. Downstream analysts had to remember
   `stringsAsFactors = FALSE` or suffer silent coercion.
   R 4.0 made `FALSE` the default, and the old default is
   remembered now only when reading elderly code.
2. `str_detect(string, pattern)` returns a logical vector of
   the same length as `string`: `TRUE` where the pattern
   matched, `FALSE` elsewhere. `str_extract()` returns a
   character vector, namely the first matching substring per
   input, or `NA_character_` where there was no match. In
   short, `str_detect` is for filtering and `str_extract` is
   for pulling out parts.
3. `parse_date_time(x, orders = c('ymd', 'mdy', 'Bdy'))`
   (or `readr::parse_date` with multiple format strings)
   is the tool. `lubridate::parse_date_time()` tries each
   order against each input and returns the first that
   parses. Inputs that match no format produce `NA`, which
   lets you flag and inspect them.
