# Joining and Reshaping {#sec-joins}

::: {.bookepigraph}
```{=latex}
\begin{flushright}\small
```
*Only connect.*

[E. M. Forster]{.smallcaps}, *Howards End* (1910)
```{=latex}
\end{flushright}
```
:::

::: {.callout-note}
## Sources

Stat 545 Chapters 14--16 (Jenny Bryan, UBC); the
`dplyr` two-table verbs reference.
:::

## Prerequisites

Answer the following questions to see if you can bypass
this chapter. You can find the answers at the end of the
chapter in @sec-quiz-answers-15.

1. What is the difference in row count between
   `inner_join()` and `left_join()` when the right
   table is missing some keys that appear in the left
   table?
2. What does `anti_join()` return, and in a biomedical
   cleanup workflow, what question does it answer?
3. Given two tables each with a `patient_id` column
   and a many-to-one relationship (many rows per
   patient in table A, one row per patient in table B),
   how do you attach the table-B row to every table-A
   row?

## Learning objectives

By the end of this chapter you should be able to:

- Select the appropriate `_join()` variant for a given
  question (inner, left, right, full, semi, anti).
- Recognize and avoid the pitfalls of many-to-many
  joins, including using `dplyr` 1.1.0+'s
  `relationship` argument.
- Use `dplyr::rows_*` for upsert-style operations.
- Apply `pivot_longer()` and `pivot_wider()` for
  wide/long reshaping beyond the basics.
- Reason about keys, uniqueness, and referential
  integrity in a small relational schema.
- Verify joins using `anti_join()` audits.

## Orientation

Almost every real biostatistical project involves joining
tables: demographics joined to outcomes, labs joined to
medications, clinical data joined to imaging indices. The
operation is one function call, and it is nonetheless
among the most reliable sources of error in applied work.

The reason is worth stating precisely, because it explains
what this chapter is defending against. A join has two
failure modes, and neither raises an error. It can drop rows,
when a key present on one side is absent on the other. The
analysis then silently proceeds on a smaller, and possibly
different, cohort than the one described in the methods
section. Or it can multiply rows, when a key that should be
unique is not. A patient then contributes several times to
an estimate that assumed a single contribution. The first
shrinks the denominator; the second corrupts the standard
errors. Both produce a data frame that looks exactly like the
one you wanted, and both survive every check that a person
makes by glancing at the output.

What catches them is not care but arithmetic. That is, a
stated expectation about how many rows the result should
have, checked against how many it does. That is the habit
this chapter is built around. The `relationship` argument
introduced in `dplyr` 1.1.0 is the tooling that makes
stating the expectation cheap enough to do every time.

## The statistician's contribution {#sec-joins-human}

Every join in this chapter is one function call. Joins are
among the most common sources of silent error in applied
work precisely because the function call is not where the
difficulty lives:

**Verify after every join.** A join that says it
returned the right number of rows usually did. A join
that returns 'about the right number' silently is
where the bugs live. After every non-trivial join,
print `nrow()` and check it against expectations.

**Detect many-to-many joins.** When you intend a
one-to-one join and accidentally trigger a
many-to-many, dplyr (1.1.0+) issues a warning. Take
the warning seriously: a many-to-many on a key that
should be unique multiplies row counts. The
`relationship` argument lets you assert the expected
cardinality and error if it does not hold.

**Use `anti_join` to find what is missing.** Two CRF
files that should cover the same patients usually do
not, exactly. `anti_join(file_a, file_b, by = "id")`
returns the patients in `file_a` not in `file_b`, the
quickest way to find missing data, ID typos, or
cohort mismatches. Run it before the analytic join.

**Document the join keys.** A join on `patient_id`
needs both tables to use the same identifier
representation (string '001' is not the same as
numeric 1). Document the canonical ID format; check
it on every read.

These judgments determine whether the joined data
mean what you think they mean.

## The `_join` family

dplyr's two-table verbs:

| Function       | Returns                                            |
|----------------|----------------------------------------------------|
| `inner_join`   | Rows with a match in both                          |
| `left_join`    | All rows from `x`, NA in `y` columns where no match |
| `right_join`   | All rows from `y`, NA in `x` columns where no match |
| `full_join`    | All rows from both, NA in non-matching columns      |
| `semi_join`    | Rows of `x` with a match in `y` (no `y` cols)       |
| `anti_join`    | Rows of `x` with no match in `y`                    |

`semi_join` and `anti_join` are filtering joins: they
filter `x` based on `y` without adding `y`'s columns.
The other four are mutating joins: they add columns.
The choice among them reduces to two questions, shown
in @fig-join-choice: do you want to keep rows that
fail to match, and if so, from which side.

```{mermaid}
%%| echo: false
%%| label: fig-join-choice
%%| fig-alt: "Decision flowchart for choosing a join. The first
%%|   question asks whether to keep rows with no match: no gives
%%|   inner_join. Yes leads to a second question asking which side
%%|   the unmatched rows come from: left x gives left_join, right y
%%|   gives right_join, and both gives full_join."
%%| fig-cap: "Choosing a join. The first question is whether unmatched
%%|   rows should survive; the second is from which table. The
%%|   filtering joins (semi, anti) answer a different question again:
%%|   keep or drop rows of x by their presence in y, adding no columns."
flowchart TD
  Q1{"Keep rows with<br/>no match?"}
  Q1 -->|"No"| INNER["inner_join"]
  Q1 -->|"Yes"| Q2{"Unmatched from<br/>which side?"}
  Q2 -->|"Left x"| LEFT["left_join"]
  Q2 -->|"Right y"| RIGHT["right_join"]
  Q2 -->|"Both"| FULL["full_join"]
```

The `dplyr` package ships two toy tables, `band_members`
and `band_instruments`, that make the row-count
differences concrete. We run the four mutating and
filtering variants and look at what each returns.

```{r}
#| label: joins-band
#| message: false
library(dplyr)

band_members
band_instruments

inner_join(band_members, band_instruments, by = "name")  # John, Paul
left_join(band_members, band_instruments, by = "name")   # + Mick (NA)
full_join(band_members, band_instruments, by = "name")   # + Keith (NA)
anti_join(band_members, band_instruments, by = "name")   # Mick only
```

The choice of join is determined by the question:

- 'Patients with both demographics and outcome
  available': `inner_join`.
- 'All patients, with their outcomes if any':
  `left_join` (demographics on the left).
- 'Find patients with demographics but no outcome':
  `anti_join`.
- 'Combine all patients from both files, even those
  in only one': `full_join`.

### The same joins in pandas

The `pandas` library folds the mutating joins into a
single `merge` call whose `how` argument selects the
variant. The filtering joins have no direct `merge`
equivalent; they are expressed with `isin`.

| dplyr | pandas |
|---|---|
| `inner_join(x, y, by = "id")` | `x.merge(y, on="id", how="inner")` |
| `left_join(x, y, by = "id")` | `x.merge(y, on="id", how="left")` |
| `right_join(x, y, by = "id")` | `x.merge(y, on="id", how="right")` |
| `full_join(x, y, by = "id")` | `x.merge(y, on="id", how="outer")` |
| `semi_join(x, y, by = "id")` | `x[x["id"].isin(y["id"])]` |
| `anti_join(x, y, by = "id")` | `x[~x["id"].isin(y["id"])]` |

```{r}
#| label: joins-bind
#| echo: false
reticulate::py_require("pandas")
band_members <- dplyr::band_members
band_instruments <- dplyr::band_instruments
```

```{python}
#| label: joins-pandas
import pandas as pd

band_members = r.band_members
band_instruments = r.band_instruments

# left join keeps Mick, who plays no instrument in the table
band_members.merge(band_instruments, on="name", how="left")
```

## Many-to-many joins

A many-to-many join occurs when both tables have
duplicate keys: each row in the left can match
multiple rows in the right, and vice versa. The
result is a Cartesian product of matches per key.

```{r}
#| label: joins-many-to-many
#| message: false
#| warning: true
library(dplyr)

a <- tibble(id = c(1, 1, 2), x = c("a", "b", "c"))
b <- tibble(id = c(1, 1, 2), y = c("p", "q", "r"))

# Five rows, not three: each id-1 in a joins each id-1 in b, giving
# four, plus the single id-2 match. dplyr warns because this is
# usually a key error rather than an intention.
inner_join(a, b, by = "id")
```

dplyr 1.1.0+ warns by default. The fix is to declare
the expected relationship:

```{r}
#| label: joins-relationship
#| error: true
# explicit; no warning, even with duplicates
inner_join(a, b, by = "id", relationship = "many-to-many")

# errors if duplicates exist on either side
inner_join(a, b, by = "id", relationship = "one-to-one")
```

Common pattern: each `patient_id` should appear once
in the demographics table. Joining demographics
(many-to-one with respect to a visits table):

```{r}
#| label: joins-fixtures
#| echo: false
# Three small tables used by the examples that follow: one row per
# patient in demographics, several visits per patient, several labs
# per visit. P004 appears in visits but not demographics, which the
# anti-join audit below is there to catch.
demographics <- tibble(
  patient_id = c("P001", "P002", "P003"),
  age        = c(64, 71, 52),
  sex        = c("F", "M", "F")
)
visits <- tibble(
  visit_id   = sprintf("V%02d", 1:6),
  patient_id = c("P001", "P001", "P002", "P002", "P003", "P004"),
  sbp        = c(148, 141, 155, 149, 132, 139)
)
labs <- tibble(
  visit_id  = c("V01", "V01", "V02", "V03"),
  test_code = c("HGB", "CRE", "HGB", "HGB"),
  value     = c(13.1, 0.9, 13.4, 11.8)
)
```

```{r}
#| label: joins-many-to-one
visits |>
  left_join(demographics, by = "patient_id",
            relationship = "many-to-one")
# errors if a patient_id appears more than once in demographics
```

The assertion catches data-quality bugs at the join
site, where they are easiest to debug.

::: {.callout-note icon=false collapse=true}
## Check your understanding: silent row inflation

**Question.** You join `visits` (10,000 rows) to
`demographics` (1,000 patients) on `patient_id`. The
result has 12,500 rows. What likely happened?

**Answer.**

The demographics table has duplicate `patient_id`
values. Each duplicate creates extra rows in the join
output. With 10,000 visits and 1,000 patients, a
clean many-to-one join produces 10,000 rows. 12,500
indicates roughly 25% of the visits' patients have
duplicates in the demographics table.

Diagnose:

```{r}
#| label: joins-key-duplicates
demographics |> count(patient_id) |> filter(n > 1)
```

Fix the duplication in demographics, or de-duplicate
before the join (`distinct(patient_id, .keep_all = TRUE)`).
Adding `relationship = "many-to-one"` to the join
catches this at the join site.

:::

## Keys and referential integrity

Before joining, verify the join columns:

```{r}
#| label: joins-key-audit
# uniqueness of keys
demographics |> distinct(patient_id) |> nrow()
demographics |> nrow()
# These should match for one-row-per-patient

# stopifnot for assertion
stopifnot(!any(duplicated(demographics$patient_id)))

# anti-join audit
visits |> anti_join(demographics, by = "patient_id") |>
  distinct(patient_id)
# patients in visits with no demographics row
```

`assertthat` and `assertr` provide more elaborate
assertions. For most work, `stopifnot()` plus
manual `anti_join` audits cover the cases.

For composite keys, name all columns:

```{r}
#| label: joins-compound-key
reference_ranges <- tibble(
  test_code = c("HGB", "HGB", "CRE"),
  sex       = c("F", "M", "F"),
  age_group = c("50-70", "50-70", "50-70"),
  lower     = c(12.0, 13.5, 0.6),
  upper     = c(15.5, 17.5, 1.1)
)

labs |>
  left_join(visits, by = "visit_id") |>
  left_join(demographics, by = "patient_id") |>
  mutate(age_group = "50-70") |>
  left_join(reference_ranges,
            by = c("test_code", "sex", "age_group"))
```

Type matching is strict: `patient_id = "001"` does
not match `patient_id = 1`. Convert to a canonical
type before joining.

::: {.callout-note icon=false collapse=true}
## Check your understanding: the join that lost rows

**Question.** You `left_join` a 500-row visits table to a
demographics table and get 500 rows back, which is what
you expected. A colleague says this proves nothing. Are
they right?

**Answer.**

They are. A row count matching your expectation rules out
duplication and rules out nothing else.

Every `patient_id` in visits could have failed to match,
and `left_join` would still return exactly 500 rows, each
with `NA` in every demographics column. The count is
identical whether the join worked perfectly or not at all.
This is the specific case where the type trap above bites:
character `"001"` against numeric `1` matches nothing, and
the row count never moves.

Two checks catch it and the count does not. Count the
`NA`s in a column that came from the right-hand table,
which should be zero if every visit has demographics. And
run the anti-join:

```r
visits |> anti_join(demographics, by = "patient_id") |>
  distinct(patient_id)
```

An empty result means every key matched. A result listing
all 500 patients means the join silently did nothing.

The general point is that `nrow()` is a structural check,
in the sense @sec-testing gives that term, and structural
checks pass on wrong answers.
:::

## Upsert operations with `rows_*`

For row-by-row updates of a table from another (the
'upsert' pattern):

```{r}
#| label: joins-rows-verbs
#| error: true
target <- tibble(id = 1:3, value = c("a", "b", NA))
source <- tibble(id = 3:4, value = c("c", "d"))

# insert new rows; error on key collision
rows_insert(target, source, by = "id")

# update existing rows; error if key not in target
rows_update(target, source, by = "id")

# upsert: insert new, update existing
rows_upsert(target, source, by = "id")

# patch: only update where target has NA
rows_patch(target, source, by = "id")
```

Useful for updating a master patient list from a new
extract: `rows_upsert` adds new patients and updates
existing ones; `rows_patch` fills in missing values
without overwriting existing.

## Reshaping for joining

Some joins need wide-format data; others need long.
The wrangling chapter (@sec-wrangling) covers `pivot_longer`
and `pivot_wider`. For joins, the typical pattern:

- Convert each source to long format with consistent
  column names.
- Stack with `bind_rows` rather than join (when sources
  are alternative measurements of the same kind).
- Pivot wide for the modeling step.

For two CRF files from sites A and B with different
column conventions:

```r
a_long <- read_csv("siteA.csv") |>
  clean_names() |>
  rename(bp_systolic = systolic_bp) |>
  pivot_longer(starts_with("bp_"), names_to = "measure")

b_long <- read_csv("siteB.csv") |>
  clean_names() |>
  pivot_longer(matches("^(bp|sbp|dbp)_"), names_to = "measure")

combined <- bind_rows(a_long, b_long)
```

`bind_rows` is for stacking; `bind_cols` is for
side-by-side concatenation (rarely correct;
`*_join` is usually what you want). `inner_join`/
`left_join` are for joining on shared keys.

## Worked example: a multi-source patient cohort

```r
library(tidyverse)
library(janitor)

# read three sources
demographics <- read_csv("data/demographics.csv") |> clean_names()
visits       <- read_csv("data/visits.csv") |> clean_names()
labs         <- read_csv("data/labs.csv") |> clean_names()

# verify keys are unique where expected
stopifnot(!anyDuplicated(demographics$patient_id))
stopifnot(!anyDuplicated(labs$visit_id))

# audit: patients in visits but no demographics
missing_demo <- visits |>
  anti_join(demographics, by = "patient_id") |>
  distinct(patient_id)

if (nrow(missing_demo) > 0) {
  warning("Patients in visits with no demographics: ",
          paste(missing_demo$patient_id, collapse = ", "))
}

# build the analytic dataset
analytic <- visits |>
  left_join(demographics, by = "patient_id",
            relationship = "many-to-one") |>
  left_join(labs, by = "visit_id",
            relationship = "many-to-many") |>
  mutate(
    age_group = cut(age, breaks = c(0, 40, 65, Inf))
  )

# final row-count check
stopifnot(nrow(analytic) == nrow(visits))   # joins should be 1-1 on visits

saveRDS(analytic, "data/derived/analytic.rds")
```

The script cleans names, verifies key uniqueness, runs an
`anti_join` audit, performs the joins with explicit
`relationship` declarations, and asserts the final row
count. A reviewer can read it and trust the join semantics
without re-deriving them.

## Collaborating with an LLM on joins

LLMs handle joins well. The trap, in our experience, is
silent join inflation.

**Prompt 1: choosing a join.** Describe the question
and the two tables; ask the LLM to recommend a join
type.

*What to watch for.* The LLM should map question to
join correctly. If it suggests `merge()` from base R,
prefer the dplyr `_join()` family for clarity.

*Verification.* Compare the join output's row count
to the expected count. If they differ, debug.

**Prompt 2: diagnosing inflated row counts.** Paste
the tables and the join, ask: 'why does the result
have more rows than I expected?'

*What to watch for.* The standard cause is many-to-
many. The LLM should know to check for duplicate keys
on each side. If it does not, run
`x |> count(key) |> filter(n > 1)` yourself.

*Verification.* The duplicate keys, if found, are the
culprits. Resolve by deduplicating, by composite
keying, or by accepting the many-to-many.

**Prompt 3: writing assertions.** Describe the
expected join cardinality and ask the LLM to add
assertions.

*What to watch for.* `stopifnot()` for simple checks;
`relationship = ...` in the join itself is the modern
idiom. The LLM should use the latter when applicable.

*Verification.* Run the assertions on real data; they
should pass. If they fail, debug the data, not the
assertions.

## Principle in use

Three habits stand between a join and a silent bug in the
downstream analysis:

1. **Pick the join that matches the question.**
   Inner for matched-only; left for left-preserving;
   anti for finding what is missing. The wrong join is
   a silent bug.
2. **Verify keys before joining.** Uniqueness checks
   and `anti_join` audits catch bugs before they
   propagate.
3. **Assert join cardinality.** dplyr's `relationship`
   argument errors loudly when expectations are
   wrong. Use it.

## Exercises

1. Use each of `inner_join`, `left_join`, `full_join`,
   and `anti_join` with the built-in
   `dplyr::band_members` and `dplyr::band_instruments`
   to produce four different tibbles. Write one
   sentence per result.
2. Construct a scenario where an unintended
   many-to-many join duplicates rows. Write an
   assertion that catches the duplication before
   downstream analysis.
3. Two wide tables represent patient visits at sites
   A and B with different column-name conventions.
   Reshape both to long and stack them so that
   downstream analysis treats them uniformly.
4. Use `rows_upsert` to merge two cohorts with
   overlapping patients (some new, some updated).
   Verify the result has the expected number of
   rows.

**On your own data.** These have no answer key.

5. Take a recent multi-table analysis of yours.
   Audit each join with `anti_join` to identify
   missing keys; document any unexpected gaps.
6. Review a colleague's multi-table script (or one
   from a public analysis repository) for join
   safety alone: for every join, determine the
   intended cardinality, check whether a
   `relationship` argument asserts it, and run an
   `anti_join` audit on the keys. Report what you
   find as you would in a pull-request comment
   (@sec-git-teams), naming the specific line and
   the specific risk.

::: {.callout-tip icon=false collapse=true}
## Solutions to exercises 1 to 4

**1. The four joins on `band_members` and `band_instruments`.**

```{r}
#| label: soln-15-1
#| message: false
inner_join(band_members, band_instruments, by = "name")
left_join(band_members, band_instruments, by = "name")
full_join(band_members, band_instruments, by = "name")
anti_join(band_members, band_instruments, by = "name")
```

`inner_join` keeps only John and Paul, who appear in both.
`left_join` keeps all three band members, giving Mick an
`NA` instrument. `full_join` keeps Keith as well, giving
him an `NA` band. `anti_join` returns Mick alone: the band
member with no instrument row, and it adds no columns.

**2. An unintended many-to-many, and the assertion that
catches it.**

```{r}
#| label: soln-15-2
#| error: true
#| warning: true
# demographics has a duplicated patient, perhaps from a
# re-consent that was appended rather than replaced
demo_dup <- tibble(patient_id = c("P001", "P001", "P002"),
                   age = c(64, 65, 71))
vis <- tibble(patient_id = c("P001", "P002"), sbp = c(148, 155))

# the silent version: two rows become three
left_join(vis, demo_dup, by = "patient_id")

# the assertion, before anything downstream runs
stopifnot(!anyDuplicated(demo_dup$patient_id))
```

The join itself does not fail; it quietly returns more rows
than it received. The `stopifnot()` fails loudly, which is
the behavior you want. Declaring
`relationship = "many-to-one"` on the join achieves the
same thing at the point of use.

**3. Reconciling two site conventions.**

```{r}
#| label: soln-15-3
#| message: false
library(tidyr)

site_a <- tibble(patient_id = "A01",
                 systolic_bp = 148, diastolic_bp = 88)
site_b <- tibble(patient_id = "B01",
                 sbp_1 = 155, dbp_1 = 91)

a_long <- site_a |>
  rename(bp_systolic = systolic_bp, bp_diastolic = diastolic_bp) |>
  pivot_longer(starts_with("bp_"), names_to = "measure")

b_long <- site_b |>
  rename(bp_systolic = sbp_1, bp_diastolic = dbp_1) |>
  pivot_longer(starts_with("bp_"), names_to = "measure")

bind_rows(a_long, b_long, .id = "site")
```

Renaming to a shared vocabulary *before* the pivot is what
makes `bind_rows()` safe. Stacking first and reconciling
after leaves you matching on strings that do not agree.

**4. `rows_upsert`, and the row-count check.**

```{r}
#| label: soln-15-4
cohort_1 <- tibble(patient_id = c("P001", "P002"),
                   status = c("enrolled", "enrolled"))
cohort_2 <- tibble(patient_id = c("P002", "P003"),
                   status = c("withdrawn", "enrolled"))

merged <- rows_upsert(cohort_1, cohort_2, by = "patient_id")
merged

# two originals plus one genuinely new patient
stopifnot(nrow(merged) == 3)
```

P002 is updated in place rather than duplicated, and P003
is inserted. The row count is the check that distinguishes
an upsert from an accidental `bind_rows()`, which would
have produced four rows and two conflicting records for
P002.
:::

## Further reading

- [@wickham2023r4ds] Chapters 19--20, the `dplyr`
  two-table verbs.
- [@bryan2019stat545] Chapters 14--16, joins and
  lookups.
- The `dplyr` 1.1.0 release notes for the
  `relationship` argument and the many-to-many
  warning.

## Prerequisites answers {#sec-quiz-answers-15}

1. `inner_join()` returns only rows whose key matches
   in both tables. `left_join()` returns every row
   from the left table; rows with no match on the
   right get `NA` in the right-hand columns. If the
   left table has 100 rows with keys A--J and the
   right table has only keys A--F, `inner_join`
   returns 60 rows and `left_join` returns 100 rows
   (40 with `NA`s).
2. `anti_join(x, y)` returns rows of `x` whose key is
   *not* present in `y`. In a biomedical cleanup workflow it
   answers: 'Which patients appear in my outcome file but not
   in the lab data?' (or vice versa). It thereby exposes
   missing data, duplicate IDs, or mismatched cohorts. Run it
   before the analytic join, so that problems surface early.
3. `table_a |> left_join(table_b, by = 'patient_id', relationship = 'many-to-one')`.
   A many-to-one join attaches the single table-B
   row to every table-A row sharing the key. The
   `relationship` argument errors if the
   relationship does not hold (e.g., if `patient_id`
   is duplicated in table B).
