# Databases and SQL for Biostatisticians {#sec-databases}

::: {.bookepigraph}
```{=latex}
\begin{flushright}\small
```
*The most valuable commodity I know of is information.*

[Gordon Gekko]{.smallcaps}, *Wall Street* (1987)
```{=latex}
\end{flushright}
```
:::

::: {.callout-note}
## Why this chapter exists

Clinical data rarely arrive as a tidy CSV. They live in a
REDCap project, an institutional data warehouse, or an
electronic health record extract, and the biostatistician
reaches them through a query rather than a file. The
survey of peer programs (@sec-peer-survey) found
relational databases and SQL taught as a core skill in
several of the comparison courses, among them Berkeley
STAT 243 [@berkeley-stat243-2024], UNC BIOS 735
[@unc-bios735-2024], Michigan [@umich-biostat615-2024],
Minnesota [@umn-biostat-2024], and the Johns Hopkins
statistical-computing sequence [@hopkins-biostat776-2024],
as well as in both Software and Data Carpentry. This
chapter brings the Practicum to parity.
:::

## Prerequisites

Answer the following questions to see if you can bypass
this chapter. You can find the answers at the end of the
chapter in @sec-quiz-answers-15b.

1. What is the difference between querying a table with
   `DBI::dbGetQuery()` and reading a whole table into
   memory with `dbReadTable()`, and why does the
   distinction matter for a ten-million-row table?
2. In the `dbplyr` model of database access, what does a
   pipeline of `dplyr` verbs actually do before you call
   `collect()`, and where does the computation happen?
3. Why is it preferable, for reproducibility, to record
   the SQL query that produced an analytic dataset rather
   than to save the extracted CSV alone?

## Learning objectives

By the end of this chapter you should be able to:

- Connect to a database from R with `DBI` and an
  appropriate backend driver (`RSQLite`, `duckdb`,
  `RPostgres`).
- Write basic `SELECT` queries with filtering, joining,
  grouping, and aggregation.
- Express the same query with `dbplyr` using familiar
  `dplyr` verbs, and inspect the SQL it generates.
- Distinguish lazy from eager evaluation and use
  `collect()` deliberately.
- Reason about when to push work to the database and when
  to pull data into R.

## Orientation

A database is a store that lets many users query structured
data efficiently and safely. For the biostatistician, the
practical point is that the data are usually too large, too
sensitive, or too frequently updated to be handed around as
flat files. Querying at the source has, we would note, three
advantages: you pull only the rows and columns you need, you
leave the authoritative copy untouched, and you record, in the
query itself, exactly how your analytic dataset was derived.

This chapter is not a database course. It teaches the subset
of SQL and the `DBI`/`dbplyr` tooling that covers the
recurring needs of applied biostatistics: connect, filter,
join, aggregate, and extract. The deeper matters of schema
design, indexing, and transaction management belong to a
database text, and to your data-engineering colleagues.

## The statistician's contribution {#sec-databases-human}

The core of SQL is small enough to learn in a day. What
matters more is where you draw the line between the
database's work and R's:

**Query at the source, extract deliberately.** The
temptation is to pull the whole table into R and work
there, because R is where you are comfortable. On a small
table this is harmless. On a large one it is slow,
memory-hungry, and it obscures the derivation. Push the
filtering and aggregation to the database; pull only the
analytic dataset.

**The query is the provenance.** A CSV named
`analytic_data_final_v3.csv` records nothing about how it
was made. The SQL that produced it records everything.
Keep the query in the compendium, under version control,
and treat the extracted file as a build product in the
manner of @sec-wrangling's raw-data rule.

**Beware the silent row multiplication.** A join in SQL
inflates rows exactly as a join in `dplyr` does, and the
database will not warn you. The verification habit from
@sec-joins, count the rows and check against
expectation, applies with equal force here.

**Respect the boundary between analyst and warehouse.**
You usually have read-only access to an institutional
warehouse, and that is correct. You are a consumer of the
data, not its custodian. Do not write back; extract, and
do your derivations in your own space.

These judgments are what make database-backed analysis
reproducible and safe.

## Connecting with DBI

The `DBI` package provides a backend-independent interface;
a driver package supplies the connection to a specific
database engine. For teaching we use `RSQLite`, which
gives a complete SQL database in a single file or in
memory, with no server to configure. The same code works
against PostgreSQL (`RPostgres`), DuckDB (`duckdb`), or
SQL Server (`odbc`) by changing only the `dbConnect` line.

To make the example self-contained, we create an in-memory
database and populate it with a tiny clinical schema: a
one-row-per-patient demographics table and a
one-row-per-visit table of blood-pressure measurements.

```{r}
#| label: db-setup
#| message: false
library(DBI)
library(RSQLite)
library(dplyr)
library(dbplyr)

con <- dbConnect(RSQLite::SQLite(), ":memory:")

demographics <- data.frame(
  patient_id = 1:5,
  age        = c(64, 71, 58, 66, 73),
  arm        = c("Active", "Placebo", "Active", "Placebo", "Active")
)

visits <- data.frame(
  patient_id = c(1, 1, 2, 3, 3, 3, 4, 5),
  visit      = c(1, 2, 1, 1, 2, 3, 1, 1),
  sbp        = c(140, 138, 150, 145, 143, 141, 139, 152)
)

dbWriteTable(con, "demographics", demographics)
dbWriteTable(con, "visits", visits)

dbListTables(con)
```

In a real setting the `dbConnect` call would name a
server, a database, and credentials, ideally drawn from an
environment variable or a `config` file rather than
written into the script:

```r
con <- dbConnect(
  RPostgres::Postgres(),
  host     = "warehouse.example.edu",
  dbname   = "clinical",
  user     = Sys.getenv("DB_USER"),
  password = Sys.getenv("DB_PASS")
)
```

## Querying with SQL

The `SELECT` statement is the workhorse. It names the
columns, the source table, and optional filtering,
joining, grouping, and ordering. Here we ask for the mean
systolic blood pressure and the number of visits per
treatment arm, joining the two tables on `patient_id`.

```{r}
#| label: db-sql
dbGetQuery(con, "
  SELECT d.arm,
         COUNT(*)     AS n_visits,
         AVG(v.sbp)   AS mean_sbp
  FROM visits AS v
  JOIN demographics AS d
    ON v.patient_id = d.patient_id
  GROUP BY d.arm
")
```

The clauses read in the order the database applies them:
`FROM` and `JOIN` assemble the rows, `WHERE` (absent here)
filters them, `GROUP BY` partitions them, the aggregate
functions summarize each partition, and `SELECT` names
what comes back. `dbGetQuery()` returns the result as an
ordinary data frame, which is what you want for a small
summary. For a large intermediate result you would not use
it, for the reason discussed next.

## Querying with dbplyr

Writing SQL by hand is a skill worth having, but for most
extractions you can stay in `dplyr`. The `dbplyr` package
translates a `dplyr` pipeline into SQL, sends it to the
database, and leaves the result there until you ask for
it. A reference to a database table behaves almost exactly
like a tibble.

```{r}
#| label: db-dbplyr
visits_tbl <- tbl(con, "visits")
demo_tbl   <- tbl(con, "demographics")

summary_q <- visits_tbl |>
  inner_join(demo_tbl, by = "patient_id") |>
  group_by(arm) |>
  summarize(n_visits = n(),
            mean_sbp = mean(sbp, na.rm = TRUE))

summary_q
```

The object `summary_q` is *lazy*: no computation has yet
crossed into R. The pipeline has been recorded, and when
printed it shows only a preview fetched from the database.
To see the SQL that `dbplyr` generated, ask for it:

```{r}
#| label: db-showquery
summary_q |> show_query()
```

The generated SQL is the same query we wrote by hand,
which is the point: `dbplyr` lets you think in `dplyr` and
still push the work to the database. When you are ready to
bring the result into R, for plotting, modeling, or
anything the database cannot do, call `collect()`.

```{r}
#| label: db-collect
analytic <- summary_q |> collect()
analytic
```

Everything up to `collect()` ran in the database. Only the
small summary crossed into memory. This is the pattern to
internalize: filter and aggregate lazily against the table,
then `collect()` the modest result. The verification habit
still applies, so we check that the row count matches what a
per-arm summary should give.

```{r}
#| label: db-verify
stopifnot(nrow(analytic) == n_distinct(demographics$arm))
```

```{r}
#| label: db-disconnect
#| echo: false
dbDisconnect(con)
```

::: {.callout-note icon=false collapse=true}
## Check your understanding: where does the work happen?

**Question.** You write
`tbl(con, "labs") |> filter(test == "HbA1c") |> summarize(m = mean(value))`
against a table of fifty million laboratory results, then
call `collect()`. How much data crosses from the database
into R, and what would have happened had you called
`collect()` before the `filter`?

**Answer.**

With `collect()` at the end, only the single summarized
value crosses into R: the database does the filtering and
the averaging, and returns one number. Had you called
`collect()` first, all fifty million rows would have been
pulled into memory before R did any work, which on most
laptops would exhaust RAM and fail. The rule is to
`collect()` as late as possible, after the database has
reduced the data to the size you actually need.

:::

## When to push, when to pull

@fig-collect-boundary is the whole idea of this chapter
in one picture: `collect()` is a boundary, everything to
its left runs in the database on the full data, and
everything to its right runs in R on whatever crossed.

```{mermaid}
%%| echo: false
%%| label: fig-collect-boundary
%%| fig-alt: "Left-to-right diagram with two grouped regions. Inside
%%|   the database region, a 50-million-row table passes through
%%|   filter, join, and group-and-aggregate. A collect() node marks
%%|   the boundary. Inside the R region, the reduced result feeds
%%|   model, plot, and iterate."
%%| fig-cap: "The collect() boundary. Operations that make the data
%%|   smaller belong on the database side; operations that need R's
%%|   statistical machinery belong on the R side. Moving collect()
%%|   earlier does not change the answer, only the amount of data
%%|   that has to cross and the memory it takes to hold it."
flowchart LR
  subgraph DB["In the database, on all rows"]
    T["table<br/><i>50 million rows</i>"] --> FI["filter"] --> JO["join"] --> AG["group and<br/>aggregate"]
  end
  AG --> CO(["<b>collect()</b>"])
  subgraph RR["In R, on the reduced result"]
    MO["model"]
    PL["plot"]
    IT["iterate"]
  end
  CO --> MO
  CO --> PL
  CO --> IT
```

The database is good at set operations on large data:
filtering, joining, grouping, aggregating. R is good at
everything else: modeling, plotting, iteration, anything
statistical. The dividing line is `collect()`. Push the
data-reduction work to the database; pull the reduced
result into R for the analysis. A useful test: if a step
makes the data smaller (a filter, an aggregate), it
usually belongs before `collect()`; if a step needs the
full statistical machinery of R (fitting a mixed model,
drawing a figure), it belongs after.

Two cautions. First, not every R function has a SQL
translation; `dbplyr` will tell you when it cannot
translate a verb, and then you must either rewrite it or
`collect()` earlier. Second, a query that is cheap to
write can be expensive to run; on a shared warehouse, a
careless cross-join can inconvenience every other user, so
test on a `LIMIT`-ed subset before running the full
extraction.

## Reproducibility of the extraction

The analytic dataset is a build product, and the query is
its recipe. Keep the query in the compendium as a `.sql`
file or an R script, so that the derivation is
version-controlled and re-runnable. Record the date of
extraction and, where the warehouse supports it, the data
snapshot or freeze identifier, because a warehouse is a
living system and today's query may return different rows
next month. The pattern mirrors the raw-data discipline of
@sec-wrangling: the query and its output date are the audit
trail that lets a reviewer reconstruct exactly which rows
your analysis rested on.

## Collaborating with an LLM on databases

LLMs write competent SQL and `dbplyr`. The traps, in our
experience, are correctness on joins and the silent scale of
a query.

**Prompt 1: drafting a query.** Describe the tables, their
keys, and the summary you want, and ask for both the SQL
and the `dbplyr` equivalent.

*What to watch for.* The join key and the join type. An
LLM will readily write an inner join where you needed a
left join, dropping patients with no visits. State the
cardinality you expect.

*Verification.* Run against a `LIMIT`-ed subset; count the
rows; confirm the count matches the expected number of
groups or patients.

**Prompt 2: translating hand-written SQL to `dbplyr`.**
Paste a `SELECT` statement and ask for the `dplyr`
pipeline that generates it.

*What to watch for.* Aggregations and `GROUP BY`
translate cleanly; window functions and correlated
subqueries less so. Compare `show_query()` on the result
against the original.

*Verification.* Confirm the two produce identical results
on a small table before trusting the translation on the
full one.

## Principle in use

Three habits keep the extraction defensible:

1. **Query at the source; `collect()` late.** Push
   filtering and aggregation to the database; pull only the
   analytic result into R.
2. **The query is the provenance.** Version-control the
   SQL that produced the analytic dataset; treat the
   extract as a build product with a recorded date.
3. **Verify joins in SQL as in R.** Count the rows; a
   silent many-to-many inflates the data in a database
   exactly as it does in `dplyr`.

## Exercises

1. Create an in-memory SQLite database with the
   `demographics` and `visits` tables from this chapter.
   Write a SQL query that returns, per patient, the number
   of visits and the change in systolic blood pressure
   from the first to the last visit.
2. Express the same query with `dbplyr`. Use
   `show_query()` to inspect the generated SQL and compare
   it to your hand-written version.
3. Add a third table, `labs`, with columns `patient_id`,
   `test`, and `value`. Write a `dbplyr` pipeline that
   returns the mean HbA1c per treatment arm, joining all
   three tables. Verify the row count.
4. Deliberately write a query that produces a
   many-to-many join between `visits` and a table with
   duplicate `patient_id` values. Observe the inflated row
   count and explain it.
**On your own machine.** No answer key; the interesting part
is the measurement, which depends on your hardware.

5. Using the `duckdb` backend instead of `RSQLite`, read a
   large CSV file directly with `duckdb`'s
   `read_csv_auto` and run an aggregation without ever
   loading the file into R. Compare the memory use to
   `readr::read_csv()` followed by `dplyr`.

::: {.callout-tip icon=false collapse=true}
## Solutions to exercises 1 to 4

The chapter's connection was closed, so these open a fresh
one and rebuild the two tables.

```{r}
#| label: soln-15b-setup
#| message: false
con2 <- dbConnect(RSQLite::SQLite(), ":memory:")
dbWriteTable(con2, "demographics", demographics)
dbWriteTable(con2, "visits", visits)
```

**1. Visit count and first-to-last change, in SQL.**

```{r}
#| label: soln-15b-1
dbGetQuery(con2, "
  SELECT patient_id,
         COUNT(*) AS n_visits,
         MAX(CASE WHEN r_last  = 1 THEN sbp END) -
         MAX(CASE WHEN r_first = 1 THEN sbp END) AS sbp_change
  FROM (
    SELECT patient_id, visit, sbp,
           ROW_NUMBER() OVER (PARTITION BY patient_id
                              ORDER BY visit)      AS r_first,
           ROW_NUMBER() OVER (PARTITION BY patient_id
                              ORDER BY visit DESC) AS r_last
    FROM visits
  )
  GROUP BY patient_id
")
```

The window functions number each patient's visits from
both ends, so the first and last rows can be picked out
inside a single aggregation. Patients with one visit get a
change of zero, which is correct and worth noticing: it is
not the same as missing, and a real analysis should decide
which it wants.

**2. The same query through `dbplyr`, and its SQL.**

```{r}
#| label: soln-15b-2
#| warning: false
q <- tbl(con2, "visits") |>
  group_by(patient_id) |>
  window_order(visit) |>
  mutate(first_sbp = first(sbp), last_sbp = last(sbp)) |>
  ungroup() |>
  group_by(patient_id) |>
  summarize(n_visits = n(),
            sbp_change = max(last_sbp) - max(first_sbp))

q |> collect()
q |> show_query()
```

Two things are worth extracting from this.

`first()` and `last()` are *window* functions, so they
belong in `mutate()` and not in `summarize()`. Writing
`summarize(sbp_change = last(sbp) - first(sbp))` looks
natural and fails, because a window function needs the
rows it is ordered over and an aggregate has already
collapsed them. The two-step form computes the values per
row, then aggregates.

`window_order()` has no dplyr equivalent in memory, where
`first()` and `last()` simply follow row order. A database
table has no inherent order, so it must be stated. Omit it
and the answer is whichever row the engine happened to
reach first, with no warning.

The generated SQL is more verbose than the hand-written
version and computes the same thing. Which to prefer is a
maintenance question, not a performance one: the `dbplyr`
version stays readable to someone who knows dplyr and not
SQL, and `show_query()` means the SQL is never hidden.

**3. A third table, and a three-way join.**

```{r}
#| label: soln-15b-3
labs <- data.frame(
  patient_id = c(1, 1, 2, 3, 3, 4, 5),
  test       = "HbA1c",
  value      = c(7.1, 6.8, 8.2, 7.5, 7.3, 6.9, 8.0)
)
dbWriteTable(con2, "labs", labs)

result <- tbl(con2, "labs") |>
  filter(test == "HbA1c") |>
  inner_join(tbl(con2, "visits") |>
               distinct(patient_id), by = "patient_id") |>
  inner_join(tbl(con2, "demographics"), by = "patient_id") |>
  group_by(arm) |>
  summarize(mean_hba1c = mean(value, na.rm = TRUE),
            n_labs = n(), .groups = "drop") |>
  collect()

result
stopifnot(nrow(result) == n_distinct(demographics$arm))
```

Note the `distinct(patient_id)` on `visits`. Without it
the join multiplies each lab result by that patient's
visit count, and the mean silently becomes a
visit-weighted average rather than a lab-weighted one. The
row-count assertion would not catch that, because the
result still has one row per arm. Exercise 4 is the same
error made visible.

**4. A deliberate many-to-many, and why the count inflates.**

```{r}
#| label: soln-15b-4
# demographics with a duplicated patient, as a re-consent
# appended rather than replacing the original row
demo_dup <- rbind(demographics,
                  data.frame(patient_id = 1, age = 65,
                             arm = "Active"))
dbWriteTable(con2, "demo_dup", demo_dup)

n_before <- tbl(con2, "visits") |> count() |> pull(n)
n_after  <- tbl(con2, "visits") |>
  inner_join(tbl(con2, "demo_dup"), by = "patient_id") |>
  count() |> pull(n)

c(visits = n_before, after_join = n_after)
```

```{r}
#| label: soln-15b-disconnect
#| echo: false
dbDisconnect(con2)
```

Patient 1 has two visits and now two demographics rows, so
those two visits become four. The join did not fail and
issued no warning; SQL has no equivalent of dplyr's
`relationship` argument. The only defense is to check the
key before joining, which is why the chapter's own
pipeline asserts uniqueness rather than trusting it.
:::

## Further reading

- The `DBI` documentation at `dbi.r-dbi.org`, the
  interface reference.
- The `dbplyr` documentation at `dbplyr.tidyverse.org`,
  including the list of translatable verbs.
- [@wickham2023r4ds], the databases chapter, applied
  introduction to `DBI` and `dbplyr`.
- The DuckDB documentation at `duckdb.org`, for
  file-backed analytical queries without a server.

## Prerequisites answers {#sec-quiz-answers-15b}

1. `dbGetQuery()` sends a query and returns only its
   result; a `SELECT` with filtering and aggregation
   returns a small summary regardless of table size.
   `dbReadTable()` reads an entire table into memory, so
   on a ten-million-row table it pulls all ten million
   rows across, which is slow and may exhaust memory. Use
   `dbGetQuery()` (or lazy `dbplyr`) to let the database do
   the reduction.
2. A `dbplyr` pipeline builds up a query lazily: each verb
   adds to a SQL statement that has not yet run. No
   computation crosses into R until you call `collect()`
   (or print, which fetches a preview). The computation
   happens in the database; R receives only the collected
   result.
3. The extracted CSV records the rows but not their
   derivation. The SQL query records exactly which tables,
   filters, joins, and aggregations produced the analytic
   dataset, so a reviewer can reconstruct or re-run it.
   A warehouse is also a living system, so pairing the
   query with an extraction date documents the snapshot the
   analysis rested on.
