# Testing Data Analysis Workflows {#sec-testing}

::: {.bookepigraph}
```{=latex}
\begin{flushright}\small
```
*Program testing can be used to show the presence of bugs,
but never to show their absence.*

[Edsger W. Dijkstra]{.smallcaps}, *Notes on Structured
Programming* (1970)
```{=latex}
\end{flushright}
```
:::

::: {.callout-note}
## Sources

Adapted from author's lecture notes and supporting
materials for a graduate practicum in biostatistics.
:::

## Prerequisites

Answer the following questions to see if you can bypass
this chapter. You can find the answers at the end of the
chapter in @sec-quiz-answers-20.

1. Why would you test a one-off data analysis script
   that will only ever run on a single dataset?
2. What is the difference between a unit test, an
   integration test, and an end-to-end test in the
   context of a data analysis pipeline?
3. What does `tinytest::expect_equal_to_reference()`
   capture, and when is a reference test preferable to a
   direct value-comparison test?

## Learning objectives

By the end of this chapter you should be able to:

- Write unit tests for analytic helper functions
  with `tinytest`.
- Add an integration test that exercises the full
  data pipeline on a small synthetic dataset.
- Use `expect_equal_to_reference()` to capture complex
  outputs (printed objects, summary tables).
- Run the test suite locally via
  `tinytest::test_all()` and in CI via GitHub
  Actions.
- Identify test smells (tests that pass for the
  wrong reason, tests that hide bugs, flaky tests).
- Apply visual regression testing to ggplot output
  via `vdiffr`.

## Orientation

Analysis code rarely gets tested, and the reason is a
respectable one. Software is written to run on inputs its
author has never seen; an analysis script is written to run on
one dataset, once, and then to be thrown away. Testing
something that runs once against known data can look like
ceremony.

The reasoning has two holes. The first is that the script does
not run once. It runs every time a data extract is refreshed,
an exclusion is revised, or a reviewer asks for a sensitivity
analysis. By then its author has forgotten which of its
branches were ever exercised. The second is more serious: the
claim 'it runs on one known dataset' is doing no work, because
the failures that matter in analysis code do not stop it
running. A filter that removes the wrong rows, a join that
duplicates a patient, a unit conversion applied twice, all
produce output. The question a test answers is not 'did it
run' but 'would I have noticed if it were wrong', and for most
analysis scripts the honest answer is no.

We bring in this chapter the habits of software engineering to
bear on that question.

The framework used in this chapter is `tinytest`,
because it is what projects scaffolded with `zzcollab`
ship with and what the rest of this book assumes. It has
no dependencies of its own, which matters when the point
of the surrounding chapters is to pin an environment;
each test file runs on its own; and the assertion
vocabulary is small enough to hold in your head.

::: {.callout-note}
## A note on testthat

`testthat` is the framework you are most likely to meet
elsewhere, and the one the companion volume uses in the
package-development context. The discipline this chapter
teaches, testing statistical content rather than plumbing,
is identical under either, and most of the translation is
mechanical. Three differences are worth knowing before you
read `testthat` code:

- **There is no wrapper.** `testthat` groups assertions
  inside `test_that("description", { ... })`. In
  `tinytest` each `expect_*()` call is itself a test, and
  the description goes in an `info=` argument. Files stay
  flatter, and a failure reports the file and line rather
  than a block name.
- **The vocabulary is smaller.** `tinytest` has no
  `expect_named()`, `expect_setequal()`, `expect_type()`
  or `expect_s3_class()`. Use `expect_equal(names(x), ...)`,
  `expect_equal(sort(a), sort(b))`,
  `expect_equal(typeof(x), ...)` and `expect_inherits()`.
  This is a feature rather than a gap: each is a plain
  comparison whose failure output shows the two values.
- **Snapshots work differently.**
  `expect_equal_to_reference()` stores an `.rds` of the
  object rather than printed text.

Where a project already depends on `testthat`, stay with
it. Mixing frameworks in one suite buys nothing.
:::

## The statistician's contribution {#sec-testing-human}

Writing a test is five minutes of work. Writing a test
that would actually fail if the code were wrong is the
part that requires thought:

**Test the statistical content, not just the
plumbing.** A test that 'returns a tibble of the
right shape' is a structural check, not a test of
correctness. A test that the regression coefficient
on a known synthetic dataset matches the closed-form
answer is a test of correctness.

**Edge cases are where bugs live.** Empty input,
single observation, all-NA, perfect collinearity,
boundary value (zero variance, all-zero counts,
class with one observation). Test these
deliberately.

**Test the pipeline, not just the helpers.** Unit
tests catch helper-function bugs. Integration tests
on synthetic data catch pipeline bugs, including
ones where helpers work individually but the
composition fails.

**Don't test what you don't trust.** Mocking out a
function call inside a test of that function makes
the test tautological. The test passes because you
made it pass, not because the code is correct.

These judgments are what make tests useful rather
than performative.

## Why test analyses?

The Orientation argued the general case. Three specific
occasions make it concrete, and each is a moment that
arrives in ordinary projects rather than unusual ones.

The first is **refactoring**. You restructure a cleaning
script to use `dplyr` where it previously used
`base::merge`, and the output should be identical. Note
what is peculiar about this situation: you have a
precise expectation, you have no way to check it by
looking, and the consequence of being wrong is a
silently different cohort. A test that compares the
output against a saved baseline settles the question in
a second, and without one the honest position after a
refactor is that you do not know whether it changed
anything.

The second is **upstream drift**, and it is the case
tests catch that nothing else does. Your script reads an
extract from an institutional warehouse, and one quarter
the warehouse begins reporting a duration in seconds
where it previously reported days. Nothing errors. The
column is still numeric, the join still succeeds, the
model still fits, and every downstream number is wrong
by a factor of 86,400. A test asserting that the column
falls in a plausible range catches this on the first run
after the change; no amount of reading your own code
ever will, because your code is not what changed.

The third is **documentation**. A test of the form
'given input X, this function returns Y' is a statement
about intent that cannot rot, because it is executed.
Comments drift away from the code they describe;
assertions cannot, since a false one fails the build.

Against these the cost is a few minutes per test,
weighed against a silent miscalculation in something you
have published under your name.

Tests come in layers, and the layers catch different
failures. @fig-test-layers is the shape, with the
important qualification that the pyramid familiar from
software engineering is somewhat flatter for analysis
code: the integration layer, a small synthetic dataset
run through the whole pipeline, does more work here than
it does in most applications.

```{mermaid}
%%| echo: false
%%| label: fig-test-layers
%%| fig-alt: "Three stacked layers in increasing scope. Unit tests
%%|   exercise one function on known input, catch helper bugs, and
%%|   cost seconds. Integration tests run the whole pipeline on a
%%|   twenty-line synthetic dataset, catch composition bugs and
%%|   silent filter and join errors, and cost seconds. End-to-end
%%|   tests run raw data through to final figures, catch everything
%%|   remaining, and cost minutes to hours."
%%| fig-cap: "The three test layers and what each one catches.
%%|   Unit tests are fast and narrow; end-to-end tests are slow and
%%|   broad; the integration layer, where a synthetic dataset is run
%%|   through the whole cleaning pipeline, has the best cost-benefit
%%|   ratio for analysis code and is the layer most often missing."
flowchart TD
  U["<b>Unit</b><br/><i>one function, known input</i><br/>catches: helper bugs<br/>cost: seconds"]
  I["<b>Integration</b><br/><i>whole pipeline, 20-line<br/>synthetic dataset</i><br/>catches: composition bugs,<br/>silent filter and join errors<br/>cost: seconds"]
  E["<b>End to end</b><br/><i>raw data to final figures</i><br/>catches: everything left<br/>cost: minutes to hours"]
  U --> I --> E
```

::: {.callout-note icon=false collapse=true}
## Check your understanding: what would you have tested?

**Question.** A cleaning script filters to adults with
`filter(age >= 18)`. Six months later someone discovers
the export encodes unknown age as `-1`, so those rows were
silently dropped as minors. Which layer of the diagram
would have caught it, and what would the test have
asserted?

**Answer.**

The integration layer, and only if the synthetic dataset
contained a `-1`.

That qualification is the whole lesson. A unit test on the
filter would pass: given ages 17, 25 and 40, it correctly
keeps two. The function is not broken. The *data* violated
an assumption the function never stated, and no test of
the function alone can see that.

An integration test catches it if the synthetic fixture
exercises the case, which is why the guidance says the
fixture should exercise every branch, including the
branches you do not think exist. A fixture built from
plausible values only would miss it too.

There are two assertions worth writing, and they fail at
different times. One on the input, asserting that `age`
falls in a plausible range before the filter runs, which
fails the day the export changes. One on the output,
asserting that the number of rows dropped is what the
codebook predicts, which fails the day the meaning
changes.

The first is a claim about the data; the second is a claim
about your understanding of it. Analysis pipelines mostly
break for the second reason, and mostly get tested only
for the first.
:::

## Unit tests with tinytest

For a function in `R/clean.R`:

```{r}
#| label: test-age-group-fn
# R/clean.R
#' Compute age groups from numeric age
#' @param age numeric vector
age_group <- function(age) {
  cut(age,
      breaks = c(0, 18, 40, 65, Inf),
      right = FALSE,
      labels = c("under-18", "18-39", "40-64", "65+"))
}
```

Tests in `inst/tinytest/test-clean.R`:

```{r}
#| label: test-age-group-tests
#| message: false
library(tinytest)

expect_equal(as.character(age_group(c(20, 50, 75))),
             c("18-39", "40-64", "65+"),
             info = "typical adult ages bin correctly")

# the interval starts are inclusive, because right = FALSE
expect_equal(as.character(age_group(c(18, 40, 65))),
             c("18-39", "40-64", "65+"),
             info = "boundary values fall in the upper bin")
expect_equal(as.character(age_group(c(17, 39, 64))),
             c("under-18", "18-39", "40-64"),
             info = "values just below a break stay below")

# NA_real_, not NA: bare NA is logical, and cut() rejects it. The
# distinction is invisible until the test is actually run.
expect_true(is.na(age_group(NA_real_)),
            info = "numeric NA propagates rather than erroring")

expect_error(age_group("twenty"),
             info = "non-numeric input is rejected")
```

Each `expect_*()` call is one test. There is no
enclosing block, which is the most visible difference
from `testthat`: the `info=` string does the work
`test_that()`'s description does, and a failure reports
the file and line number alongside it.

Run all tests:

```r
tinytest::test_all()          # from the project root
# or, for a package:
tinytest::test_package("practicum")
```

The layout `tinytest` expects is `inst/tinytest/` for the
test files and a one-line `tests/tinytest.R` that runs
them, which is what `zzcollab` scaffolds. Shipping tests
under `inst/` has a consequence worth knowing: they are
installed with the package, so a user who has your
compendium can run its test suite without your source
tree. For an analysis whose credibility rests on being
checkable by someone else, that is the right default.

## Integration tests on synthetic data

Unit tests catch helper-function bugs. They do not
catch bugs where the helpers work individually but
the pipeline composition fails.

Integration tests run the full pipeline on a small
synthetic dataset:

```{r}
#| label: test-integration
#| message: false
# inst/tinytest/test-pipeline.R
# In a package the test file would load run_pipeline from R/; it is
# defined here so the example executes as written.
run_pipeline <- function(raw) {
  raw |>
    janitor::clean_names() |>
    dplyr::filter(age >= 18) |>
    dplyr::mutate(age_group = age_group(age)) |>
    tidyr::pivot_longer(dplyr::starts_with("bp_"),
                        names_to = "visit", names_prefix = "bp_v",
                        values_to = "bp") |>
    dplyr::filter(!is.na(bp)) |>
    dplyr::mutate(visit = as.integer(visit)) |>
    dplyr::select(patient_id, age_group, sex, treatment, visit, bp)
}

# synthetic data that exercises every branch
raw <- tibble::tibble(
  patient_id = c(1, 2, 3, 4),
  age        = c(25, 50, 75, 17),     # one excluded
  bp_v1      = c(120, 140, 160, NA),
  bp_v2      = c(118, 138, NA,  NA),
  sex        = c("M", "F", "F", "M"),
  treatment  = c("placebo", "active", "placebo", "active")
)

clean <- run_pipeline(raw)

# structural assertions. tinytest has no expect_s3_class or
# expect_named, and does not need them: both are ordinary
# comparisons, and a failure prints the two values.
expect_inherits(clean, "tbl_df", info = "returns a tibble")
expect_equal(names(clean),
             c("patient_id", "age_group", "sex",
               "treatment", "visit", "bp"),
             info = "column names and order are as specified")

# row count: 3 adults x 2 visits, minus NAs
expect_equal(nrow(clean), 5,
             info = "patient 4 excluded; patient 3 has one NA bp")

# specific values
expect_equal(clean$bp[clean$patient_id == 1 & clean$visit == 1],
             120,
             info = "a known cell survives the pipeline unchanged")
```

The synthetic dataset is small (4 patients, 2
visits) but exercises each cleaning rule: age
exclusion (patient 4), missing values (patient 3
visit 2). A test that runs in a fraction of a second
covers the pipeline's logic.

For each cleaning rule (filter, derive, pivot,
recode), the synthetic dataset should have at least
one row that exercises it.

::: {.callout-note icon=false collapse=true}
## Check your understanding: structural vs. correctness

**Question.** Your test asserts
`expect_equal(nrow(result), 100)`. The test passes.
Does this mean the result is correct?

**Answer.**

No. The test asserts only that the result has 100
rows. It says nothing about which rows, what
values, or whether the right rows were filtered.
A bug that produces 100 rows of the wrong patients
passes the test silently. To test correctness:

```{r}
#| label: test-structural-vs-correct
result        <- data.frame(patient_id = 1:100, age = rep(40:49, 10))
expected_ids  <- 1:100
expected_ages <- rep(40:49, 10)

expect_equal(nrow(result), 100)            # structural
expect_equal(sort(result$patient_id),      # correctness
             sort(expected_ids))
expect_equal(result$age, expected_ages)
```

The combination tests structure *and* content. The
structural test alone is a false positive
generator. This pattern, 'looks right at the
surface, wrong underneath', is the most common
testing failure mode in data analysis.

:::

## Reference tests

For complex outputs that are hard to assert
literally, compare against a stored reference:

```r
fit <- lm(mpg ~ wt + hp, data = mtcars)
expect_equal_to_reference(coef(fit), "ref/mtcars-coefs.rds",
                          info = "coefficients unchanged")

expect_equal_to_reference(summary(fit)$r.squared,
                          "ref/mtcars-r2.rds")
```

On first run the object is written to the named file and
the test passes trivially. On subsequent runs the value is
compared against it; if it differs the test fails, and you
decide whether the change is intended (delete the file and
regenerate) or a regression.

One difference from `testthat` is worth understanding
rather than glossing. `expect_snapshot()` captures
*printed output* as text, so it also detects changes in
formatting. `expect_equal_to_reference()` stores the
*object*, so it compares values and ignores how they
print. For analysis code the object comparison is usually
what you want: a coefficient that changed matters, and a
`print` method that gained a blank line does not.

Reference tests are useful when:

- Output is complex and tedious to assert literally.
- Output is human-readable formatting (tables,
  printed summaries, plot text).
- You care more about 'is this what was reviewed' than
  'is this exactly the right value'.

Snapshots are not useful when:

- Output changes for trivial reasons (random seed,
  system locale, an unstable sort). These produce false
  positives.
- The stored object is too large to inspect when it
  changes, which makes 'accept or investigate' a coin
  toss.

## Visual regression testing with vdiffr

For ggplot output, `vdiffr` saves an SVG of the
rendered plot and compares on subsequent runs:

```r
library(vdiffr)

fit <- lm(mpg ~ wt, data = mtcars)
p <- ggplot(broom::augment(fit), aes(.fitted, .resid)) +
      geom_point() +
      geom_hline(yintercept = 0)
expect_doppelganger("regression-residual-plot", p)
```

vdiffr handles cross-platform rendering issues
(font differences, anti-aliasing) better than raw
image comparison. Useful for catching unintended
changes to plot code (an axis label change, a
geom swap).

Note that `vdiffr` is built on `testthat` and expects to
run inside it, so this is the one place in the chapter
where the two frameworks do not substitute for each
other. If visual regression matters to a project, that is
a reason to use `testthat` there; `tinysnapshot` is the
`tinytest`-native alternative and is younger.

## Continuous integration

`usethis::use_github_action("check-standard")` sets
up GitHub Actions to run `R CMD check` on every
push and pull request, on multiple OS / R-version
combinations. (Older material calls this
`use_github_action_check_standard()`, a function
`usethis` has since removed in favor of naming the
workflow as an argument.) For a non-package analysis
project, the equivalent is a workflow that runs tests:

```yaml
# .github/workflows/test.yaml
name: tests

on: [push, pull_request]

jobs:
  test:
    runs-on: ubuntu-latest
    steps:
      - uses: actions/checkout@v4
      - uses: r-lib/actions/setup-r@v2
      - uses: r-lib/actions/setup-r-dependencies@v2
        with:
          packages: |
            local::.
            any::tinytest
      - name: Run tests
        run: Rscript -e 'tinytest::test_all()'
```

Push the workflow; on every commit, GitHub runs the
tests and reports pass/fail. Regressions caught at
push time are far cheaper to fix than regressions
caught in production.

## Test smells

Common patterns that look like tests but produce
false confidence:

**Tautological tests.**

```{r}
#| label: test-antipattern-tautology
# foo ignores its argument and always returns 42. It is wrong for
# every input, and the test below still passes.
foo <- function(x) 42

expect_equal(foo(1), foo(1),      # always passes
             info = "foo works")
```

The test cannot fail because it tests `foo()` against
itself. Useless.

**Tests that mock the tested function.**

```r
fit_model <- function(d) list()          # stub, for the test
expect_equal(typeof(fit_model(d)), "list")
```

You replaced the function under test with a stub; the
test passes because the stub returns a list, not because
`fit_model` works. `tinytest` has no mocking facility,
which makes this smell harder to write by accident than
it is under `testthat`'s `with_mocked_bindings()`.

**Brittle snapshots.**

```r
expect_equal_to_reference(do_thing(today()),
                          "ref/do_thing.rds")   # date in output
```

The stored object includes the current date, so the test
fails every day after the first. Either take the date out
of the returned object or compare the parts that should be
stable, which usually means asserting on specific fields
rather than on the whole object.

**Overly permissive tolerances.**

```{r}
#| label: test-antipattern-tolerance
result <- 0.9                                # nowhere near 0.5

expect_equal(result, 0.5, tolerance = 0.5)   # passes for any value
```

A 100% tolerance is no test.

**Tests that test the test framework, not the
code.**

```{r}
#| label: test-antipattern-vacuous
expect_true(TRUE)
```

These are placeholders that should be filled in.
Watch for them in code reviews.

## Worked example: testing a cleaning pipeline

```{r}
#| label: test-worked-cleaning
#| message: false
# R/cleaning.R
clean_visits <- function(raw) {
  raw |>
    janitor::clean_names() |>
    dplyr::filter(age >= 18) |>
    tidyr::pivot_longer(
      cols = dplyr::starts_with("bp_"),
      names_to = "visit",
      names_prefix = "bp_v",
      values_to = "bp"
    ) |>
    dplyr::filter(!is.na(bp))
}

# inst/tinytest/test-cleaning.R
# In the package this line loads the function above; here it is
# already in scope, so the tests run as written.
# source("../../R/cleaning.R")

# the age filter
raw <- tibble::tibble(
  patient_id = 1:3,
  age        = c(25, 17, 30),
  bp_v1      = c(120, 110, 130),
  bp_v2      = c(122, 112, 128)
)
result <- clean_visits(raw)
expect_equal(sort(unique(result$patient_id)), c(1, 3),
             info = "cleaning excludes minors")

# the NA filter
raw <- tibble::tibble(
  patient_id = c(1, 1, 2),
  age        = c(25, 25, 30),
  bp_v1      = c(120, NA, 130),
  bp_v2      = c(NA,  118, 128)
)
result <- clean_visits(raw)
expect_equal(nrow(result), 4,
             info = "6 cells, 2 NA, so 4 rows survive")
expect_true(all(!is.na(result$bp)),
            info = "no NA blood pressures remain")

# the reshape
raw <- tibble::tibble(
  patient_id = 1,
  age        = 25,
  bp_v1      = 120,
  bp_v2      = 118
)
result <- clean_visits(raw)
expect_equal(sort(names(result)),
             sort(c("patient_id", "age", "visit", "bp")),
             info = "long format has the expected columns")
expect_equal(nrow(result), 2,
             info = "one patient with two visits gives two rows")
```

The tests cover: the age filter, the NA filter, the
pivot. Each test is a single named scenario; each
asserts specific values, not just structure.
Running the test suite catches any regression to
the cleaning logic.

## Collaborating with an LLM on tests

LLMs draft tests well; the judgment about what to
test needs human input.

**Prompt 1: drafting tests.** Paste the function and
ask: 'write `tinytest` tests covering happy path,
edge cases, and error handling. Make the assertions
specific to values, not just structure.'

*What to watch for.* Default LLM tests tend to be
shape-checks. Push for value-checks. Edge cases:
empty, NA, single observation, type mismatch.

*Verification.* Introduce a bug into the function
and re-run the tests. If the bug is caught, the
tests are useful. If not, add a more specific
test.

**Prompt 2: integration test on synthetic data.**
Describe the pipeline; ask the LLM to generate a
small synthetic dataset that exercises every
branch.

*What to watch for.* The LLM may generate a dataset
that exercises the happy path but not the edges.
Verify each branch (filter rules, missing-data
paths, factor levels) is covered.

*Verification.* Add a deliberate bug to one branch;
the test should fail. If it does not, the synthetic
data does not exercise that branch.

**Prompt 3: diagnosing a flaky test.** Paste the
test and the failure; ask: 'what's the source of
non-determinism?'

*What to watch for.* Common causes: random number
generation without seed, parallel processing
without seed-aware RNG, system-time-dependent
output, locale issues. The LLM should know these.

*Verification.* Apply the fix and run the test
multiple times. If it stops being flaky, fixed.

## Principle in use

Three habits separate a suite that catches bugs from one
that reports green:

1. **Test the math, not just the shape.**
   Structural checks alone produce false confidence.
2. **Integration tests on synthetic data.** A
   20-line synthetic dataset that exercises every
   branch catches pipeline bugs that unit tests
   miss.
3. **CI as a tripwire.** GitHub Actions running
   tests on every push catches regressions at the
   commit, not at the publication.

## Exercises

1. `age_group()` earlier in this chapter is tested
   for typical ages, boundaries, `NA`, and
   non-numeric input. One gap remains: what the
   function does with an age outside the range the
   breaks cover. Find the behavior, decide whether
   it is the behavior you want, and write the test
   that pins whichever answer you choose.
2. Write a `summarize_species()` function returning
   mean body mass and count per species from
   `palmerpenguins::penguins`. Give it one
   structural test and one correctness test, and say
   in a comment which is which and why the
   structural one is insufficient on its own.
3. Take the tautological test from this chapter,
   `expect_equal(foo(1), foo(1))`, and replace it
   with an assertion that fails against
   `foo <- function(x) 42` while passing against a
   correct implementation.

**On your own project.** No answer key; these need code,
a repository, or a colleague you supply.

4. Add unit tests for two functions from an
   existing analysis of yours. Aim for at least: a
   happy-path test, an edge-case test, and an
   error-path test.
5. Write an integration test: a 20-line synthetic
   dataset that exercises every branch of your
   pipeline. Run the full pipeline in under a
   second.
6. Set up GitHub Actions with a workflow that runs
   `tinytest::test_all()` on every push. Push and
   verify the workflow runs green.
7. Introduce a deliberate bug into your pipeline.
   Verify the test suite catches it. If it does
   not, add a test that does.
8. Apply `vdiffr::expect_doppelganger` to one of
   your project's plots. Modify the plot's theme
   slightly; verify the test fails and you can
   accept or reject the change.
9. Review a colleague's test suite rather than their
   code. For each test, decide whether it checks
   statistical content or only structure, and find
   at least one test that would pass even if the
   function under test were subtly wrong. Propose the
   value-level assertion that would catch the bug,
   and deliver the review in writing as a
   pull-request comment (@sec-git-teams).

::: {.callout-tip icon=false collapse=true}
## Solutions to exercises 1 to 3

**1. What happens outside the range of the breaks.**

```{r}
#| label: soln-20-1
as.character(age_group(c(-5, 0, 200)))
```

A negative age returns `NA`. The breaks begin at 0, and
`cut()` places anything below the first break outside every
interval. An age of 200 is absorbed into `65+`, because
the last break is `Inf`.

The asymmetry is the finding. An impossible age at the top
of the range is silently accepted as plausible; an
impossible age at the bottom becomes `NA`, which is the
same value a genuinely missing age produces. Downstream,
a data-entry error of `-5` is indistinguishable from an
age nobody recorded.

Whether that is acceptable is a judgment, and the test
should pin whichever judgment you make. Pinning the
current behavior:

```{r}
#| label: soln-20-1b
expect_true(is.na(age_group(-5)),
            info = "age below the first break gives NA")
```

Preferring an error instead, the test comes first and the
function changes to satisfy it:

```{r}
#| label: soln-20-1c
age_group_strict <- function(age) {
  stopifnot(is.numeric(age))
  if (any(age < 0, na.rm = TRUE))
    stop("negative age")
  cut(age, breaks = c(0, 18, 40, 65, Inf), right = FALSE,
      labels = c("under-18", "18-39", "40-64", "65+"))
}

expect_error(age_group_strict(-5), pattern = "negative age",
             info = "negative age is rejected")
expect_equal(as.character(age_group_strict(30)), "18-39",
             info = "a valid age still bins correctly")
```

Either is defensible. What is not defensible is leaving
the behavior untested, because then it is not a decision
at all, only whatever `cut()` happened to do.

**2. A structural test and a correctness test.**

```{r}
#| label: soln-20-2
#| message: false
library(palmerpenguins)

summarize_species <- function(data) {
  data |>
    dplyr::filter(!is.na(body_mass_g)) |>
    dplyr::group_by(species) |>
    dplyr::summarize(mean_mass = mean(body_mass_g),
                     n = dplyr::n(), .groups = "drop")
}

# STRUCTURAL: shape only. Every line below passes even if
# mean_mass is computed as a median, a sum, or the wrong column.
out <- summarize_species(penguins)
expect_inherits(out, "tbl_df", info = "structural: a tibble")
expect_equal(names(out), c("species", "mean_mass", "n"),
             info = "structural: column names")
expect_equal(nrow(out), 3, info = "structural: one row per species")

# CORRECTNESS: values the test computes independently.
adelie <- penguins$body_mass_g[penguins$species == "Adelie"]
expect_equal(out$mean_mass[out$species == "Adelie"],
             mean(adelie, na.rm = TRUE),
             info = "correctness: the mean is the mean")
expect_equal(out$n[out$species == "Adelie"],
             sum(!is.na(adelie)),
             info = "correctness: n counts non-missing")
```

The structural test is worth keeping, because a change in
column names breaks callers. It is insufficient because
every assertion in it would still pass if `mean()` were
replaced by `median()`. Only the second test would fail,
and only because it recomputes the answer a different way
rather than comparing the function to itself.

**3. Replacing the tautology.**

```{r}
#| label: soln-20-3
#| error: true
foo <- function(x) 42          # wrong for every input

expect_equal(foo(1), 2, info = "foo doubles 1")
expect_equal(foo(3), 6, info = "foo doubles 3")
```

The fix is to assert against a value computed
independently of the function, here by hand. Any assertion
whose expected value comes from calling the function is
circular, and no amount of adding cases repairs that.
:::

## Further reading

- [@wickham2023rpkgs] testing chapters, the
  canonical `testthat` reference.
- `vdiffr` documentation on CRAN, visual
  regression testing for `ggplot2`.
- The `testthat` package vignettes.

## Prerequisites answers {#sec-quiz-answers-20}

1. Even a one-off analysis benefits from tests:
   they catch regressions when you refactor
   mid-analysis, they serve as executable
   documentation of expected behavior, and they
   detect silent failures when upstream data
   changes format. The cost of writing a handful of
   tests is small; the cost of silently
   mis-analyzing data is much larger.
2. A unit test exercises a single function with
   known inputs and checks the output. An
   integration test runs a pipeline segment
   (multiple functions) on synthetic data and
   checks the pipeline's output. An end-to-end
   test runs the full analysis pipeline from raw
   data to final figures/tables. Unit tests are
   fast and narrow; end-to-end tests are slow and
   broad; integration tests sit in between, with
   good cost-benefit for analytic pipelines.
3. `expect_snapshot()` captures the output of an
   expression (printed text, numerical result,
   plot) to a file on first run. On subsequent
   runs it compares the new output against the
   saved snapshot. Use it when the output is
   complex (a printed object, a plot, a formatted
   table) and not easily expressed as a literal
   value. Review changes manually when they
   occur; do not auto-accept.
